### 2017-12-13 paper/main.tex
Finished Python scripts to 1) translate jobs into different languages 2) organize everything in .csv tables and 3) translate sentences back into english, extracting their predicted gender. The results of this extraction are saved in the table job-genders.csv

\documentclass{article}

\usepackage{multirow}
\usepackage{hyperref}
\usepackage{float}
\usepackage{array}
\usepackage{dirtytalk}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\title{Assessing Gender Bias in Machine Translation}
\author{Marcelo Prates \and Luis Lamb}

\begin{document}

\maketitle

\begin{abstract}
\end{abstract}

\section{Introduction}

Although the idea of automated translation can in principle be traced back to as long as the 17th century with René Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by Warren Weaver \cite{weaver1955translation} discussing the possibility of employing digital computers to perform automated translation. The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}. Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation. As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential ALPAC report \cite{hutchins1986machine}. Research was almost completely abandoned in the United States, making a shy re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical}. Statistical and example-based machine translation have been on the rise ever since \cite{carl2003recent}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily \cite{bengio2015deep}.

In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism. Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data. Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics:

\begin{quotation}
I mean actually you could do physics this way, instead of studying things like balls rolling down frictionless planes, which can't happen in nature, if you took a ton of video tapes of what's happening outside my office window, let's say, you know, leaves flying and various things, and you did an extensive analysis of them, you would get some kind of prediction of what's likely to happen next, certainly way better than anybody in the physics department could do. Well that's a notion of success which is I think novel, I don't know of anything like it in the history of science.
\end{quotation}

Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained}:

\begin{quotation}
As another example, consider the Newtonian model of gravitational attraction, which says that the force between two objects of mass $m_1$ and $m_2$ a distance $r$ apart is given by

\begin{equation*}
F = G m_1 m_2 / r^2
\end{equation*}

where $G$ is the universal gravitational constant. This is a trained model because the gravitational constant G is determined by statistical inference over the results of a series of experiments that contain stochastic experimental error. It is also a deterministic (non-probabilistic) model because it states an exact functional relationship. I believe that Chomsky has no objection to this kind of statistical model. Rather, he seems to reserve his criticism for statistical models like Shannon's that have quadrillions of parameters, not just one or two.
\end{quotation}

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society. On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English, as Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies.

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
	\label{fig:screenshot-gtranslate-hungarian}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
\end{figure}

\section{Motivation}

\section{Methods}

We believe that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse}. The same structure can be

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|m{3cm}|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported \\ \hline
	\multirow{2}{*}{Austronesian} & Malay & No & Yes \\
	& Tagalog & No & \textbf{No} \\ \hline
	\multirow{3}{*}{Finno-Ugric} & Estonian & No & Yes \\
	& Finnish & No & Yes\\
	& Hungarian & No & Yes \\ \hline
	\multirow{4}{*}{Indo-European} & Armenian & No & Yes \\
	& Bengali & No & Yes \\
	& English & \textbf{Yes} & Yes \\
	& Persian & \textbf{Yes} & Yes \\ \hline
	\multirow{3}{*}{Indo-Aryan} & Maithili & No & \textbf{No} \\
	& Nepali & No & Yes \\
	& Oriya & No & \textbf{No} \\ \hline
	\multirow{10}{*}{} & Japanese & No & Yes \\
	& Korean & \textbf{Optional} & Yes \\
	& Turkish & No & Yes \\
	& Yoruba & No & Yes \\
	& Basque & No & Yes \\
	& Swahili & No & Yes \\
	& Chinese & \textbf{Optional} & Yes \\
	& Cantonese & No & \textbf{No} \\
	& Pipil & No & \textbf{No} \\
	& Quechuan & No & \textbf{No} \\ \hline
	\multirow{4}{*}{Constructed} & Esperanto & \textbf{Yes} & Yes \\
	& Ido & \textbf{Optional} & \textbf{No} \\
	& Lingua Franca Nova & No & \textbf{No} \\
	& Interlingua & No & \textbf{No} \\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system. For the purposes of this work, we have decided to work only with languages lacking such a system, and as such Persian and Esperanto have been discarded. Because Tagalog, Maithili, Oriya, Cantonese, Pipil, Quechuan, Lingua Franca Nova and Interlingua lack support from Google Translate, these languages have also been ommitted from this work.}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Category 		& \# Occupations 	\\ \hline
	Artistic 		& $102$ 			\\ \hline
	Computer 		& $19$ 				\\ \hline
	Corporate 		& $50$ 				\\ \hline
	Dance 			& $9$ 				\\ \hline
	Film/Television & $26$ 				\\ \hline
	Healthcare 		& $88$ 				\\ \hline
	Industrial 		& $26$ 				\\ \hline
	Science 		& $50$ 				\\ \hline
	Service 		& $10$ 				\\ \hline
	Theatre 		& $52$ 				\\ \hline
	Writing 		& $29$ 				\\ \hline
	\hline
	Total			& $461$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the Wikipedia article \url{en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category. We have selected a total of $453$ occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
\end{table}

\section{Results}

\section{Discussion}

\bibliographystyle{unsrt}
\bibliography{main}

\end{document}

### 2017-12-13 gender-bias.py

@@ -1,0 +1,79 @@
+
+from googletrans import Translator
+
+translator = Translator()
+
+def get_gender(occupation,language):
+
+	translation = ''
+
+	if(language == 'Malay'):
+		translation = translator.translate('dia adalah %s' % occupation, src=language, dest='en').text
+	elif(language == 'Estonian'):
+		translation = translator.translate('ta on %s' % occupation, src=language, dest='en').text
+	elif(language == 'Finnish'):
+		translation = translator.translate('hän on %s' % occupation, src=language, dest='en').text
+	elif(language == 'Hungarian'):
+		translation = translator.translate('ő %s' % occupation, src=language, dest='en').text
+	elif(language == 'Armenian'):
+		translation = translator.translate('նա %s է' % occupation, src=language, dest='en').text
+	#else if(language == 'Bengali'):
+	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
+	#else if(language == 'Nepali'):
+	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
+	#elif(language == 'Japanese'):
+	#	translation = translator.translate('あいつ %s' % occupation, src=language, dest='en').text
+	#elif(language == 'Korean'):
+	#	translation = translator.translate('그는 %s' % occupation, src=language, dest='en').text
+	elif(language == 'Turkish'):
+		translation = translator.translate('o bir %s' % occupation, src=language, dest='en').text
+	elif(language == 'Yoruba'):
+		translation = translator.translate('o jẹ %s' % occupation, src=language, dest='en').text
+	elif(language == 'Basque'):
+		translation = translator.translate('%s da' % occupation, src=language, dest='en').text
+	elif(language == 'Swahili'):
+		translation = translator.translate('yeye ni %s' % occupation, src=language, dest='en').text
+	#elif(language == 'Chinese'):
+	#	translation = translator.translate('tā %s' % occupation, src='zh-cn', dest='en').text
+
+	translation = translation.lower()
+
+	if(translation[0:4].find("she") != -1 or translation[0:4].find("she's") != -1):
+		return 'Female'
+	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1):
+		return 'Male'
+	else:
+		return '?'
+
+with open('jobs.csv','r') as jobs:
+	with open('job-genders.csv','w') as output:
+
+		languages = list(map(lambda x: x.rstrip(), jobs.readline().split(';')[2:]))
[…]

### 2017-12-13 script.py

@@ -1,0 +1,49 @@
+
+from googletrans import Translator
+from googletrans import LANGUAGES
+translator = Translator()
+
+# Get language list
+languages = []
+with open('languages.csv', 'r') as f:
+    f.readline()
+    for line in f:
+    	language,family,pronomial_gender_system,supported = line.split(";")
+    	if(pronomial_gender_system != 'Yes' and supported.rstrip() == 'Yes'):
+    		languages.append(language)
+
+# Initialize job categories list
+job_categories = ["artistic","computer","corporate","dance","film-television","healthcare","science","service","theatre","writing"]
+
+with open('jobs.csv', 'w') as output:
+
+	output.write('category')
+	output.write(';english')
+	for language in languages:
+		output.write(';' + language)
+	output.write('\n')
+
+	for category in job_categories:
+		print('Category: %s' % category)
+		with open('jobs/%s-jobs-english.txt' % category,'r') as jobs:
+			i = 0
+			for job in jobs:
+				print('Occupation: %s' % job.rstrip())
+				translations = []
+				for language in languages:
+					try:
+						if language == 'Chinese':
+							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest='zh-cn').text).lower()
+							translations.append(translated_job)
+						else:
+							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest=language).text).lower()
+							translations.append(translated_job)
+					except Exception:
+						print("Could not translate occupation %s to language %s" % (job.rstrip(),language))
+						translations.append("?")
+
+				output.write(category)
+				output.write(';'+job.rstrip().lower())
+				for translation in translations:
+					output.write(';'+translation)
+				output.write('\n')

### 2017-12-14 paper/main.tex
Updated scripts, paper, tables, languages, jobs and added julia script to generate plots

@@
 \documentclass{article}
+
+\usepackage[utf8]{inputenc}
+\usepackage{fontspec}
+\usepackage{xeCJK}
 
@@
 \usepackage{amsmath}
-\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
+%
+%
+%\usepackage{xltxtra}
 
@@
 	\hline
-	Language Family & Language & Pronominal Gender System & Supported \\ \hline
+	Language Family & Language & Pronominal Gender System & Supported \\ \hline \hline
 	\multirow{2}{*}{Austronesian} & Malay & No & Yes \\
@@
 	\hline
-	Category 		& \# Occupations 	\\ \hline
+	Category 		& \# Occupations 	\\ \hline \hline
 	Artistic 		& $102$ 			\\ \hline
@@
 
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|}
+	\hline
+	Language 	& Sentence template \\ \hline \hline
+	Malay		& dia adalah $\langle occupation \rangle$ \\ \hline
+	Estonian	& ta on $\langle occupation \rangle$ \\ \hline
+	Finnish		& hän on $\langle occupation \rangle$ \\ \hline
+	Hungarian	& ő $\langle occupation \rangle$ \\ \hline
+	Armenian	& նա $\langle occupation \rangle$ է \\ \hline
+	Japanese	& は $\langle occupation \rangle$ です\\ \hline
+	Turkish		& o bir $\langle occupation \rangle$ \\ \hline
+	Yoruba		& o jẹ $\langle occupation \rangle$ \\ \hline
+	Basque		& $\langle occupation \rangle$ da \\ \hline
+	Swahili		& yeye ni $\langle occupation \rangle$ \\ \hline
+	Chinese		& ta $\langle occupation \rangle$ \\ \hline
+	\end{tabular}
+	\caption{Templates used to infer gender biases in the translation to the English language.}
+\end{table}
+
 \section{Results}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|}
+	\hline
+	Category & Female & Male & Neutral & Ratio & Total \\ \hline
+	\hline
+	Artistic        & 179    & 504  & 206     & 2.816 	& 918 \\ \hline
+	Computer        & 7      & 125  & 39      & 17.857 	& 171 \\ \hline
+	Corporate       & 36     & 340  & 74      & 9.444 	& 450 \\ \hline
+	Dance           & 31     & 33   & 17      & 1.064 	& 81  \\ \hline
+	Film-television & 54     & 125  & 55      & 2.315 	& 234 \\ \hline
+	Healthcare      & 174    & 475  & 130     & 2.730 	& 792 \\ \hline
+	Science         & 34     & 357  & 59      & 10.500 	& 450 \\ \hline
+	Service         & 5      & 63   & 22      & 12.600 	& 90  \\ \hline
+	Theatre         & 75     & 296  & 106     & 3.947 	& 477 \\ \hline
+	Writing         & 41     & 148  & 72      & 3.610 	& 261 \\ \hline
+	\end{tabular}
+	\label{tab:gender-by-category}
+	\caption{Number of female, male and neutral pronominal genders per occupation category in the translated sentences.
+The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in male dominated fields such as computer science, with up to $\approx 18$ occurrences of male pronouns for each of a female one.}
+\end{table}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/gender-by-category}
+	\label{fig:gender-by-category}
+	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $\approx 73\%$ in computer related jobs and $\approx 76\%$ in corporate jobs respectively.
+Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/gender-by-language}
+	\label{fig:gender-by-language}
+	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $\approx 73\%$ in computer related jobs and $\approx 76\%$ in corporate jobs respectively.
+Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
+\end{figure}
 

### 2017-12-14 gender-bias.py

@@ -1,7 +1,9 @@
 
 from googletrans import Translator
+from xpinyin import Pinyin
 
 translator = Translator()
+p = Pinyin()
 
 def get_gender(occupation,language):
 
@@ -21,8 +23,8 @@ def get_gender(occupation,language):
 	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
 	#else if(language == 'Nepali'):
 	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
-	#elif(language == 'Japanese'):
-	#	translation = translator.translate('あいつ %s' % occupation, src=language, dest='en').text
+	elif(language == 'Japanese'):
+		translation = translator.translate('は %s です' % occupation, src=language, dest='en').text
 	#elif(language == 'Korean'):
 	#	translation = translator.translate('그는 %s' % occupation, src=language, dest='en').text
 	elif(language == 'Turkish'):
@@ -33,23 +35,23 @@ def get_gender(occupation,language):
 		translation = translator.translate('%s da' % occupation, src=language, dest='en').text
 	elif(language == 'Swahili'):
 		translation = translator.translate('yeye ni %s' % occupation, src=language, dest='en').text
-	#elif(language == 'Chinese'):
-	#	translation = translator.translate('tā %s' % occupation, src='zh-cn', dest='en').text
+	elif(language == 'Chinese'):
+		translation = translator.translate('ta %s' % p.get_pinyin(occupation,''), src='zh-cn', dest='en').text
 
 	translation = translation.lower()
 
-	if(translation[0:4].find("she") != -1 or translation[0:4].find("she's") != -1):
+	if(translation[0:4].find("she") != -1 or translation[0:4].find("she's") != -1 or translation[0:4].find("her") != -1):
 		return 'Female'
-	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1):
+	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1 or translation[0:4].find("his") != -1):
 		return 'Male'
 	else:
-		return '?'
+		return 'Neutral'
 
 with open('jobs.csv','r') as jobs:
-	with open('job-genders.csv','w') as output:
+	with open('job-genders-2.csv','w') as output:
 
 		languages = list(map(lambda x: x.rstrip(), jobs.readline().split(';')[2:]))
-		discarded_languages = ['Bengali','Nepali','Japanese','Korean','Chinese']
[…]

### 2017-12-14 stats.jl

@@ -1,0 +1,50 @@
+
+using DataFrames, CSV, Gadfly
+
+# Read job-genders.csv into a Julia DataFrame
+dat = CSV.read("job-genders-2.csv",delim=';',nullable=false)
+
+# Get list of occupation categories
+categories = unique(dat[:Category])
+
+# Initialize results table
+#results = DataFrame(Category=[],Female=[],Male=[],Neutral=[],Ratio=[],Total=[])
+#results = DataFrame(Category=[],Gender=[],Count=[])
+
+#=
+results = DataFrame(Category=[],Gender=[],Count=[])
+for category in categories
+
+    aux = dat[dat[:,:Category] .== category, 3:11]
+
+    female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(aux) ])
+    male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(aux) ])
+    neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(aux) ])
+    total           = sum([ count(x -> true,            aux[language]) for language in names(aux) ]
+
+    #push!(results,[ucfirst(category),female_count,male_count,neutral_count,male_count/female_count,total])
+    push!(results,[ucfirst(category),"Female",  100*female_count/total])
+    push!(results,[ucfirst(category),"Male",    100*male_count/total])
+    push!(results,[ucfirst(category),"Neutral", 100*neutral_count/total])
+end
+=#
+
+results = DataFrame(Language=[],Gender=[],Count=[])
+
+languages = names(dat)[3:end]
+for language in languages
+    female_count    = count(x -> x=="Female", dat[language])
+    male_count      = count(x -> x=="Male", dat[language])
+    neutral_count   = count(x -> x=="Neutral", dat[language])
+    total           = count(x -> true, dat[language])
+    push!(results,[language,"Female",   100*female_count/total    ])
+    push!(results,[language,"Male",     100*male_count/total      ])
+    push!(results,[language,"Neutral",  100*neutral_count/total   ])
+end
+
+println(results)
+#p = plot(results, x="Category", y="Count", color="Gender", Coord.Cartesian(ymin=0,ymax=100), Geom.bar(position=:stack), Theme(background_color="white",grid_color="gray",bar_highlight=colorant"black"), Guide.ylabel("\%"))
[…]

### 2017-12-15 paper/main.tex
Updated scripts for generating plots, added plots to paper/pictures and updated main.tex

@@
 \usepackage[utf8]{inputenc}
-\usepackage{fontspec}
+%\usepackage{fontspec}
 \usepackage{xeCJK}
@@
 	Writing         & 41     & 148  & 72      & 3.610 	& 261 \\ \hline
+	\hline
+	Total           & 636    & 2466 & 780     & 3.877   &    \\ \hline
 	\end{tabular}
@@
 	\label{fig:gender-by-category}
-	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $\approx 73\%$ in computer related jobs and $\approx 76\%$ in corporate jobs respectively.
+	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively.
 Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
 \end{figure}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|}
+	\hline
+	Language & Female & Male & Neutral & Ratio & Total \\ \hline
+	\hline
+	Malay     & 43     & 392  & 0       & 9.116 	& 436   \\ \hline
+	Estonian  & 121    & 309  & 0       & 2.554 	& 436   \\ \hline
+	Finnish   & 167    & 263  & 0       & 1.575 	& 436   \\ \hline
+	Hungarian & 174    & 255  & 2       & 1.465 	& 436   \\ \hline
+	Armenian  & 94     & 337  & 1       & 3.585 	& 436   \\ \hline
+	Japanese  & 2      & 207  & 222     & 103.500   & 436   \\ \hline
+	Turkish   & 19     & 368  & 44      & 19.368 	& 436   \\ \hline
+	Yoruba    & 11     & 290  & 131     & 26.364 	& 436   \\ \hline
+	Basque    & 5      & 45   & 380     & 9.000		& 436   \\ \hline
+	Swahili   & 68     & 363  & 0       & 5.338 	& 436   \\ \hline
+	Chinese   & 13     & 359  & 58      & 27.615 	& 436   \\ \hline
+	\hline
+	Total 	  & 717    & 3188 & 838     & 4.446     &   \\ \hline
+	\end{tabular}
+	\label{tab:gender-by-category}
+	\caption{Number of female, male and neutral pronominal genders per language in the translated sentences.
+The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in some languages such as Chinese, with almost 30 male pronouns for each female one.}
+\end{table}
 
@@
 	\label{fig:gender-by-language}
-	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $\approx 73\%$ in computer related jobs and $\approx 76\%$ in corporate jobs respectively.
-Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
+	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively.
+Some languages such as Japanese (and particularly Basque) were observed to yield a high number of neutral pronouns, but that is the exception rather than the rule among the tested idioms.
+Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences.}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/scatterplot-languages}
+	\label{fig:scatterplot-languages}
+	\caption{Scatter plot of translated sentences' statistics.
+Each point (color coded according to its category) corresponds to a single occupation, of which the sex ratio and the percentage of gender neutral pronouns are averaged over all tested languages (Malay, Estonian, Finnish, Hungarian, Armenian, Japanese, Turkish, Yoruba, Basque, Swahili and Chinese).}
 \end{figure}

### 2017-12-15 stats.jl

@@ -4,15 +4,28 @@ using DataFrames, CSV, Gadfly
 # Read job-genders.csv into a Julia DataFrame
 dat = CSV.read("job-genders-2.csv",delim=';',nullable=false)
 
-# Get list of occupation categories
-categories = unique(dat[:Category])
+# Compute results table (grouped by occupation, averaged among languages)
+results_by_occupation = DataFrame(Occupation=[],Category=[],Ratio=[],Neutral=[])
+for occupation in dat[:Occupation]
 
-# Initialize results table
-#results = DataFrame(Category=[],Female=[],Male=[],Neutral=[],Ratio=[],Total=[])
-#results = DataFrame(Category=[],Gender=[],Count=[])
+    aux = dat[ dat[:,:Occupation] .== occupation, : ]
+    
+    female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat)[3:11] ])
+    male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat)[3:11] ])
+    neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat)[3:11] ])
+    ratio           = male_count/female_count
+
+    push!(results_by_occupation, [occupation,ucfirst(aux[1,:Category]),ratio,100*neutral_count/(11-3+1)])
+end
+# Save scatter plot
+p = plot(results_by_occupation, x=:Ratio, y=:Neutral, color=:Category, Guide.xlabel("Sex Ratio"), Guide.ylabel("\% of Gender Neutral Pronouns"), Stat.x_jitter(range=2), Stat.y_jitter(range=2), Coord.Cartesian(xmin=0, xmax=10, ymin=0), Geom.point)
+draw(PDF("paper/pictures/scatterplot-languages.pdf", 5inch, 3.75inch),p)
 
 #=
-results = DataFrame(Category=[],Gender=[],Count=[])
+# Compute results table (grouped by category)
+# Get list of occupation categories
+categories = unique(dat[:Category])
+results_by_category = DataFrame(Category=[],Gender=[],Count=[])
 for category in categories
 
     aux = dat[dat[:,:Category] .== category, 3:11]
@@ -20,31 +33,31 @@ for category in categories
     female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(aux) ])
     male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(aux) ])
[…]

### 2017-12-15 paper/main.tex
Added 'industrial' category and wrote more in the paper

@@
 We believe that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
-Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse}.
-The same structure can be
+Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} in Google Translate.
+The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
+Given the success of Google Translate, which amounts to 200 million users daily, we have decided to exploit its API to obtain the desired thermometer of gender bias.
+Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they 1) exhibit a pronominal gender system and 2) are supported by Google Translate.
+Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
+
+There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}.
+To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
+Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) is shown in Table \ref{tab:occupations}.
+Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
 
@@
 	& Bengali & No & Yes \\
-	& English & \textbf{Yes} & Yes \\
+	& English & \textbf{Yes} & Yes \\{}
 	& Persian & \textbf{Yes} & Yes \\ \hline
@@
 	\end{tabular}
-	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+	\label{tab:gender-neutral-languages}
+	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
 Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system.
@@
 	\hline
-	Total			& $461$				\\ \hline
+	Total			& $436$				\\ \hline
 	\end{tabular}
-	\caption{Selected occupations obtained from the Wikipedia article \url{en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category.
+	\label{tab:occupations}
+	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category.
 We have selected a total of $453$ occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
@@
 	Armenian	& նա $\langle occupation \rangle$ է \\ \hline
-	Japanese	& は $\langle occupation \rangle$ です\\ \hline
+	Japanese	& $\langle occupation \rangle$ です\\ \hline
 	Turkish		& o bir $\langle occupation \rangle$ \\ \hline
@@
 	\end{tabular}
+	\label{tab:templates}
 	\caption{Templates used to infer gender biases in the translation to the English language.}
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	\hline
+	stagehands & author & neurologist \\ \hline
+	screenwriter & animator & marketing director \\ \hline
+	biochemist & endocrinologist & freelancer \\ \hline
+	neurosurgeon & computer scientist & petrochemical engineer \\ \hline
+	food stylist & cardiothoracic surgeon & property master \\ \hline
+	literary editor & video editor & animation director \\ \hline
+	house manager & chief administrative officer & arts administration \\ \hline
+	actor & dialysis technician & family nurse practitioner \\ \hline
+	psychologist & chief creative officer & flash developer \\ \hline
+	scenic artist & producer & medical laboratory scientist \\ \hline
+	\end{tabular}
+	\label{tab:occupations-examples}
+	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 436 different occupations.}
 \end{table}

### 2017-12-15 gender-bias.py

@@ -48,7 +48,7 @@ def get_gender(occupation,language):
 		return 'Neutral'
 
 with open('jobs.csv','r') as jobs:
-	with open('job-genders-2.csv','w') as output:
+	with open('job-genders.csv','w') as output:
 
 		languages = list(map(lambda x: x.rstrip(), jobs.readline().split(';')[2:]))
 		discarded_languages = ['Bengali','Nepali','Korean']
@@ -64,16 +64,18 @@ with open('jobs.csv','r') as jobs:
 		# For each job, translate into each language and fetch the corresponding gender
 		for line in jobs:
 			category = line.split(';')[0]
-			english_name = line.split(';')[1]
+			occupation = line.split(';')[1]
 			foreign_names = list(map(lambda x: x.rstrip(), line.split(';')[2:]))
 
+			print(occupation)
+
 			output.write('%s' % category)
-			output.write(';%s' % english_name)
+			output.write(';%s' % occupation)
 	
-			for (language,occupation) in zip(languages,foreign_names):
+			for (language,foreign_name) in zip(languages,foreign_names):
 				if(language not in discarded_languages):
 					try:
-						gender = get_gender(occupation,language)
+						gender = get_gender(foreign_name,language)
 						output.write(';%s' % gender)
 					except ValueError:
 						output.write(';?')

### 2017-12-15 sample-jobs.jl

@@ -1,0 +1,20 @@
+
+using DataFrames, CSV, StatsBase
+
+dat = CSV.read("jobs.csv",delim=';',nullable=false)
+
+n,m = 10,3
+occupations = sample(dat[:Occupation],n*m,replace=false)
+
+output = open("test.txt","w")
+for i=1:n
+    for j=1:m
+        if j > 1
+            print(output,"& $(occupations[m*(i-1)+j]) ")
+        else
+            print(output,"$(occupations[m*(i-1)+j]) ")
+        end
+    end
+    println(output,"\\\\ \\hline")
+end
+close(output)

### 2017-12-15 script.py

@@ -13,9 +13,9 @@ with open('languages.csv', 'r') as f:
     		languages.append(language)
 
 # Initialize job categories list
-job_categories = ["artistic","computer","corporate","dance","film-television","healthcare","science","service","theatre","writing"]
+job_categories = ["artistic","computer","corporate","dance","film-television","healthcare","industrial","science","service","theatre","writing"]
 
-with open('jobs.csv', 'w') as output:
+with open('jobs-2.csv', 'w') as output:
 
 	output.write('category')
 	output.write(';english')

### 2017-12-15 stats.jl

@@ -2,7 +2,7 @@
 using DataFrames, CSV, Gadfly
 
 # Read job-genders.csv into a Julia DataFrame
-dat = CSV.read("job-genders-2.csv",delim=';',nullable=false)
+dat = CSV.read("job-genders.csv",delim=';',nullable=false)
 
 # Compute results table (grouped by occupation, averaged among languages)
 results_by_occupation = DataFrame(Occupation=[],Category=[],Ratio=[],Neutral=[])

### 2017-12-15 paper/main.tex
Updates on paper

@@
 \usepackage{graphicx}
-%
-%
-%\usepackage{xltxtra}
+\usepackage{amssymb}
+\usepackage{pifont}
+
+\newcommand{\cmark}{\ding{51}}%
+\newcommand{\xmark}{\ding{53}}%
 
@@
 We believe that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
-Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} in Google Translate.
+Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate.
 The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
 Given the success of Google Translate, which amounts to 200 million users daily, we have decided to exploit its API to obtain the desired thermometer of gender bias.
-Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}).
 Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they 1) exhibit a pronominal gender system and 2) are supported by Google Translate.
 Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
+Following difficulties with Bengali, Nepali and Korean we have decided not to work with these languages also.
 
@@
 To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) is shown in Table \ref{tab:occupations}.
+Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) are shown in Table \ref{tab:occupations}.
 Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
@@
 	\centering
-	\begin{tabular}{|c|c|m{3cm}|c|}
+	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
 	\hline
-	Language Family & Language & Pronominal Gender System & Supported \\ \hline \hline
-	\multirow{2}{*}{Austronesian} & Malay & No & Yes \\
-	& Tagalog & No & \textbf{No} \\ \hline
-	\multirow{3}{*}{Finno-Ugric} & Estonian & No & Yes \\
-	& Finnish & No & Yes\\
-	& Hungarian & No & Yes \\ \hline
-	\multirow{4}{*}{Indo-European} & Armenian & No & Yes \\
-	& Bengali & No & Yes \\
-	& English & \textbf{Yes} & Yes \\{}
-	& Persian & \textbf{Yes} & Yes \\ \hline
-	\multirow{3}{*}{Indo-Aryan} & Maithili & No & \textbf{No} \\
-	& Nepali & No & Yes \\
-	& Oriya & No & \textbf{No} \\ \hline
-	\multirow{10}{*}{} & Japanese & No & Yes \\
-	& Korean & \textbf{Optional} & Yes \\
-	& Turkish & No & Yes \\
-	& Yoruba & No & Yes \\
-	& Basque & No & Yes \\
-	& Swahili & No & Yes \\
-	& Chinese & \textbf{Optional} & Yes \\
-	& Cantonese & No & \textbf{No} \\
-	& Pipil & No & \textbf{No} \\
-	& Quechuan & No & \textbf{No} \\ \hline
-	\multirow{4}{*}{Constructed} & Esperanto & \textbf{Yes} & Yes \\
-	& Ido & \textbf{Optional} & \textbf{No} \\
-	& Lingua Franca Nova & No & \textbf{No} \\
-	& Interlingua & No & \textbf{No} \\ \hline
+	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
+	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
+									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
+	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
+									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
+	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
+									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
+									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
+	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
+									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
+									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
+	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Korean 				& O 				& \checkmark 	& \xmark		\\
+									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
+									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
+									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
+									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
+									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
+	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
+									& Ido 					& O 				& \xmark 		& \xmark 		\\
+									\cline{2-2}
+									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
+									\cline{2-2}
+									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
 	\end{tabular}
+	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
+For the purposes of this work, we have decided to work only with languages lacking such a system, and as such Persian and Esperanto have been discarded.
+Languages lacking support from Google Translate have been discarded.
+Following difficulties with Bengali, Nepali and Korean, these languages have also been discarded.}
 	\label{tab:gender-neutral-languages}
-	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
-Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system.
-For the purposes of this work, we have decided to work only with languages lacking such a system, and as such Persian and Esperanto have been discarded.
-Because Tagalog, Maithili, Oriya, Cantonese, Pipil, Quechuan, Lingua Franca Nova and Interlingua lack support from Google Translate, these languages have also been ommitted from this work.}
 \end{table}
@@
 	\hline
-	Total			& $436$				\\ \hline
+	Total			& $462$				\\ \hline
 	\end{tabular}
+	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category.
+We have selected a total of 462 occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
 	\label{tab:occupations}
-	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category.
-We have selected a total of $453$ occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
 \end{table}
@@
 	\end{tabular}
+	\caption{Templates used to infer gender biases in the translation to the English language.}
 	\label{tab:templates}
-	\caption{Templates used to infer gender biases in the translation to the English language.}
 \end{table}
@@
 	\end{tabular}
+	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 462 different occupations.}
 	\label{tab:occupations-examples}
-	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 436 different occupations.}
 \end{table}
@@
 \section{Results}
+
+For each one of the tested 462 occupations (see Tables \ref{tab:occupations}, \ref{tab:occupations-examples}), we used the Python Google Translate API (\url{http://py-googletrans.readthedocs.io/en/latest/}) to translate sentences built with the templates in Table \ref{tab:templates} from each one of the tested languages in Table \ref{tab:gender-neutral-languages} to English.
+The resulting sentences are then classified as \emph{female}, \emph{male} or \emph{neutral} according to their respective pronouns.
+Sentences starting with \say{She/She's/Her} are classified as female, sentences starting with \say{He/He's/His} are classified as male and sentences starting with \say{It/It's/Its/They/They're/Their} are classified as (gender) neutral.
+The results from this analysis, which can be found in \url{https://github.com/marceloprates/Gender-Bias}, are further discussed below.
+
+One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
+Sentences about occupations from the \emph{Corporate} and \emph{Science} category are also disproportionately translated with male pronouns (sex ratios $9.444$ and $10.5$ respectively), while those containing occupations from the \emph{Dance} category achieve a sex ratio of almost one ($1.064$).
+Not one category has achieved a balanced sex ratio, neither does any category exhibit more gender neutral than male pronouns.
+In total, female pronouns add up to $16.522\%$ among all categories, while male pronouns add up to $63.083\%$ and gender neutral pronouns to just $7.912\%$, yielding an average sex ratio of $3.818$.
 
@@
 	\hline
-	Artistic        & 179    & 504  & 206     & 2.816 	& 918 \\ \hline
-	Computer        & 7      & 125  & 39      & 17.857 	& 171 \\ \hline
-	Corporate       & 36     & 340  & 74      & 9.444 	& 450 \\ \hline
-	Dance           & 31     & 33   & 17      & 1.064 	& 81  \\ \hline
-	Film-television & 54     & 125  & 55      & 2.315 	& 234 \\ \hline
-	Healthcare      & 174    & 475  & 130     & 2.730 	& 792 \\ \hline
-	Science         & 34     & 357  & 59      & 10.500 	& 450 \\ \hline
-	Service         & 5      & 63   & 22      & 12.600 	& 90  \\ \hline
-	Theatre         & 75     & 296  & 106     & 3.947 	& 477 \\ \hline
-	Writing         & 41     & 148  & 72      & 3.610 	& 261 \\ \hline
+	Artistic        & 188 & 518 & 72 & 2.755 	& 918 \\ \hline
+	Computer        & 7   & 125 & 12 & 17.857 	& 171 \\ \hline
+	Corporate       & 36  & 340 & 23 & 9.444 	& 450 \\ \hline
+	Dance           & 31  & 33  & 8  & 1.064 	& 81  \\ \hline
+	Film-television & 54  & 125 & 18 & 2.315 	& 234 \\ \hline
+	Healthcare      & 176 & 483 & 64 & 2.744 	& 792 \\ \hline
+	Industrial      & 40  & 135 & 29 & 3.375   	& 234 \\ \hline
+	Science         & 34  & 357 & 25 & 10.500	& 450 \\ \hline
+	Service         & 5   & 63  & 10 & 12.600	& 90  \\ \hline
+	Theatre         & 75  & 296 & 38 & 3.947 	& 477 \\ \hline
+	Writing         & 41  & 148 & 30 & 3.610 	& 261 \\ \hline
 	\hline
-	Total           & 636    & 2466 & 780     & 3.877   &    \\ \hline
+	Total           & 687 & 2623 & 329 & 3.818 & 4158 \\ \hline
 	\end{tabular}
@@
 
+While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each different culture has on this issue.
+Table \ref{tab:gender-by-language} for instance shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns.
+Figure \ref{fig:gender-by-language} shows that, when grouping by language, gender neutral pronouns can be more prominent than male pronouns at least in one case: translating sentences from Basque yields 153 neutral vs 50 male and 5 female pronouns.
+Unfortunately this is the exception rather than the rule, with Yoruba following after with 131 neutral vs 304 male and 12 female pronouns.
+In total, female pronouns add up to $18.687\%$ among all categories, while male pronouns add up to $81.626\%$ and gender neutral pronouns to $8.466\%$, yielding an average sex ratio of $4.368$ ($14.405\%$ larger than what we get from grouping among categories).
+It should be noted however that Japanese and Basque, the two languages which stood out from the behavior observed in Figure \ref{fig:gender-by-category}, are precisely the two that Google Translate found hardest to translate.
+These findings should, as a result, be taken with a grain of salt.
+
 \begin{table}[H]
@@
 	\hline
-	Malay     & 43     & 392  & 0       & 9.116 	& 436   \\ \hline
-	Estonian  & 121    & 309  & 0       & 2.554 	& 436   \\ \hline
-	Finnish   & 167    & 263  & 0       & 1.575 	& 436   \\ \hline
-	Hungarian & 174    & 255  & 2       & 1.465 	& 436   \\ \hline
-	Armenian  & 94     & 337  & 1       & 3.585 	& 436   \\ \hline
-	Japanese  & 2      & 207  & 222     & 103.500   & 436   \\ \hline
-	Turkish   & 19     & 368  & 44      & 19.368 	& 436   \\ \hline
-	Yoruba    & 11     & 290  & 131     & 26.364 	& 436   \\ \hline
-	Basque    & 5      & 45   & 380     & 9.000		& 436   \\ \hline
-	Swahili   & 68     & 363  & 0       & 5.338 	& 436   \\ \hline
-	Chinese   & 13     & 359  & 58      & 27.615 	& 436   \\ \hline
+	Malay     & 47  & 415  & 0   & 8.830 	& 462 \\ \hline
+	Estonian  & 130 & 332  & 0   & 2.554 	& 462 \\ \hline
+	Finnish   & 179 & 283  & 0   & 1.581 	& 462 \\ \hline
+	Hungarian & 189 & 270  & 1   & 1.429 	& 462 \\ \hline
+	Armenian  & 101 & 360  & 1   & 3.564 	& 462 \\ \hline
+	Japanese  & 2   & 215  & 0   & 107.500  & 462 \\ \hline
+	Turkish   & 22  & 394  & 43  & 17.909	& 462 \\ \hline
+	Yoruba    & 12  & 304  & 131 & 25.333	& 462 \\ \hline
+	Basque    & 5   & 50   & 153 & 10.000   & 462 \\ \hline
+	Swahili   & 76  & 386  & 0   & 5.079 	& 462 \\ \hline
+	Chinese   & 14  & 385  & 23  & 27.500   & 462 \\ \hline
 	\hline
-	Total 	  & 717    & 3188 & 838     & 4.446     &   \\ \hline
+	Total     & 777 & 3394 & 352 & 4.368	& 4158   \\ \hline
 	\end{tabular}
-	\label{tab:gender-by-category}
 	\caption{Number of female, male and neutral pronominal genders per language in the translated sentences.
 The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in some languages such as Chinese, with almost 30 male pronouns for each female one.}
+	\label{tab:gender-by-language}
 \end{table}
@@
 	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively.
-Some languages such as Japanese (and particularly Basque) were observed to yield a high number of neutral pronouns, but that is the exception rather than the rule among the tested idioms.
-Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences.}
+Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
+Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns.}
 \end{figure}
+
+Instead of grouping sentences either by category or language, we can also visualize each of them individually on a scatter plot.
+Figure \ref{fig:scatterplot-languages} shows each occupation as a point on a bi-dimensional lattice, arranged horizontally by their sex ratio and vertically by the proportion of gender neutral pronouns, both averaged over translations from all tested languages.
+Each point is also color coded according to that occupation's respective category (Artistic, Writing, Science, etc.).
 

### 2017-12-16 paper/main.tex
Wrapping up

@@
 With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation.
-We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English, as Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies.
+We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English.
+As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach
 

### 2017-12-16 gender-bias.py

@@ -44,8 +44,10 @@ def get_gender(occupation,language):
 		return 'Female'
 	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1 or translation[0:4].find("his") != -1):
 		return 'Male'
-	else:
+	elif(translation[0:4].find("it") != -1 or translation[0:4].find("it's") != -1 or translation[0:4].find("its") != -1 or translation[0:7].find("they") != -1 or translation[0:7].find("they're") != -1 or translation[0:4].find("them") != -1):
 		return 'Neutral'
+	else:
+		return '?'
 
 with open('jobs.csv','r') as jobs:
 	with open('job-genders.csv','w') as output:
@@ -54,12 +56,12 @@ with open('jobs.csv','r') as jobs:
 		discarded_languages = ['Bengali','Nepali','Korean']
 
 		# Write header
-		output.write('category')
-		output.write(';English')
-		output.write('\n')
+		output.write('Category')
+		output.write(';Occupation')
 		for language in languages:
 			if language not in discarded_languages:
 				output.write(';%s' % language)
+		output.write('\n')
 	
 		# For each job, translate into each language and fetch the corresponding gender
 		for line in jobs:

### 2017-12-16 sample-jobs.jl

@@ -3,6 +3,11 @@ using DataFrames, CSV, StatsBase
 
 dat = CSV.read("jobs.csv",delim=';',nullable=false)
 
+for category in unique(dat[:Category])
+    println("$category $( length(dat[ dat[:,:Category] .== category, :Occupation ]) )")
+end
+
+#=
 n,m = 10,3
 occupations = sample(dat[:Occupation],n*m,replace=false)
 
@@ -18,3 +23,4 @@ for i=1:n
     println(output,"\\\\ \\hline")
 end
 close(output)
+=#

### 2017-12-16 stats.jl

@@ -10,22 +10,23 @@ for occupation in dat[:Occupation]
 
     aux = dat[ dat[:,:Occupation] .== occupation, : ]
     
-    female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat)[3:11] ])
-    male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat)[3:11] ])
-    neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat)[3:11] ])
+    female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat)[3:end] ])
+    male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat)[3:end] ])
+    neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat)[3:end] ])
     ratio           = male_count/female_count
 
     push!(results_by_occupation, [occupation,ucfirst(aux[1,:Category]),ratio,100*neutral_count/(11-3+1)])
 end
 # Save scatter plot
-p = plot(results_by_occupation, x=:Ratio, y=:Neutral, color=:Category, Guide.xlabel("Sex Ratio"), Guide.ylabel("\% of Gender Neutral Pronouns"), Stat.x_jitter(range=2), Stat.y_jitter(range=2), Coord.Cartesian(xmin=0, xmax=10, ymin=0), Geom.point)
+p = plot(results_by_occupation, x=:Ratio, y=:Neutral, color=:Category, Guide.xlabel("Sex Ratio"), Guide.ylabel("\% of Gender Neutral Pronouns"), Stat.x_jitter(range=2), Stat.y_jitter(range=2), Coord.Cartesian(xmin=0, ymin=0), Geom.point)
 draw(PDF("paper/pictures/scatterplot-languages.pdf", 5inch, 3.75inch),p)
 
 #=
 # Compute results table (grouped by category)
 # Get list of occupation categories
 categories = unique(dat[:Category])
-results_by_category = DataFrame(Category=[],Gender=[],Count=[])
+#results_by_category = DataFrame(Category=[],Gender=[],Count=[])
+results_by_category = DataFrame(Category=[],Female=[],Male=[],Neutral=[],Ratio=[],Total=[])
 for category in categories
 
     aux = dat[dat[:,:Category] .== category, 3:11]
@@ -35,16 +36,21 @@ for category in categories
[…]

### 2017-12-17 paper/main.tex
Updated paper

@@
 
-\title{Assessing Gender Bias in Machine Translation}
+\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
 \author{Marcelo Prates \and Luis Lamb}
@@
 \begin{abstract}
+In recent times there has been a growing concern both inside and outside academia about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of their own prejudices.
+A large number of AI tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas.
+Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI.
+In this paper, we construct sentences with gender neutral pronouns in different languages which support them and translate these sentences into English using the Google Translate API.
+The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate assumes for each sentence.
+We show that not only there is a tendency towards translating pronouns to \say{he} but also that this tendency is exaggerated for sentences about male dominated fields such as computer science.
+At the other hand, sentences about artistic fields are less likely to yield male defaults.
+This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of those people behind its training dataset -- ourselves.
 \end{abstract}
@@
 Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}.
+Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}.
 \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
@@
 We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English.
-As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach
+As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.
 
@@
 \section{Motivation}
+
+If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights.
+One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others.
+We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market.
+We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.
+
+If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models.
+Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed.
+This poses an ethical rather than a technical limitation to statistical machine translation and big data statistical models in general.
 
@@
 Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
+Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.
 
@@
 
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|}
+	\hline
+	happy 		& sad 		& right 	\\ \hline
+	wrong 		& afraid	& brave 	\\ \hline
+	smart		& dumb		& proud 	\\ \hline
+	ashamed		& kind		& cruel 	\\ \hline
+	envious		& loving	& hateful 	\\ \hline
+	modest 		& arrogant	& guilty	\\ \hline
+	innocent	& helpless	& shy		\\ \hline
+	\end{tabular}
+	\caption{Adjectives}
+	\label{tab:adjectives}
+\end{table}
+
 \section{Results}
@@
 
-While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each different culture has on this issue.
-Table \ref{tab:gender-by-language} for instance shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns.
+While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each culture possibly has on this issue.
+Table \ref{tab:gender-by-language} shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns.
 Figure \ref{fig:gender-by-language} shows that, when grouping by language, gender neutral pronouns can be more prominent than male pronouns at least in one case: translating sentences from Basque yields 153 neutral vs 50 male and 5 female pronouns.
@@
 
+Figures \ref{fig:histogram-female}, \ref{fig:histogram-female} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns.
+While the number of occurences of male pronouns tends towards a normal distribution, the figure is changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation.
+The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/histogram-female}
+	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0.
+The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
+	\label{fig:histogram-female}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/histogram-male}
+	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9.
+In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
+	\label{fig:histogram-male}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/histogram-neutral}
+	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
+	\label{fig:histogram-neutral}
+\end{figure}
+
+Finally, Figure \ref{fig:barplot-adjectives} and Table \ref{tab:gender-by-adjective} show how stereotypical gender roles possibly play a part when simple adjectives are translated.
+We observed that objective statements such as \say{he/she is wrong/right/guilty/innocent} are biased towards male defaults, while statements concerning emotional states (\say{sad},\say{kind},\say{shy}) amass at the opposite extreme of the sex ratio spectrum.
+Unsurprisingly, the statement \say{he/she is attractive} is translated predominately with female pronouns.
+
+\begin{table}[H]
+	\centering
+	\begin{tabular}{|c|c|c|c|c|c|}
+	\hline
+	Adjective 	& Female & Male & Neutral & Ratio & Total \\ \hline
+	\hline
+	Shy     	& 6  & 2   & 2  & 0.333 	& 12  \\ \hline
+	Attractive 	& 4  & 2   & 4  & 0.500   	& 12  \\ \hline
+	Happy   	& 5  & 3   & 2  & 0.600   	& 12  \\ \hline
+	Kind    	& 4  & 3   & 1  & 0.750 	& 12  \\ \hline
+	Ashamed 	& 4  & 5   & 1  & 1.250 	& 12  \\ \hline
+	Smart   	& 2  & 5   & 3  & 2.500   	& 12  \\ \hline
+	Envious 	& 2  & 6   & 1  & 3.000   	& 12  \\ \hline
+	Sad     	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
+	Loving  	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
+	Helpless 	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
+	Brave   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
+	Proud   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
+	Hateful 	& 1  & 5   & 3  & 5.000   	& 12  \\ \hline
+	Dumb    	& 1  & 6   & 2  & 6.000   	& 12  \\ \hline
+	Innocent 	& 1  & 8   & 1  & 8.000   	& 12  \\ \hline
+	Right   	& 0  & 7   & 3  & -   		& 12  \\ \hline
+	\hline
+	Total   	& 43 & 129 & 41 & 3     	& 264 \\ \hline
+	\end{tabular}
+	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
+	\label{tab:gender-by-adjective}
+\end{table}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{pictures/barplot-adjectives}
+	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
+One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns.
+Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
+	\label{fig:barplot-adjectives}
+\end{figure}
+
 \section{Discussion}
 
+In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them.
+As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society.
+By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process.
+Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis.
+Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{Guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
+We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese.
+Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.
+
+We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it arouses discussions about the role of AI engineers on minimizing the harmful effects of the pressing modern concern of machine bias.
+
+\clearpage
 \bibliographystyle{unsrt}

### 2017-12-17 gender-bias.py

@@ -19,14 +19,8 @@ def get_gender(occupation,language):
 		translation = translator.translate('ő %s' % occupation, src=language, dest='en').text
 	elif(language == 'Armenian'):
 		translation = translator.translate('նա %s է' % occupation, src=language, dest='en').text
-	#else if(language == 'Bengali'):
-	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
-	#else if(language == 'Nepali'):
-	#	translation = translator.translate('dia adalah %s' % occupation,dest='en').text
 	elif(language == 'Japanese'):
 		translation = translator.translate('は %s です' % occupation, src=language, dest='en').text
-	#elif(language == 'Korean'):
-	#	translation = translator.translate('그는 %s' % occupation, src=language, dest='en').text
 	elif(language == 'Turkish'):
 		translation = translator.translate('o bir %s' % occupation, src=language, dest='en').text
 	elif(language == 'Yoruba'):
@@ -49,10 +43,10 @@ def get_gender(occupation,language):
 	else:
 		return '?'
 
-with open('jobs.csv','r') as jobs:
+with open('jobs.csv','r') as adjectives:
 	with open('job-genders.csv','w') as output:
 
-		languages = list(map(lambda x: x.rstrip(), jobs.readline().split(';')[2:]))
+		languages = list(map(lambda x: x.rstrip(), adjectives.readline().split(';')[1:]))
 		discarded_languages = ['Bengali','Nepali','Korean']
 
 		# Write header
@@ -64,15 +58,15 @@ with open('jobs.csv','r') as jobs:
 		output.write('\n')
 	
 		# For each job, translate into each language and fetch the corresponding gender
-		for line in jobs:
+		for line in adjectives:
 			category = line.split(';')[0]
 			occupation = line.split(';')[1]
-			foreign_names = list(map(lambda x: x.rstrip(), line.split(';')[2:]))
-
-			print(occupation)
+			foreign_names = list(map(lambda x: x.rstrip(), line.split(';')[1:]))
 
 			output.write('%s' % category)
 			output.write(';%s' % occupation)
+
+			output.write('%s' % english_name)
 	
 			for (language,foreign_name) in zip(languages,foreign_names):
 				if(language not in discarded_languages):

### 2017-12-17 stats.jl

@@ -2,71 +2,105 @@
 using DataFrames, CSV, Gadfly
 
 # Read job-genders.csv into a Julia DataFrame
-dat = CSV.read("job-genders.csv",delim=';',nullable=false)
+dat_jobs = CSV.read("job-genders.csv",delim=';',nullable=false)
+# Read adjective-genders.csv into a Julia DataFrame
+dat_adjectives = CSV.read("adjective-genders.csv",delim=';',nullable=false)
 
-# Compute results table (grouped by occupation, averaged among languages)
-results_by_occupation = DataFrame(Occupation=[],Category=[],Ratio=[],Neutral=[])
-for occupation in dat[:Occupation]
+function barplot_adjectives()
+    results_by_adjective = DataFrame(Adjective=[],Gender=[],Count=[],Ratio=[])
 
-    aux = dat[ dat[:,:Occupation] .== occupation, : ]
-    
-    female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat)[3:end] ])
-    male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat)[3:end] ])
-    neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat)[3:end] ])
-    ratio           = male_count/female_count
+    for adjective in dat_adjectives[:Adjective]
+        aux = dat_adjectives[ dat_adjectives[:,:Adjective] .== adjective, : ]
+        
+        female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat_adjectives)[1:end] ])
+        male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat_adjectives)[1:end] ])
+        neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat_adjectives)[1:end] ])
+        total           = sum([ count(x -> true,            aux[language]) for language in names(dat_adjectives)[1:end] ])
+        ratio = male_count/female_count
+        
+        push!(results_by_adjective, [ucfirst(adjective),"Female",   100*female_count/total,   ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Male",     100*male_count/total,     ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  100*neutral_count/total,  ratio])
+    end
 
[…]

### 2017-12-18 paper/main.tex
Fixed problems with references

@@
 Research was almost completely abandoned in the United States, making a shy re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical}.
-Statistical and example-based machine translation have been on the rise ever since \cite{carl2003recent}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily \cite{bengio2015deep}.
+Statistical and example-based machine translation have been on the rise ever since \cite{carl2003recent}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google,gtranslateneuralnet2017}) amounting to over $200$ million users daily \cite{gtranslateneuralnet2017,gtranslatezeroshot2017}.
 
@@
 \clearpage
-\bibliographystyle{unsrt}
+\bibliographystyle{alphadin}
 \bibliography{main}

### 2017-12-18 paper/main.tex
Fixed references

@@
 \usepackage{xeCJK}
+%\usepackage{armtex}
 
@@
 	Hungarian	& ő $\langle occupation \rangle$ \\ \hline
-	Armenian	& նա $\langle occupation \rangle$ է \\ \hline
+	Armenian	& na $\langle occupation \rangle$ e \\ \hline
 	Japanese	& $\langle occupation \rangle$ です\\ \hline
@@
 	\includegraphics[width=\linewidth]{pictures/gender-by-language}
-	\label{fig:gender-by-language}
 	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively.
@@
 Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns.}
+	\label{fig:gender-by-language}
 \end{figure}
@@
 	\includegraphics[width=\linewidth]{pictures/scatterplot-languages}
-	\label{fig:scatterplot-languages}
 	\caption{Scatter plot of translated sentences' statistics.
 Each point (color coded according to its category) corresponds to a single occupation, of which the sex ratio and the percentage of gender neutral pronouns are averaged over all tested languages (Malay, Estonian, Finnish, Hungarian, Armenian, Japanese, Turkish, Yoruba, Basque, Swahili and Chinese).}
+	\label{fig:scatterplot-languages}
 \end{figure}

### 2017-12-19 paper/main.tex
Fixed references

@@
 
+\usepackage[numbers]{natbib}
+\bibliographystyle{plainnat}
+
 \title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
@@
 
-Although the idea of automated translation can in principle be traced back to as long as the 17th century with René Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by Warren Weaver \cite{weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
+Although the idea of automated translation can in principle be traced back to as long as the 17th century with René Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by Warren Weaver \cite{locke1955machine} discussing the possibility of employing digital computers to perform automated translation.
 The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}.
@@
 \clearpage
-\bibliographystyle{alphadin}
+
+%\bibliographystyle{di}
 \bibliography{main}

### 2017-12-25 paper/main.tex
Update main.tex
Updated caption for scatterplot figure

@@
 	\includegraphics[width=\linewidth]{pictures/scatterplot-languages}
-	\caption{Scatter plot of translated sentences' statistics.
-Each point (color coded according to its category) corresponds to a single occupation, of which the sex ratio and the percentage of gender neutral pronouns are averaged over all tested languages (Malay, Estonian, Finnish, Hungarian, Armenian, Japanese, Turkish, Yoruba, Basque, Swahili and Chinese).}
+	\caption{Scatter plot of the translated sentences' statistics.
+Each point, corresponding to a single occupation, is positioned in the horizontal axis according to its sex ratio and in the vertical axis according to the frequency with which it is translated using gender neutral pronouns.
+Both quantities are averaged over all tested languages.}
 	\label{fig:scatterplot-languages}

### 2017-12-26 paper/main.tex
Fixed problem with JAIR format

@@
 \documentclass{article}
+%\usepackage{jair, theapa, rawfonts}
 
-\usepackage[utf8]{inputenc}
+%\usepackage[utf8]{inputenc}
 %\usepackage{fontspec}
@@
 
-\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
+%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
+%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
+%\name Luis Lamb \email lamb@inf.ufrgs.br \\
+%\addr Institute of Informatics \\
+%Federal University of Rio Grande do Sul \\
+%Porto Alegre, 91501-970 Brazil}
+
+\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
 \author{Marcelo Prates \and Luis Lamb}
@@
 \begin{abstract}
-In recent times there has been a growing concern both inside and outside academia about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of their own prejudices.
-A large number of AI tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas.
+Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices.
+A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas.
 Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI.
-In this paper, we construct sentences with gender neutral pronouns in different languages which support them and translate these sentences into English using the Google Translate API.
-The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate assumes for each sentence.
-We show that not only there is a tendency towards translating pronouns to \say{he} but also that this tendency is exaggerated for sentences about male dominated fields such as computer science.
-At the other hand, sentences about artistic fields are less likely to yield male defaults.
-This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of those people behind its training dataset -- ourselves.
+In this paper, we construct sentences with gender neutral pronouns in different languages 
+and translate these sentences into English using the Google Translate API. 
+The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate
+API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
+masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science.
+On the other hand, sentences about artistic fields are less likely to yield male defaults.
+This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
 \end{abstract}
@@
 
-Although the idea of automated translation can in principle be traced back to as long as the 17th century with René Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by Warren Weaver \cite{locke1955machine} discussing the possibility of employing digital computers to perform automated translation.
+Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
 The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}.
-Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation.
-As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential ALPAC report \cite{hutchins1986machine}.
-Research was almost completely abandoned in the United States, making a shy re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical}.
-Statistical and example-based machine translation have been on the rise ever since \cite{carl2003recent}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google,gtranslateneuralnet2017}) amounting to over $200$ million users daily \cite{gtranslateneuralnet2017,gtranslatezeroshot2017}.
+Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Noam Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation.
+As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \cite{hutchins1986machine}.
+Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical,Moses2007}.
+Statistical and example-based machine translation have been on the rise ever since \cite{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily.
 
 In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism.
-Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data.
-Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics:
+Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \cite{Chomsky2011,norvig2017chomsky}.
+Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \cite{Chomsky2011}:
 
 \begin{quotation}
-I mean actually you could do physics this way, instead of studying things like balls rolling down frictionless planes, which can't happen in nature, if you took a ton of video tapes of what's happening outside my office window, let's say, you know, leaves flying and various things, and you did an extensive analysis of them, you would get some kind of prediction of what's likely to happen next, certainly way better than anybody in the physics department could do.
-Well that's a notion of success which is I think novel, I don't know of anything like it in the history of science.
+\textsl{I mean actually you could do physics this way, instead of studying things like balls rolling down frictionless planes, which can't happen in nature, if you took a ton of video tapes of what's happening outside my office window, let's say, you know, leaves flying and various things, and you did an extensive analysis of them, you would get some kind of prediction of what's likely to happen next, certainly way better than anybody in the physics department could do.
+Well that's a notion of success which is I think novel, I don't know of anything like it in the history of science}.
 \end{quotation}
 
-Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained}:
+Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \cite{norvig2017chomsky}:
 
 \begin{quotation}
-As another example, consider the Newtonian model of gravitational attraction, which says that the force between two objects of mass $m_1$ and $m_2$ a distance $r$ apart is given by
-
+\textsl{As another example, consider the Newtonian model of gravitational attraction, which says that the force between two objects of mass $m_1$ and $m_2$ a distance $r$ apart is given by}
 \begin{equation*}
@@
 \end{equation*}
-
-where $G$ is the universal gravitational constant.
+\textsl{where $G$ is the universal gravitational constant.
 This is a trained model because the gravitational constant G is determined by statistical inference over the results of a series of experiments that contain stochastic experimental error.
@@
 I believe that Chomsky has no objection to this kind of statistical model.
-Rather, he seems to reserve his criticism for statistical models like Shannon's that have quadrillions of parameters, not just one or two.
+Rather, he seems to reserve his criticism for statistical models like Shannon's that have quadrillions of parameters, not just one or two.}
 \end{quotation}
@@
 Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society.
+
 On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}.
@@
 With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation.
-We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English.
+We illustrate how this can be done by simply exploiting Google Translate to map sentences from a gender neutral language into English.
 As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.
@@
 Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed.
-This poses an ethical rather than a technical limitation to statistical machine translation and big data statistical models in general.
+This poses
+mostly an ethical rather than technical limitation to statistical machine translation and big data statistical models in general.
 
-\section{Methods}
+\section{Materials and Methods}
 
-We believe that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
+We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
 Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate.
 The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
-Given the success of Google Translate, which amounts to 200 million users daily, we have decided to exploit its API to obtain the desired thermometer of gender bias.
+Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
 Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}).
-Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they 1) exhibit a pronominal gender system and 2) are supported by Google Translate.
-Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
-Following difficulties with Bengali, Nepali and Korean we have decided not to work with these languages also.
+Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
+( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
 
@@
 Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
-For the purposes of this work, we have decided to work only with languages lacking such a system, and as such Persian and Esperanto have been discarded.
-Languages lacking support from Google Translate have been discarded.
-Following difficulties with Bengali, Nepali and Korean, these languages have also been discarded.}
+For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded.
+Languages lacking support from Google Translate have been discarded.}
 	\label{tab:gender-neutral-languages}
@@
 
-Figures \ref{fig:histogram-female}, \ref{fig:histogram-female} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns.
-While the number of occurences of male pronouns tends towards a normal distribution, the figure is changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation.
+Figures \ref{fig:histogram-female}, \ref{fig:histogram-male} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns.
+While the number of occurrences of male pronouns tends towards a normal distribution, the figure changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation.
 The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.
@@
 
-\section{Discussion}
-
+\section{Conclusions}
 In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them.
@@
 By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process.
-Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis.
-Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{Guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
+Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis yet.
+Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
 We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese.
@@
 
-We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it arouses discussions about the role of AI engineers on minimizing the harmful effects of the pressing modern concern of machine bias.
+We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
 
-\clearpage
-
-%\bibliographystyle{di}
 \bibliography{main}
+\bibliographystyle{named}
 

### 2017-12-27 paper/Scientific Reports/coverletterSciRep.tex
Updated scientific reports .tex file with bibliography inside .tex file and added covertletterSciRep.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Plain Cover Letter
% LaTeX Template
%
% This template has been downloaded from:
% http://www.latextemplates.com
%
% Original author:
% Rensselaer Polytechnic Institute (http://www.rpi.edu/dept/arc/training/latex/resumes/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[10pt]{letter} % Default font size of the document, change to 10pt to fit more text
\newcommand{\degree}{\ensuremath{^\circ}}
\usepackage{color}

\usepackage{fancyhdr,lastpage}
\usepackage{graphicx}
\usepackage{newcent} % Default font is the New Century Schoolbook PostScript font 
%\usepackage{helvet} % Uncomment this (while commenting the above line) to use the Helvetica font

% Margins
\topmargin=-1.8in % Moves the top of the document 1 inch above the default
\textheight=11.5in % Total height of the text on the page before text goes on to the next page, this can be increased in a longer letter
\oddsidemargin=-10pt % Position of the left margin, can be negative or positive if you want more or less room
\textwidth=6.5in % Total width of the text, increase this if the left margin was decreased and vice-versa

%\let\raggedleft\raggedright % Pushes the date (at the top) to the left, comment this line to have the date on the right

\begin{document}

%\pagestyle{fancy}

% Cabealho
% ~~~~~~~~~

%----------------------------------------------------------------------------------------
%	ADDRESSEE SECTION
%----------------------------------------------------------------------------------------

\begin{letter}{To Scientific Reports Office\\
SpringerNature\\
London, UK
 }

\includegraphics[scale=0.48]{logo.png}\\ \ \\ \ \\

%\includegraphics[scale=0.45]{logo.png}

%----------------------------------------------------------------------------------------
%	YOUR NAME & ADDRESS SECTION
%----------------------------------------------------------------------------------------

%----------------------------------------------------------------------------------------
%	LETTER CONTENT SECTION
%----------------------------------------------------------------------------------------

\opening{Re New submission to Scientific Reports\\
\\ \ \\}

 Dear Editors of Scientific Reports:

We hereby submit the manuscript titled \emph{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}, by Marcelo Prates and Luis C. Lamb to \emph{Scientific Reports}. \\
\\
In a nutshell, considering the recent and growing concerns in both industrial research labs,
academia and mainstream media on the impact of Artificial Intelligence in society, the analysis of  machine bias is relevant to a large number of AI systems and applications.  
Although a systematic study of such biases is indeed difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of 
gender bias in AI.  
In our paper we investigate gender bias in machine translation. Specifically, we present several case studies and experiments with the Google Translate API. We collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate API computes for each sentence.

We show that not only there is a tendency towards translating pronouns to masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This contributes to investigate how our society sees stereotypical gender roles, and suggests points to the improvement of   statistical machine translation systems.

Suggested names of Referees: \\
Prof. Pascal Hizler, Wright State University, Dayton, OH -  pascal.hitzler@wright.edu \\
Prof. Moshe Vardi, Rice University, Houston, TX - vardi@cs.rice.edu\\
Prof. Artur Garcez, City, University of London - aag@soi.city.ac.uk

Data disclosure: All data used in this research will be freely available.\\
M.P. and L.C.L. created the models and analyzed the results. M.P. wrote the code
of the experiments.
M.P. and L.C.L. wrote the manuscript. All authors gave final approval for submission.

We look forward to hearing from you.\\ \ \\ \ \\ \ \\

Yours sincerely,\\

\begin{flushleft}
Marcelo Prates and Luis C. Lamb
 \end{flushleft}

\let\thefootnote\relax\footnote{Authors' Contact:\\
Institute of Informatics \\
Federal University of Rio Grande do Sul\\ 
91501-970 Porto Alegre, RS, Brazil \\
http://www.inf.ufrgs.br/  \\
\textbf{morprates@inf.ufrgs.br}\\
\textbf{lamb@inf.ufrgs.br}\\
Tel +55 (51) 3308 9464 \\
Fax +55 (51) 3308 7308}

\end{letter}

\end{document}

### 2017-12-27 paper/Scientific Reports/main.tex
Added Scientific Reports version
Fixed problem with labels
Updated scientific reports .tex file with bibliography inside .tex file and added covertletterSciRep.tex

--- 2017-12-26 paper/main.tex
@@
-\documentclass{article}
-%\usepackage{jair, theapa, rawfonts}
-
-%\usepackage[utf8]{inputenc}
-%\usepackage{fontspec}
-\usepackage{xeCJK}
-%\usepackage{armtex}
+\documentclass[fleqn,10pt]{wlscirep}
 
@@
 
-\usepackage[numbers]{natbib}
-\bibliographystyle{plainnat}
-
-%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
-%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
-%\name Luis Lamb \email lamb@inf.ufrgs.br \\
-%\addr Institute of Informatics \\
-%Federal University of Rio Grande do Sul \\
-%Porto Alegre, 91501-970 Brazil}
+%\usepackage[numbers]{natbib}
+%\bibliographystyle{plainnat}
 
 \title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
-\author{Marcelo Prates \and Luis Lamb}
 
-\begin{document}
+\author[1,*]{Marcelo Prates}
+\author[1,+]{Luis Lamb}
+\affil[1]{Federal University of Rio Grande do Sul, Institute of Informatics, Porto Alegre, 91501-970, Brazil}
 
-\maketitle
+\affil[*]{morprates@inf.ufrgs.br}
+\affil[+]{lamb@inf.ufrgs.br}
+
+%\keywords{Keyword1, Keyword2, Keyword3}
 
@@
 \end{abstract}
+\begin{document}
+
+\flushbottom
+\maketitle
+% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
+%
+%  Click the title above to edit the author information and abstract
+%
+\thispagestyle{empty}
 
@@
 	\centering
-	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
-	\label{fig:screenshot-gtranslate-hungarian}
+	\fbox{\includegraphics[width=14cm]{pictures/screenshot-gtranslate-hungarian}}
 	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation.
 This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
+	\label{fig:screenshot-gtranslate-hungarian}
 \end{figure}
-
-\section{Motivation}
-
 If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights.
@@
 \begin{table}[H]
+\begin{small}
 	\centering
@@
 	\label{tab:gender-neutral-languages}
+	\end{small}
 \end{table}
@@
 	\label{tab:occupations}
+	\end{small}
 \end{table}
@@
 \begin{table}[H]
+\small{
 	\centering
@@
 	\label{tab:templates}
+	}
 \end{table}
@@
 	\label{tab:occupations-examples}
+	}
 \end{table}
@@
 	\label{tab:adjectives}
+	}
 \end{table}
@@
 	\end{tabular}
-	\label{tab:gender-by-category}
 	\caption{Number of female, male and neutral pronominal genders per occupation category in the translated sentences.
 The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in male dominated fields such as computer science, with up to $\approx 18$ occurrences of male pronouns for each of a female one.}
+	\label{tab:gender-by-category}
+	}
 \end{table}
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/gender-by-category}
-	\label{fig:gender-by-category}
+	\includegraphics[width=10cm]{pictures/gender-by-category}
 	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively.
 Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
+	\label{fig:gender-by-category}
 \end{figure}
@@
 	\label{tab:gender-by-language}
+	}
 \end{table}
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/gender-by-language}
+	\includegraphics[width=10cm]{pictures/gender-by-language}
 	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively.
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/scatterplot-languages}
+	\includegraphics[width=10cm]{pictures/scatterplot-languages}
 	\caption{Scatter plot of the translated sentences' statistics.
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/histogram-female}
+	\includegraphics[width=10cm]{pictures/histogram-female}
 	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0.
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/histogram-male}
+	\includegraphics[width=10cm]{pictures/histogram-male}
 	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9.
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/histogram-neutral}
+	\includegraphics[width=10cm]{pictures/histogram-neutral}
 	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
@@
 	\label{tab:gender-by-adjective}
+	}
 \end{table}
@@
 	\centering
-	\includegraphics[width=\linewidth]{pictures/barplot-adjectives}
+	\includegraphics[width=10cm]{pictures/barplot-adjectives}
 	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
@@
 
-\bibliography{main}
-\bibliographystyle{named}
+\begin{thebibliography}{10}
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+\expandafter\ifx\csname doiprefix\endcsname\relax\def\doiprefix{DOI }\fi
+\providecommand{\bibinfo}[2]{#2}
+\providecommand{\eprint}[2][]{\url{#2}}
+
+\bibitem{dascal1982universal}
+\bibinfo{author}{Dascal, M.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Universal language schemes in
+  {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson}}.
+\newblock {\emph{\JournalTitle{Studia leibnitiana}}}
+  \textbf{\bibinfo{volume}{14}}, \bibinfo{pages}{98--109}
+  (\bibinfo{year}{1982}).
+
+\bibitem{locke1955machine}
+\bibinfo{author}{Locke, W.~N.} \& \bibinfo{author}{Booth, A.~D.}
+\newblock \emph{\bibinfo{title}{Machine translation of languages: fourteen
+  essays}} (\bibinfo{publisher}{Published jointly by Technology Press of the
+  Massachusetts Institute of Technology and Wiley, New York},
+  \bibinfo{year}{1955}).
+
+\bibitem{weaver1955translation}
+\bibinfo{author}{Weaver, W.}
+\newblock \bibinfo{title}{Translation}.
+\newblock In \bibinfo{editor}{Locke, W.~N.} \& \bibinfo{editor}{Booth, A.~D.}
+  (eds.) \emph{\bibinfo{booktitle}{Machine translation of languages}},
+  vol.~\bibinfo{volume}{14}, \bibinfo{pages}{15--23}
+  (\bibinfo{publisher}{Cambridge: Technology Press, MIT},
+  \bibinfo{year}{1955}).
+\newblock \urlprefix\url{http://www.mt-archive.info/Weaver-1949.pdf}.
+\newblock \bibinfo{note}{Last visited 2017-12-17}.
+
+\bibitem{gordin2015scientific}
+\bibinfo{author}{Gordin, M.~D.}
+\newblock \emph{\bibinfo{title}{Scientific Babel: How science was done before
+  and after global English}} (\bibinfo{publisher}{University of Chicago Press},
+  \bibinfo{year}{2015}).
+
+\bibitem{hutchins1986machine}
+\bibinfo{author}{Hutchins, W.~J.}
+\newblock \emph{\bibinfo{title}{Machine translation: past, present, future}}
+  (\bibinfo{publisher}{Ellis Horwood Chichester}, \bibinfo{year}{1986}).
+
+\bibitem{koehn2009statistical}
+\bibinfo{author}{Koehn, P.}
+\newblock \emph{\bibinfo{title}{Statistical machine translation}}
+  (\bibinfo{publisher}{Cambridge University Press}, \bibinfo{year}{2009}).
+
+\bibitem{Moses2007}
+\bibinfo{author}{Koehn, P.} \emph{et~al.}
+\newblock \bibinfo{title}{Moses: Open source toolkit for statistical machine
+  translation}.
+\newblock In \emph{\bibinfo{booktitle}{{ACL} 2007, Proceedings of the 45th
+  Annual Meeting of the Association for Computational Linguistics, June 23-30,
+  2007, Prague, Czech Republic}} (\bibinfo{year}{2007}).
+\newblock \urlprefix\url{http://aclweb.org/anthology/P07-2045}.
+
+\bibitem{Bahdanau2014}
+\bibinfo{author}{Bahdanau, D.}, \bibinfo{author}{Cho, K.} \&
+  \bibinfo{author}{Bengio, Y.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Neural machine translation by
+  jointly learning to align and translate}}.
+\newblock {\emph{\JournalTitle{CoRR}}} \textbf{\bibinfo{volume}{abs/1409.0473}}
+  (\bibinfo{year}{2014}).
+\newblock \urlprefix\url{http://arxiv.org/abs/1409.0473}.
+\newblock \eprint{1409.0473}.
+
+\bibitem{carl2003recent}
+\bibinfo{author}{Carl, M.} \& \bibinfo{author}{Way, A.}
+\newblock \emph{\bibinfo{title}{Recent advances in example-based machine
+  translation}}, vol.~\bibinfo{volume}{21} (\bibinfo{publisher}{Springer
+  Science \& Business Media}, \bibinfo{year}{2003}).
+
+\bibitem{Firat2017}
+\bibinfo{author}{Firat, O.}, \bibinfo{author}{Cho, K.},
+  \bibinfo{author}{Sankaran, B.}, \bibinfo{author}{Yarman{-}Vural, F.~T.} \&
+  \bibinfo{author}{Bengio, Y.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Multi-way, multilingual neural
+  machine translation}}.
+\newblock {\emph{\JournalTitle{Computer Speech {\&} Language}}}
+  \textbf{\bibinfo{volume}{45}}, \bibinfo{pages}{236--252}
+  (\bibinfo{year}{2017}).
+\newblock \urlprefix\url{https://doi.org/10.1016/j.csl.2016.10.006}.
+\newblock \doiprefix 10.1016/j.csl.2016.10.006.
+
+\bibitem{wu2016google}
+\bibinfo{author}{Johnson, M.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Google's multilingual neural
+  machine translation system: Enabling zero-shot translation}}.
+\newblock {\emph{\JournalTitle{{TACL}}}} \textbf{\bibinfo{volume}{5}},
+  \bibinfo{pages}{339--351} (\bibinfo{year}{2017}).
+\newblock
+  \urlprefix\url{https://transacl.org/ojs/index.php/tacl/article/view/1081}.
+
+\bibitem{Chomsky2011}
+\bibinfo{author}{Chomsky, N.}
+\newblock \bibinfo{title}{The golden age: A look at the original roots of
+  artificial intelligence, cognitive science, and neuroscience (partial
+  transcript of an interview with {N}. {C}homsky at {MIT150} {S}ymposia:
+  Brains, minds and machines symposium} (\bibinfo{year}{2011}).
+\newblock \urlprefix\url{https://chomsky.info/20110616/}.
+\newblock \bibinfo{note}{Last visited 2017-12-26}.
+
+\bibitem{norvig2017chomsky}
+\bibinfo{author}{Norvig, P.}
+\newblock \bibinfo{title}{On {C}homsky and the two cultures of statistical
+  learning} (\bibinfo{year}{2017}).
+\newblock \urlprefix\url{http://norvig.com/chomsky.html}.
+\newblock \bibinfo{note}{Last visited 2017-12-17}.
+
+\bibitem{schiebinger2014scientific}
+\bibinfo{author}{Schiebinger, L.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Scientific research must take
+  gender into account}}.
+\newblock {\emph{\JournalTitle{Nature}}} \textbf{\bibinfo{volume}{507}},
+  \bibinfo{pages}{9} (\bibinfo{year}{2014}).
+
+\bibitem{garcia2016racist}
+\bibinfo{author}{Garcia, M.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Racist in the machine: The
+  disturbing implications of algorithmic bias}}.
+\newblock {\emph{\JournalTitle{World Policy Journal}}}
+  \textbf{\bibinfo{volume}{33}}, \bibinfo{pages}{111--117}
+  (\bibinfo{year}{2016}).
+
+\bibitem{angwin2016machine}
+\bibinfo{author}{Angwin, J.}, \bibinfo{author}{Larson, J.},
+  \bibinfo{author}{Mattu, S.} \& \bibinfo{author}{Kirchner, L.}
+\newblock \bibinfo{title}{Machine bias: {T}here's software used across the
+  country to predict future criminals and it's biased against blacks}
+  (\bibinfo{year}{2016}).
+\newblock \urlprefix\url{https://www.
+  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}.
+\newblock \bibinfo{note}{Last visited 2017-12-17}.
+
+\bibitem{wang2017deep}
+\bibinfo{author}{Wang, Y.} \& \bibinfo{author}{Kosinski, M.}
+\newblock \bibinfo{title}{Deep neural networks are more accurate than humans at
+  detecting sexual orientation from facial images} (\bibinfo{year}{2017}).
+
+\bibitem{womanunlockphone2017}
+\bibinfo{author}{Papenfuss, M.}
+\newblock \bibinfo{title}{Woman in {C}hina says colleague's face was able to
+  unlock her i{P}hone {X}} (\bibinfo{year}{2017}).
+\newblock
+  \urlprefix\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}.
+\newblock \bibinfo{note}{Last visited 2017-12-17}.
+
+\bibitem{racistsoapdispenser2017}
+\bibinfo{author}{Mills, K.-A.}
+\newblock \bibinfo{title}{'{R}acist' soap dispenser refuses to help
+  dark-skinned man wash his hands - but {T}witter blames 'technology'}
+  (\bibinfo{year}{2017}).
+\newblock
+  \urlprefix\url{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}.
+\newblock \bibinfo{note}{Last visited 2017-12-17}.
+
+\end{thebibliography}
+
+%\bibliography{main}
+%\bibliographystyle{named}
 

### 2018-02-26 paper/JAIR/jair.tex

--- 2017-12-26 paper/main.tex
@@
-\documentclass{article}
-%\usepackage{jair, theapa, rawfonts}
+\documentclass[jair,twoside,11pt,theapa]{article}
+\usepackage[bookmarks=true]{hyperref}
+\hypersetup{%
+    bookmarks=false,                            % show bookmarks bar?
+    pdftitle={TeX and friends hackers manual},  % title
+    pdfauthor={Yiannis Lazarides},              % author
+    pdfsubject={TeX and LaTeX},                 % subject of the document
+    pdfkeywords={TeX, LaTeX, graphics, images}, % list of keywords
+    %colorlinks=true,                            % false: boxed links; true: colored links
+    %linkcolor=blue,                             % color of internal links
+    %citecolor=black,                            % color of links to bibliography
+    %filecolor=black,                            % color of file links
+    %urlcolor=purple,                            % color of external links
+    linktoc=page                                % only page is linked
+}
+\usepackage{jair, theapa, rawfonts}
 
-%\usepackage[utf8]{inputenc}
-%\usepackage{fontspec}
+\usepackage[utf8]{inputenc}
 \usepackage{xeCJK}
-%\usepackage{armtex}
-
-\usepackage{multirow}
-\usepackage{hyperref}
 \usepackage{float}
+\usepackage{graphicx}
+\usepackage{dirtytalk}
+\usepackage{amsmath}
 \usepackage{array}
-\usepackage{dirtytalk}
 \usepackage{csquotes}
-\usepackage{amsmath}
-\usepackage{graphicx}
 \usepackage{amssymb}
 \usepackage{pifont}
+\usepackage{multirow}
 
@@
 
-\usepackage[numbers]{natbib}
-\bibliographystyle{plainnat}
+\jairheading{1}{2018}{1-15}{?/2017}{?/2018}
+\ShortHeadings{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
+{Prates \& Lamb}
+\firstpageno{25}
 
-%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
-%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
-%\name Luis Lamb \email lamb@inf.ufrgs.br \\
-%\addr Institute of Informatics \\
-%Federal University of Rio Grande do Sul \\
-%Porto Alegre, 91501-970 Brazil}
+\begin{document}
 
 \title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
-\author{Marcelo Prates \and Luis Lamb}
 
-\begin{document}
+\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
+       \name Luis Lamb \email lamb@inf.ufrgs.br \\
+       \addr Institute of Informatics \\
+Federal University of Rio Grande do Sul \\
+Porto Alegre, 91501-970 Brazil}
+
+% For research notes, remove the comment character in the line below.
+% \researchnote
 
@@
 
-\bibliography{main}
-\bibliographystyle{named}
+\vskip 0.2in
+
+\bibliographystyle{theapa}
+\bibliography{jair}
 

### 2018-02-26 paper/Palgrave/main.tex

--- 2017-12-26 paper/main.tex
@@
-\documentclass{article}
-%\usepackage{jair, theapa, rawfonts}
-
-%\usepackage[utf8]{inputenc}
-%\usepackage{fontspec}
-\usepackage{xeCJK}
-%\usepackage{armtex}
+\documentclass[fleqn,10pt]{article}
 
 \usepackage{multirow}
-\usepackage{hyperref}
 \usepackage{float}
@@
 
+\usepackage{natbib}
+\bibliographystyle{agsm}
+\let\hardvardurl\url
+\usepackage{url}
+\usepackage[breaklinks]{hyperref}
+
 \newcommand{\cmark}{\ding{51}}%
@@
 
-\usepackage[numbers]{natbib}
-\bibliographystyle{plainnat}
-
-%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
-%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
-%\name Luis Lamb \email lamb@inf.ufrgs.br \\
-%\addr Institute of Informatics \\
-%Federal University of Rio Grande do Sul \\
-%Porto Alegre, 91501-970 Brazil}
-
 \title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
-\author{Marcelo Prates \and Luis Lamb}
 
@@
 
-\bibliography{main}
-\bibliographystyle{named}
+\begin{thebibliography}{xx}
 
+\harvarditem[Angwin et~al.]{Angwin, Larson, Mattu \harvardand\
+  Kirchner}{2016}{angwin2016machine}
+Angwin, J., Larson, J., Mattu, S. \harvardand\ Kirchner, L.  \harvardyearleft
+  2016\harvardyearright , `Machine bias: {T}here's software used across the
+  country to predict future criminals and it's biased against blacks'.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{https://www.
+  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}
+
+\harvarditem[Bahdanau et~al.]{Bahdanau, Cho \harvardand\
+  Bengio}{2014}{Bahdanau2014}
+Bahdanau, D., Cho, K. \harvardand\ Bengio, Y.  \harvardyearleft
+  2014\harvardyearright , `Neural machine translation by jointly learning to
+  align and translate', {\em CoRR} {\bf abs/1409.0473}.
+\newline\harvardurl{http://arxiv.org/abs/1409.0473}
+
+\harvarditem{Carl \harvardand\ Way}{2003}{carl2003recent}
+Carl, M. \harvardand\ Way, A.  \harvardyearleft 2003\harvardyearright , {\em
+  Recent advances in example-based machine translation}, Vol.~21, Springer
+  Science \& Business Media.
+
+\harvarditem{Chomsky}{2011}{Chomsky2011}
+Chomsky, N.  \harvardyearleft 2011\harvardyearright , `The golden age: A look
+  at the original roots of artificial intelligence, cognitive science, and
+  neuroscience (partial transcript of an interview with {N}. {C}homsky at
+  {MIT150} {S}ymposia: Brains, minds and machines symposium'.
+\newblock Last visited 2017-12-26.
+\newline\harvardurl{https://chomsky.info/20110616/}
+
+\harvarditem{Dascal}{1982}{dascal1982universal}
+Dascal, M.  \harvardyearleft 1982\harvardyearright , `Universal language
+  schemes in {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson',
+  {\em Studia leibnitiana} {\bf 14}(1),~98--109.
+
+\harvarditem[Firat et~al.]{Firat, Cho, Sankaran, Yarman{-}Vural \harvardand\
+  Bengio}{2017}{Firat2017}
+Firat, O., Cho, K., Sankaran, B., Yarman{-}Vural, F.~T. \harvardand\ Bengio, Y.
+   \harvardyearleft 2017\harvardyearright , `Multi-way, multilingual neural
+  machine translation', {\em Computer Speech {\&} Language} {\bf 45},~236--252.
+\newline\harvardurl{https://doi.org/10.1016/j.csl.2016.10.006}
+
+\harvarditem{Garcia}{2016}{garcia2016racist}
+Garcia, M.  \harvardyearleft 2016\harvardyearright , `Racist in the machine:
+  The disturbing implications of algorithmic bias', {\em World Policy Journal}
+  {\bf 33}(4),~111--117.
+
+\harvarditem{Gordin}{2015}{gordin2015scientific}
+Gordin, M.~D.  \harvardyearleft 2015\harvardyearright , {\em Scientific Babel:
+  How science was done before and after global English}, University of Chicago
+  Press.
+
+\harvarditem{Hutchins}{1986}{hutchins1986machine}
+Hutchins, W.~J.  \harvardyearleft 1986\harvardyearright , {\em Machine
+  translation: past, present, future}, Ellis Horwood Chichester.
+
+\harvarditem[Johnson et~al.]{Johnson, Schuster, Le, Krikun, Wu, Chen, Thorat,
+  Vi{\'{e}}gas, Wattenberg, Corrado, Hughes \harvardand\
+  Dean}{2017}{wu2016google}
+Johnson, M., Schuster, M., Le, Q.~V., Krikun, M., Wu, Y., Chen, Z., Thorat, N.,
+  Vi{\'{e}}gas, F.~B., Wattenberg, M., Corrado, G., Hughes, M. \harvardand\
+  Dean, J.  \harvardyearleft 2017\harvardyearright , `Google's multilingual
+  neural machine translation system: Enabling zero-shot translation', {\em
+  {TACL}} {\bf 5},~339--351.
+\newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}
+
+\harvarditem{Koehn}{2009}{koehn2009statistical}
+Koehn, P.  \harvardyearleft 2009\harvardyearright , {\em Statistical machine
+  translation}, Cambridge University Press.
+
+\harvarditem[Koehn et~al.]{Koehn, Hoang, Birch, Callison{-}Burch, Federico,
+  Bertoldi, Cowan, Shen, Moran, Zens, Dyer, Bojar, Constantin \harvardand\
+  Herbst}{2007}{Moses2007}
+Koehn, P., Hoang, H., Birch, A., Callison{-}Burch, C., Federico, M., Bertoldi,
+  N., Cowan, B., Shen, W., Moran, C., Zens, R., Dyer, C., Bojar, O.,
+  Constantin, A. \harvardand\ Herbst, E.  \harvardyearleft
+  2007\harvardyearright , Moses: Open source toolkit for statistical machine
+  translation, {\em in} `{ACL} 2007, Proceedings of the 45th Annual Meeting of
+  the Association for Computational Linguistics, June 23-30, 2007, Prague,
+  Czech Republic'.
+\newline\harvardurl{http://aclweb.org/anthology/P07-2045}
+
+\harvarditem{Locke \harvardand\ Booth}{1955}{locke1955machine}
+Locke, W.~N. \harvardand\ Booth, A.~D.  \harvardyearleft 1955\harvardyearright
+  , {\em Machine translation of languages: fourteen essays}, Published jointly
+  by Technology Press of the Massachusetts Institute of Technology and Wiley,
+  New York.
+
+\harvarditem{Mills}{2017}{racistsoapdispenser2017}
+Mills, K.-A.  \harvardyearleft 2017\harvardyearright , `'{R}acist' soap
+  dispenser refuses to help dark-skinned man wash his hands - but {T}witter
+  blames 'technology''.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}
+
+\harvarditem{Norvig}{2017}{norvig2017chomsky}
+Norvig, P.  \harvardyearleft 2017\harvardyearright , `On {C}homsky and the two
+  cultures of statistical learning'.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://norvig.com/chomsky.html}
+
+\harvarditem{Papenfuss}{2017}{womanunlockphone2017}
+Papenfuss, M.  \harvardyearleft 2017\harvardyearright , `Woman in {C}hina says
+  colleague's face was able to unlock her i{P}hone {X}'.
+\newblock Last visited 2017-12-17.
+\newline\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}
+
+\harvarditem{Schiebinger}{2014}{schiebinger2014scientific}
+Schiebinger, L.  \harvardyearleft 2014\harvardyearright , `Scientific research
+  must take gender into account', {\em Nature} {\bf 507}(7490),~9.
+
+\harvarditem{Wang \harvardand\ Kosinski}{2017}{wang2017deep}
+Wang, Y. \harvardand\ Kosinski, M.  \harvardyearleft 2017\harvardyearright ,
+  `Deep neural networks are more accurate than humans at detecting sexual
+  orientation from facial images'.
+
+\harvarditem{Weaver}{1955}{weaver1955translation}
+Weaver, W.  \harvardyearleft 1955\harvardyearright , Translation, {\em in}
+  W.~N. Locke \harvardand\ A.~D. Booth, eds, `Machine translation of
+  languages', Vol.~14, Cambridge: Technology Press, MIT, pp.~15--23.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://www.mt-archive.info/Weaver-1949.pdf}
+
+\end{thebibliography}
 \end{document}

### 2018-03-12 paper/main.tex
Added 'revisions-subtractive.txt', updated with the remaining modifications needed to complete the paper. Applied some modifications to the text, rewriting the motivation section

@@
 On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}.
-Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears.
+Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view.
 Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}.
@@
 We illustrate how this can be done by simply exploiting Google Translate to map sentences from a gender neutral language into English.
-As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.
+As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that sentences about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.
 
@@
 
-If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights.
-One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others.
-We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market.
-We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.
-
-If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models.
-Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed.
-This poses
-mostly an ethical rather than technical limitation to statistical machine translation and big data statistical models in general.
+As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}.
+Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}.
+Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}.
+This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}.
+Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure.
+The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+This paper intends to investigate whether it does.
+We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
 
@@
 The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
-Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
+Given the success of Google Translate, which amounts to 200 million users daily \cite{Gtranslate200daily2017}, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
 Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}).
@@
 By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process.
-Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis yet.
+Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet.
 Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
@@
 
-We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
+We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
 

### 2018-03-12 paper/main.tex
Cited two papers as requested by reviewer 1

@@
 \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
+In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.
 
@@
 Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure.
-The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+The researchers are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
 This paper intends to investigate whether it does.

### 2018-03-13 Paper/main.tex
AA

--- 2018-02-26 paper/Palgrave/main.tex
@@
 \end{figure}
-If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights.
-One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others.
-We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market.
-We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.
 
-If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models.
-Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed.
-This poses
-mostly an ethical rather than technical limitation to statistical machine translation and big data statistical models in general.
+\section{Motivation}
+
+As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}.
+Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}.
+Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}.
+This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}.
+Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure.
+The researchers are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+This paper intends to investigate whether it does.
+We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
 
@@
 	\centering
+	\begin{tabular}{|m{3.5cm}|c|c|c|}
+	\hline	
+	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
+	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
+	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
+	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
+	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
+	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
+	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
+	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
+	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
+	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
+	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
+	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
+	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
+	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
+	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
+	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
+	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
+	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
+	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
+	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
+	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
+	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
+	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
+	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
+	\end{tabular}
+	\caption{Selected occupations obtained from the U.S.
+Bureau of Labor Statistics \url{https://???}, grouped by category.
+We obtained a total of $1019$ occupations from $22$ distinct categories.
+We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
+	\label{tab:occupations}
+	\end{small}
+\end{table}
+
+\begin{table}[H]
+\begin{small}
+	\centering
 	\begin{tabular}{|c|c|}
@@
 	\hline
-	stagehands & author & neurologist \\ \hline
-	screenwriter & animator & marketing director \\ \hline
-	biochemist & endocrinologist & freelancer \\ \hline
-	neurosurgeon & computer scientist & petrochemical engineer \\ \hline
-	food stylist & cardiothoracic surgeon & property master \\ \hline
-	literary editor & video editor & animation director \\ \hline
-	house manager & chief administrative officer & arts administration \\ \hline
-	actor & dialysis technician & family nurse practitioner \\ \hline
-	psychologist & chief creative officer & flash developer \\ \hline
-	scenic artist & producer & medical laboratory scientist \\ \hline
+	Insurance sales agent & Editor & Rancher \\ \hline
+	Ticket taker & Pile-driver operator & Tool maker \\ \hline
+	Jeweler & Judicial law clerk & Auditing clerk \\ \hline
+	Physician & Embalmer & Door-to-door salesperson \\ \hline
+	Packer & Bookkeeping clerk & Community health worker \\ \hline
+	Sales worker & Floor finisher & Social science technician \\ \hline
+	Probation officer & Paper goods machine setter & Heating installer \\ \hline
+	Animal breeder & Instructor & Teacher assistant \\ \hline
+	Statistical assistant & Shipping clerk & Trapper \\ \hline
+	Pharmacy aide & Sewing machine operator & Service unit operator \\ \hline
 	\end{tabular}
-	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 462 different occupations.}
+	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
 	\label{tab:occupations-examples}
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|c|c|}
+	\begin{tabular}{|c|c|c|c|}
 	\hline
-	Category & Female & Male & Neutral & Ratio & Total \\ \hline
+	Category 								& Female ($\%$)	& Male ($\%$)	& Neutral ($\%$) \\ \hline
 	\hline		
-	Artistic        & 188 & 518 & 72 & 2.755 	& 918 \\ \hline
-	Computer        & 7   & 125 & 12 & 17.857 	& 171 \\ \hline
-	Corporate       & 36  & 340 & 23 & 9.444 	& 450 \\ \hline
-	Dance           & 31  & 33  & 8  & 1.064 	& 81  \\ \hline
-	Film-television & 54  & 125 & 18 & 2.315 	& 234 \\ \hline
-	Healthcare      & 176 & 483 & 64 & 2.744 	& 792 \\ \hline
-	Industrial      & 40  & 135 & 29 & 3.375   	& 234 \\ \hline
-	Science         & 34  & 357 & 25 & 10.500	& 450 \\ \hline
-	Service         & 5   & 63  & 10 & 12.600	& 90  \\ \hline
-	Theatre         & 75  & 296 & 38 & 3.947 	& 477 \\ \hline
-	Writing         & 41  & 148 & 30 & 3.610 	& 261 \\ \hline
+	Education / Training / Library 			& $25.207$ 		& $59.917$		& $2.893$		\\ \hline
+	Business / Financial 					& $6.324$ 		& $67.194$		& $7.312$		\\ \hline
+	Office / Administrative					& $10.031$		& $62.382$		& $7.732$		\\ \hline
+	Healthcare support						& $29.545$		& $51.705$		& $7.955$		\\ \hline
+	Management								& $9.486$		& $68.182$		& $3.953$		\\ \hline
+	Installation / Maintenance / Repair		& $3.596$		& $68.332$		& $11.588$		\\ \hline
+	Healthcare practitioners / technical	& $21.564$		& $61.734$		& $6.977$		\\ \hline
+	Community and social service			& $15.584$ 		& $63.636$		& $4.545$		\\ \hline
+	Sales and related						& $8.117$ 		& $69.156$		& $8.766$		\\ \hline
+	Production								& $14.36$ 		& $53.306$		& $13.12$		\\ \hline
+	Architecture and engineering			& $2.194$		& $79.624$		& $5.016$		\\ \hline
+	Life, physical, and social sciences		& $3.743$ 		& $82.888$		& $3.743$		\\ \hline
+	Transportation / material moving		& $7.532$ 		& $66.234$		& $11.688$		\\ \hline
+	Arts / Entertainment					& $7.862$ 		& $71.007$		& $4.423$		\\ \hline
+	Legal									& $11.688$ 		& $67.532$		& $5.195$		\\ \hline
+	Protective service						& $6.818$ 		& $75.0$		& $5.844$		\\ \hline
+	Food prepaparation / serving			& $16.578$ 		& $59.893$		& $10.16$		\\ \hline
+	Farming, fishing, and forestry			& $10.49$ 		& $62.238$		& $8.392$		\\ \hline
+	Computer and mathematical				& $2.841$ 		& $76.136$		& $4.545$		\\ \hline
+	Personal care and service				& $20.937$ 		& $55.096$		& $9.091$		\\ \hline
+	Construction and extraction				& $9.492$ 		& $61.631$		& $11.096$		\\ \hline
+	Cleaning / Maintenance					& $11.818$		& $63.636$		& $8.182$		\\ \hline
 	\hline
-	Total           & 687 & 2623 & 329 & 3.818 & 4158 \\ \hline
+	Total           & ? & ? & ? \\ \hline
 	\end{tabular}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/gender-by-category}
+	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
 	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively.
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|c|c|}
+	\begin{tabular}{|c|c|c|c|}
 	\hline
-	Language & Female & Male & Neutral & Ratio & Total \\ \hline
+	Language 	& Female ($\%$) 	& Male ($\%$)		& Neutral ($\%$)	\\ \hline
 	\hline
-	Malay     & 47  & 415  & 0   & 8.830 	& 462 \\ \hline
-	Estonian  & 130 & 332  & 0   & 2.554 	& 462 \\ \hline
-	Finnish   & 179 & 283  & 0   & 1.581 	& 462 \\ \hline
-	Hungarian & 189 & 270  & 1   & 1.429 	& 462 \\ \hline
-	Armenian  & 101 & 360  & 1   & 3.564 	& 462 \\ \hline
-	Japanese  & 2   & 215  & 0   & 107.500  & 462 \\ \hline
-	Turkish   & 22  & 394  & 43  & 17.909	& 462 \\ \hline
-	Yoruba    & 12  & 304  & 131 & 25.333	& 462 \\ \hline
-	Basque    & 5   & 50   & 153 & 10.000   & 462 \\ \hline
-	Swahili   & 76  & 386  & 0   & 5.079 	& 462 \\ \hline
-	Chinese   & 14  & 385  & 23  & 27.500   & 462 \\ \hline
-	\hline
-	Total     & 777 & 3394 & 352 & 4.368	& 4158   \\ \hline
+	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
+	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
+	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
+	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
+	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
+	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
+	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
+	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
+	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
+	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
+	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline \hline
+	Total     	& ? 				& ? 				& ? 				\\ \hline
 	\end{tabular}
@@
 	\label{fig:gender-by-language}
-\end{figure}
-
-Instead of grouping sentences either by category or language, we can also visualize each of them individually on a scatter plot.
-Figure \ref{fig:scatterplot-languages} shows each occupation as a point on a bi-dimensional lattice, arranged horizontally by their sex ratio and vertically by the proportion of gender neutral pronouns, both averaged over translations from all tested languages.
-Each point is also color coded according to that occupation's respective category (Artistic, Writing, Science, etc.).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/scatterplot-languages}
-	\caption{Scatter plot of the translated sentences' statistics.
-Each point, corresponding to a single occupation, is positioned in the horizontal axis according to its sex ratio and in the vertical axis according to the frequency with which it is translated using gender neutral pronouns.
-Both quantities are averaged over all tested languages.}
-	\label{fig:scatterplot-languages}
 \end{figure}
@@
 	\label{fig:histogram-neutral}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
+	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0.
+The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
+	\label{fig:histogram-female-grouped}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
+	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9.
+In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
+	\label{fig:histogram-male-grouped}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
+	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
+	\label{fig:histogram-neutral-grouped}
 \end{figure}
@@
 
-We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
+We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
 
+%\bibliography{main}
+%\bibliographystyle{agsm}
+
+%\ifx
 \begin{thebibliography}{xx}
@@
 \newline\harvardurl{http://arxiv.org/abs/1409.0473}
+
+\harvarditem[Bolukbasi et~al.]{Bolukbasi, Chang, Zou, Saligrama \harvardand\
+  Kalai}{2016}{bolukbasi2016man}
+Bolukbasi, T., Chang, K.-W., Zou, J.~Y., Saligrama, V. \harvardand\ Kalai,
+  A.~T.  \harvardyearleft 2016\harvardyearright , Man is to computer programmer
+  as woman is to homemaker? debiasing word embeddings, {\em in} `Advances in
+  Neural Information Processing Systems', pp.~4349--4357.
 
@@
 
+\harvarditem[Hajian et~al.]{Hajian, Bonchi \harvardand\
+  Castillo}{2016}{hajian2016algorithmic}
+Hajian, S., Bonchi, F. \harvardand\ Castillo, C.  \harvardyearleft
+  2016\harvardyearright , Algorithmic bias: From discrimination discovery to
+  fairness-aware data mining, {\em in} `Proceedings of the 22nd ACM SIGKDD
+  international conference on knowledge discovery and data mining', ACM,
+  pp.~2125--2126.
+
 \harvarditem{Hutchins}{1986}{hutchins1986machine}
@@
 \newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}
+
+\harvarditem{Kelman}{2014}{TranslateCommunity}
+Kelman, S.  \harvardyearleft 2014\harvardyearright , `Translate community: Help
+  us improve google translate!'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://search.googleblog.com/2014/07/translate-community-help-us-improve.html}
+
+\harvarditem{Kirkpatrick}{2016}{kirkpatrick2016battling}
+Kirkpatrick, K.  \harvardyearleft 2016\harvardyearright , `Battling algorithmic
+  bias: how do we ensure algorithms treat us fairly?', {\em Communications of
+  the ACM} {\bf 59}(10),~16--17.
 
@@
 
+\harvarditem{Olson}{2018}{AlgorithmGtranslateSexist2018}
+Olson, P.  \harvardyearleft 2018\harvardyearright , `The algorithm that helped
+  google translate become sexist'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://www.forbes.com/sites/parmyolson/2018/02/15/the-algorithm-that-helped-google-translate-become-sexist/\#1c1122c27daa}
+
 \harvarditem{Papenfuss}{2017}{womanunlockphone2017}
@@
   must take gender into account', {\em Nature} {\bf 507}(7490),~9.
+
+\harvarditem{Shankland}{2017}{Gtranslate200daily2017}
+Shankland, S.  \harvardyearleft 2017\harvardyearright , `Google translate now
+  serves 200 million people daily'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://www.cnet.com/news/google-translate-now-serves-200-million-people-daily/}
 
@@
 \end{thebibliography}
+%\fi
+
 \end{document}

### 2018-03-13 gender-bias.py

@@ -1,10 +1,84 @@
 
 from googletrans import Translator
+from googletrans import LANGUAGES
 from xpinyin import Pinyin
+import csv
+import numpy as np
 
 translator = Translator()
 p = Pinyin()
 
+# Get language list
+languages = []
+with open('languages.csv', 'r') as f:
+    f.readline()
+    for line in f:
+    	language,family,pronomial_gender_system,supported = line.split(";")
+    	if(pronomial_gender_system != 'Yes' and supported.rstrip() == 'Yes'):
+    		languages.append(language)
+#end with
+
+# Read job occupations list into table
+table = np.array(list(csv.reader(open('jobs/bureau_of_labor_statistics_profession_list_gender_filtered_expanded.tsv','r'), delimiter='\t')))
+
+# Compute list of categories
+categories = list(set(table[1:,-3]))
+
+# Get dictionary of category -> jobs (jobs per category)
+categories_dict = dict([ (category, table[table[:,-3] == category,0]) for category in categories ])
+
+"""
+	Create a table with the translated version (provided by Google Translate) of each job,
+	in the following structure:
+		1. Each line corresponds to a single occupation
+		2. The first column is the occupation category ("Computer and mathematical occupations", "Healthcare practitioners and technical occupations", etc.)
+		3. The following N columns give a translated version of that occupation for each language (of N) in our list
+"""
+if False:
+	with open("Results/jobs-translations.tsv","w") as output:
+	
+		# Write header
+		# First column is category
+		output.write("Category")
+		# Then follows one column per language
+		output.write("\tEnglish")
+		for language in languages:
+			output.write('\t' + language)
+		#end for
+		output.write('\n')
+	
+		# Not iterate over all categories
+		for category in categories:
+			print("Translating occupations from category {} ...".format(category))
+			# Get all jobs for this category
+			for job in categories_dict[category]:
+				print("\tTranslating occupation \"{}\" ...".format(job))
+				output.write(category)
+				output.write('\t' + job)
+				# For each language L in our list, translate 'job' from English to L
+				for language in languages:
+					try:
+						if language == 'Chinese':
+							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest='zh-cn').text).lower()
+						else:
+							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest=language).text).lower()
+						#end if
[…]

### 2018-03-13 sample-jobs.jl

@@ -1,26 +1,0 @@
-
-using DataFrames, CSV, StatsBase
-
-dat = CSV.read("jobs.csv",delim=';',nullable=false)
-
-for category in unique(dat[:Category])
-    println("$category $( length(dat[ dat[:,:Category] .== category, :Occupation ]) )")
-end
-
-#=
-n,m = 10,3
-occupations = sample(dat[:Occupation],n*m,replace=false)
-
-output = open("test.txt","w")
-for i=1:n
-    for j=1:m
-        if j > 1
-            print(output,"& $(occupations[m*(i-1)+j]) ")
-        else
-            print(output,"$(occupations[m*(i-1)+j]) ")
-        end
-    end
-    println(output,"\\\\ \\hline")
-end
-close(output)
-=#

### 2018-03-13 script.py

@@ -1,49 +1,0 @@
-
-from googletrans import Translator
-from googletrans import LANGUAGES
-translator = Translator()
-
-# Get language list
-languages = []
-with open('languages.csv', 'r') as f:
-    f.readline()
-    for line in f:
-    	language,family,pronomial_gender_system,supported = line.split(";")
-    	if(pronomial_gender_system != 'Yes' and supported.rstrip() == 'Yes'):
-    		languages.append(language)
-
-# Initialize job categories list
-job_categories = ["artistic","computer","corporate","dance","film-television","healthcare","industrial","science","service","theatre","writing"]
-
-with open('jobs-2.csv', 'w') as output:
-
-	output.write('category')
-	output.write(';english')
-	for language in languages:
-		output.write(';' + language)
-	output.write('\n')
-
-	for category in job_categories:
-		print('Category: %s' % category)
-		with open('jobs/%s-jobs-english.txt' % category,'r') as jobs:
-			i = 0
-			for job in jobs:
-				print('Occupation: %s' % job.rstrip())
-				translations = []
-				for language in languages:
-					try:
-						if language == 'Chinese':
-							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest='zh-cn').text).lower()
-							translations.append(translated_job)
-						else:
-							translated_job = (translator.translate(job.rstrip().lower(),src='en',dest=language).text).lower()
-							translations.append(translated_job)
-					except Exception:
-						print("Could not translate occupation %s to language %s" % (job.rstrip(),language))
-						translations.append("?")
-
-				output.write(category)
-				output.write(';'+job.rstrip().lower())
-				for translation in translations:
-					output.write(';'+translation)
-				output.write('\n')

### 2018-03-13 stats.jl

@@ -2,7 +2,87 @@
 using DataFrames, CSV, Gadfly
 
 # Read job-genders.csv into a Julia DataFrame
-dat_jobs = CSV.read("job-genders.csv",delim=';',nullable=false)
+dat_jobs = CSV.read("Results/job-genders.tsv",delim='\t',nullable=false)
+
+# Remove "occupations" from the end of all category names
+dat_jobs[:Category] = map(x -> replace(x," occupations",""),dat_jobs[:Category])
+
+function rename_category(label)
+    if label=="Education, training, and library"
+        return "Education / Training / Library"
+    elseif label=="Business and financial operations"
+        return "Business / Financial"
+    elseif label=="Office and administrative support"
+        return "Office / Administrative"
+    elseif label=="Installation, maintenance, and repair"
+        return "Installation / Maintenance / Repair"
+    elseif label=="Healthcare practitioners and technical"
+        return "Healthcare practitioners / technical"
+    elseif label=="Life, physical, and social science"
+        return "Life, physical, and social sciences"
+    elseif label=="Arts, design, entertainment, sports, and media"
+        return "Arts / Entertainment"
+    elseif label=="Building and grounds cleaning and maintenance"
+        return "Cleaning / Maintenance"
+    elseif label=="Food preparation and serving related"
+        return "Food prepaparation / serving"
+    elseif label=="Transportation and material moving"
+        return "Transportation / material moving"
+    else
+        return label
+    end
+end
+
+function group_categories(label)
+    if label == "Education, training, and library"
+        return "Education"
+    elseif label == "Business and financial operations"
+        return "Corporate"
+    elseif label == "Office and administrative support"
+        return "Service"
+    elseif label == "Healthcare support"
+        return "Healthcare"
+    elseif label == "Management"
+        return "Corporate"
+    elseif label == "Installation, maintenance, and repair"
+        return "Service"
+    elseif label == "Healthcare practitioners and technical"
+        return "Healthcare"
+    elseif label == "Community and social service"
+        return "Service"
+    elseif label == "Sales and related"
+        return "Corporate"
+    elseif label == "Production"
[…]

### 2018-03-14 Paper/main.tex
Major changes to the text and structure of the paper. Added TODO.txt

@@
 \usepackage{natbib}
-\bibliographystyle{agsm}
+%\bibliographystyle{agsm}
+\bibliographystyle{named}
 \let\hardvardurl\url
@@
 Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}.
-\emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
-In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.
+\emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal assymetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
+In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.
 
@@
 	\centering
-	\fbox{\includegraphics[width=14cm]{pictures/screenshot-gtranslate-hungarian}}
+	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
 	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation.
@@
 This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}.
-Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding procedure.
-The researchers are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique.
+They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
 This paper intends to investigate whether it does.
-We are optimistic that our research endeavors, if conclusive, can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
+We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
 
@@
 
+% The hungarian punctuation in 'ő' is not rendered in the PDF. This needs to be fixed somehow. The same problem extends to Table 4 (label tab:templates)
 We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
 Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate.
+As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when \say{nurse} is replaced by \say{engineer}.
 The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
 Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
-Also, in order to solidify our results, we have decided to work with as many gender neutral languages as possible, obtaining a list of these from the Wikipedia article on the issue (\url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}).
-Table \ref{tab:gender-neutral-languages} compiles all languages from said article, with additional columns informing whether they
-( 1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
+Also, in order to solidify our results, we have decided to work
+% PEDRO: Removed "with as many gender neutral languages as possible"
+ with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures \cite{wals} and other sources.
+Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
+(1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
 
@@
 To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) are shown in Table \ref{tab:occupations}.
+This time we resorted to using the Bureau of Labor Statistics' detailed occupations table \cite{BLS2017}, from the United States Department of Labor, for providing such a list.
+The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
+Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers.
+We then separated the curated jobs into categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}.
 Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
@@
 
+% Table containing a list of gender neutral languages with information about pronominal gender systems and GTranslate support
 \begin{table}[H]
@@
 	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
-									& Ido 					& O 				& \xmark 		& \xmark 		\\
-									\cline{2-2}
-									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
-									\cline{2-2}
+									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
+									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
 									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
@@
 
+% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
 \begin{table}[H]
@@
 
-\begin{table}[H]
-\begin{small}
-	\centering
-	\begin{tabular}{|c|c|}
-	\hline
-	Category 		& \# Occupations 	\\ \hline \hline
-	Artistic 		& $102$ 			\\ \hline
-	Computer 		& $19$ 				\\ \hline
-	Corporate 		& $50$ 				\\ \hline
-	Dance 			& $9$ 				\\ \hline
-	Film/Television & $26$ 				\\ \hline
-	Healthcare 		& $88$ 				\\ \hline
-	Industrial 		& $26$ 				\\ \hline
-	Science 		& $50$ 				\\ \hline
-	Service 		& $10$ 				\\ \hline
-	Theatre 		& $52$ 				\\ \hline
-	Writing 		& $29$ 				\\ \hline
-	\hline
-	Total			& $462$				\\ \hline
-	\end{tabular}
-	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category.
-We have selected a total of 462 occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
-	\label{tab:occupations}
-	\end{small}
-\end{table}
-
+% List of sentence templates for assessing gender bias in the translation from source language to English
 \begin{table}[H]
@@
 	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
-	Japanese	& $\langle occupation \rangle$ です\\ \hline
+	Japanese	& $\langle occupation \rangle$ です			\\ \hline
 	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
@@
 
+% Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
 \begin{table}[H]
@@
 
+% Table with all adjectives we experimented with
 \begin{table}[H]
@@
 	\hline
-	happy 		& sad 		& right 	\\ \hline
-	wrong 		& afraid	& brave 	\\ \hline
-	smart		& dumb		& proud 	\\ \hline
-	ashamed		& kind		& cruel 	\\ \hline
-	envious		& loving	& hateful 	\\ \hline
-	modest 		& arrogant	& guilty	\\ \hline
-	innocent	& helpless	& shy		\\ \hline
+	Happy 		& Sad 		& Right 	\\ \hline
+	Wrong 		& Afraid	& Brave 	\\ \hline
+	Smart		& Dumb		& Proud 	\\ \hline
+	Ashamed		& Kind		& Cruel 	\\ \hline
+	Envious		& Loving	& Hateful 	\\ \hline
+	Modest 		& Arrogant	& Guilty	\\ \hline
+	Innocent	& Helpless	& Shy		\\ \hline
 	\end{tabular}
@@
 
-\section{Results}
+\section{Distribution of translated gender pronouns per occupation category}
 
-For each one of the tested 462 occupations (see Tables \ref{tab:occupations}, \ref{tab:occupations-examples}), we used the Python Google Translate API (\url{http://py-googletrans.readthedocs.io/en/latest/}) to translate sentences built with the templates in Table \ref{tab:templates} from each one of the tested languages in Table \ref{tab:gender-neutral-languages} to English.
-The resulting sentences are then classified as \emph{female}, \emph{male} or \emph{neutral} according to their respective pronouns.
-Sentences starting with \say{She/She's/Her} are classified as female, sentences starting with \say{He/He's/His} are classified as male and sentences starting with \say{It/It's/Its/They/They're/Their} are classified as (gender) neutral.
-The results from this analysis, which can be found in \url{https://github.com/marceloprates/Gender-Bias}, are further discussed below.
+A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
+What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability
+% Add statistical test here
+than it does either female or gender-neutral pronouns. Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes,
+% Add reference here
+such as life and physical sciences, architecture, engineering, computer science and mathematics.
+Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
+For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).
 
-One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
-Sentences about occupations from the \emph{Corporate} and \emph{Science} category are also disproportionately translated with male pronouns (sex ratios $9.444$ and $10.5$ respectively), while those containing occupations from the \emph{Dance} category achieve a sex ratio of almost one ($1.064$).
-Not one category has achieved a balanced sex ratio, neither does any category exhibit more gender neutral than male pronouns.
-In total, female pronouns add up to $16.522\%$ among all categories, while male pronouns add up to $63.083\%$ and gender neutral pronouns to just $7.912\%$, yielding an average sex ratio of $3.818$.
-
+% Table with our gender bias results grouped by category
 \begin{table}[H]
@@
 	Sales and related						& $8.117$ 		& $69.156$		& $8.766$		\\ \hline
-	Production								& $14.36$ 		& $53.306$		& $13.12$		\\ \hline
+	Production								& $14.36$ 		& $53.306$		& $13.120$		\\ \hline
 	Architecture and engineering			& $2.194$		& $79.624$		& $5.016$		\\ \hline
@@
 	Legal									& $11.688$ 		& $67.532$		& $5.195$		\\ \hline
-	Protective service						& $6.818$ 		& $75.0$		& $5.844$		\\ \hline
-	Food prepaparation / serving			& $16.578$ 		& $59.893$		& $10.16$		\\ \hline
-	Farming, fishing, and forestry			& $10.49$ 		& $62.238$		& $8.392$		\\ \hline
+	Protective service						& $6.818$ 		& $75.000$		& $5.844$		\\ \hline
+	Food prepaparation / serving			& $16.578$ 		& $59.893$		& $10.160$		\\ \hline
+	Farming, fishing, and forestry			& $10.490$ 		& $62.238$		& $8.392$		\\ \hline
 	Computer and mathematical				& $2.841$ 		& $76.136$		& $4.545$		\\ \hline
@@
 	\hline
-	Total           & ? & ? & ? \\ \hline
+	Total           						& $11.107$ 		& $63.030$ 		& $9.278$ 		\\ \hline
 	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders per occupation category in the translated sentences.
-The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in male dominated fields such as computer science, with up to $\approx 18$ occurrences of male pronouns for each of a female one.}
+	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
 	\label{tab:gender-by-category}
@@
 
+% Table with our gender bias results grouped by category (but categories are further grouped into "super-categories")
+\begin{table}[H]
+\small{
+	\centering
+	\begin{tabular}{|c|c|c|c|}
+	\hline
+	Category 								& Female ($\%$)			& Male ($\%$)			& Neutral ($\%$)	\\ \hline
+	\hline
+	Education								& $25.207$				& $59.917$				& $2.893$			\\ \hline
+	Corporate								& $7.955$				& $68.030$				& $6.364$			\\ \hline
+	Service									& $9.221$				& $64.987$				& $9.506$			\\ \hline
+	Healthcare								& $23.729$				& $59.014$				& $7.242$			\\ \hline
+	Production								& $14.360$				& $53.306$				& $13.120$			\\ \hline
+	STEM									& $2.992$				& $80.322$				& $4.373$			\\ \hline
+	Arts / Entertainment					& $7.862$				& $71.007$				& $4.423$			\\ \hline
+	Legal									& $11.688$				& $67.532$				& $5.195$			\\ \hline
+	Farming / Fishing / Forestry			& $10.490$				& $62.238$				& $8.392$			\\ \hline
+	Construction / Extraction				& $9.492$				& $61.631$				& $11.096$			\\ \hline
+	\hline
+	Total           						& $11.107$ 				& $63.030$ 				& $9.278$ 			\\ \hline
+	\end{tabular}
+	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\label{tab:gender-by-category-grouped}
+	}
+\end{table}
+
+Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed.
+The histogram in Figure \ref{fig:histogram-female} (and its coalesced variant in Figure \ref{fig:histogram-female-grouped}) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures \ref{fig:histogram-neutral} and \ref{fig:histogram-neutral-grouped} --, while the same data for male pronouns (shown in Figures \ref{fig:histogram-male}, \ref{fig:histogram-male-grouped}) suggests a skew normal distribution.
+Furthermore we can see both on Figures \ref{fig:histogram-female-grouped} and \ref{fig:histogram-male-grouped} how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.
+
+% PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
+These values contrast with LBS' report of gender participation.
+Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
+The variance reported for the translation results is also lower, at $0.0145145752$ in contrast with the report's $0.0666766998$.
+Altogether, the Pearson correlation between the two values is $0.3035612739$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$).
+This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.
+
+% Histogram with the distribution of female pronouns for each occupation category
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-female}
+	\caption{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
+STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table \ref{tab:gender-by-category}) extend to higher values.}
+	\label{fig:histogram-female}
+\end{figure}
+
+% Histogram with the distirbution of female pronouns for each occupation category (but categories are grouped into 'super-categories')
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
+	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure \ref{fig:histogram-female}.}
+	\label{fig:histogram-female-grouped}
+\end{figure}
+
+% Histogram with the distirbution of male prononuns for each occupation category
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-male}
+	\caption{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$. One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}
+	\label{fig:histogram-male}
+\end{figure}
+
+% Histogram with the distirbution of male pronouns for each occupation category (but categories are grouped into 'super-categories')
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
+	\caption{Coalescing the data from table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure \ref{fig:histogram-female-grouped}).}
+	\label{fig:histogram-male-grouped}
+\end{figure}
+
+% Histogram with the distribution of neutral pronouns for each occupation category
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-neutral}
+	\caption{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure \ref{fig:histogram-female}).
+Once again, STEM fields are predominantly concentrated on $X = 0$.}
+	\label{fig:histogram-neutral}
+\end{figure}
+
+% Histogram with the distirbution of neutral pronouns for each occupation category (but categories are grouped into 'super-categories')
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
+	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+Once again, STEM fields are predominantly concentrated at $X = 0$.}
+	\label{fig:histogram-neutral-grouped}
+\end{figure}
+
+The barplots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
+In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.
+
+% Barplot with % of female, male and neutral translated pronouns for each occupation category
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
-	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively.
-Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
+	\caption{Barplots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
+STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
+Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
 	\label{fig:gender-by-category}
 \end{figure}
+
+\section{Distribution of translated gender pronouns per language}
+
+We have taken the care of experimenting with a fair amount of different gender neutral languages.
+Because of that, another sensible way of coalescing our data is by language groups, as shown in Table \ref{tab:gender-by-language}.
+This can help us visualize the effect of different cultures in the genesis -- or lack thereof -- of gender bias.
+Nevertheless, the barplots in Figure \ref{fig:gender-by-language} are perhaps most useful to identifying the difficulty of extracting a gender pronoun when translating from certain languages.
+Japanese and Basque are good examples of this difficulty, although the quality of Turkish, Chinese and Yoruba translations are also compromised.
+
+% Table with our gender bias results grouped by language
+\begin{table}[H]
+\small{
+	\centering
+	\begin{tabular}{|c|c|c|c|}
+	\hline
+	Language 	& Female ($\%$) 	& Male ($\%$)		& Neutral ($\%$)	\\ \hline
+	\hline
+	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
+	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
+	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
+	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
+	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
+	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
+	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
+	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
+	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
+	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
+	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline \hline
+	Total     	& ? 				& ? 				& ? 				\\ \hline
+	\end{tabular}
+	\caption{Percentage of female, male and neutral gender pronouns obtained for each language, averaged over all occupations detailed in Table \ref{tab:gender-neutral-languages}.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\label{tab:gender-by-language}
+	}
+\end{table}
+
+% Barplot with % of female, male and neutral translated pronouns for each language
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/gender-by-language}
+	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
+Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
+Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}
+	\label{fig:gender-by-language}
+\end{figure}
+
+\section{Distribution of translated gender pronouns for varied adjectives}
+
+Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
+Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.
+
+Once again the data points towards male defaults, but some variation can be observed throughout different adjectives.
+Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).
+
+% Table with our gender bias results for each adjective
+\begin{table}[H]
+\small{
+	\centering
+	\begin{tabular}{|c|c|c|c|c|}
+	\hline
+	Adjective 	& Female 	& Male 	& Neutral 	& Total 	\\ \hline
+	\hline
+	Shy     	& 6  		& 2   	& 2  		& 12		\\ \hline
+	Attractive 	& 4  		& 2   	& 4  		& 12		\\ \hline
+	Happy   	& 5  		& 3   	& 2  		& 12		\\ \hline
+	Kind    	& 4  		& 3   	& 1  		& 12		\\ \hline
+	Ashamed 	& 4  		& 5   	& 1  		& 12		\\ \hline
+	Smart   	& 2  		& 5   	& 3  		& 12		\\ \hline
+	Envious 	& 2  		& 6   	& 1  		& 12		\\ \hline
+	Sad     	& 2  		& 6   	& 2  		& 12		\\ \hline
+	Loving  	& 2  		& 6   	& 2  		& 12		\\ \hline
+	Helpless 	& 2  		& 6   	& 2  		& 12		\\ \hline
+	Brave   	& 2  		& 7   	& 1  		& 12		\\ \hline
+	Proud   	& 2  		& 7   	& 1  		& 12		\\ \hline
+	Hateful 	& 1  		& 5   	& 3  		& 12		\\ \hline
+	Dumb    	& 1  		& 6   	& 2  		& 12		\\ \hline
+	Innocent 	& 1  		& 8   	& 1  		& 12		\\ \hline
+	Right   	& 0  		& 7   	& 3  		& 12		\\ \hline
+	Guilty 		& 0 		& 9		& 0			& 12		\\ \hline \hline	
+	Total   	& 43 		& 129 	& 41 		& 264		\\ \hline
+	\end{tabular}
+	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
+	\label{tab:gender-by-adjective}
+	}
+\end{table}
+
+% Barplot with % of female, male and neutral translated pronouns for each adjective
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/barplot-adjectives}
+	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
+One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns.
+Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
+	\label{fig:barplot-adjectives}
+\end{figure}
+
+\ifx
+\section{Results}
+
+For each one of the tested 462 occupations (see Tables \ref{tab:occupations}, \ref{tab:occupations-examples}), we used the Python Google Translate API (\url{http://py-googletrans.readthedocs.io/en/latest/}) to translate sentences built with the templates in Table \ref{tab:templates} from each one of the tested languages in Table \ref{tab:gender-neutral-languages} to English.
+The resulting sentences are then classified as \emph{female}, \emph{male} or \emph{neutral} according to their respective pronouns.
+Sentences starting with \say{She/She's/Her} are classified as female, sentences starting with \say{He/He's/His} are classified as male and sentences starting with \say{It/It's/Its/They/They're/Their} are classified as (gender) neutral.
+The results from this analysis, which can be found in \url{https://github.com/marceloprates/Gender-Bias}, are further discussed below.
+
+One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
+Sentences about occupations from the \emph{Corporate} and \emph{Science} category are also disproportionately translated with male pronouns (sex ratios $9.444$ and $10.5$ respectively), while those containing occupations from the \emph{Dance} category achieve a sex ratio of almost one ($1.064$).
+Not one category has achieved a balanced sex ratio, neither does any category exhibit more gender neutral than male pronouns.
+In total, female pronouns add up to $16.522\%$ among all categories, while male pronouns add up to $63.083\%$ and gender neutral pronouns to just $7.912\%$, yielding an average sex ratio of $3.818$.
 
@@
 
+% Table with our gender bias results grouped by language
 \begin{table}[H]
@@
 
+% Barplot with % of female, male and neutral translated pronouns for each language
 \begin{figure}[H]
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female}
-	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0.
-The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
-	\label{fig:histogram-female}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male}
-	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9.
-In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
-	\label{fig:histogram-male}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral}
-	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
-	\label{fig:histogram-neutral}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
-	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0.
-The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
-	\label{fig:histogram-female-grouped}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
-	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9.
-In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
-	\label{fig:histogram-male-grouped}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
-	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
-	\label{fig:histogram-neutral-grouped}
-\end{figure}
-
 Finally, Figure \ref{fig:barplot-adjectives} and Table \ref{tab:gender-by-adjective} show how stereotypical gender roles possibly play a part when simple adjectives are translated.
@@
 
+% Table with our gender bias results for each adjective
 \begin{table}[H]
@@
 
+% Barplot with % of female, male and neutral translated pronouns for each adjective
 \begin{figure}[H]
@@
 
+\fi
+
 \section{Conclusions}
@@
 By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process.
-Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet.
+Because Google Translate typically uses English as a \emph{lingua franca}
+% PEDRO: reference?
+ to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet.
 Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
@@
 
-%\bibliography{main}
-%\bibliographystyle{agsm}
+\bibliography{main}
 
-%\ifx
+\end{document}N
+
 \begin{thebibliography}{xx}
@@
 \end{thebibliography}
-%\fi
-
-\end{document}

### 2018-03-14 stats.jl

@@ -98,10 +98,12 @@ function barplot_adjectives()
         total           = sum([ count(x -> true,            aux[language]) for language in names(dat_adjectives)[1:end] ])
         ratio = male_count/female_count
         
-        push!(results_by_adjective, [ucfirst(adjective),"Female",   100*female_count/total,   ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Male",     100*male_count/total,     ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  100*neutral_count/total,  ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Female",   female_count    /1,   ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Male",     male_count      /1,     ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  neutral_count   /1,  ratio])
     end
+
+    CSV.write("Results/results-by-adjective.dat",results_by_adjective)
 
     sort!(results_by_adjective, cols = [order(:Ratio)])
 
@@ -124,6 +126,8 @@ function histograms_occupations()
 
         push!(results_by_occupation, [female_count,male_count,neutral_count,ucfirst(aux[1,:Category])])
     end
+
+    CSV.write("Results/results-by-occupation.dat",results_by_occupation)
 
     # Save histograms
     p = plot(results_by_occupation, x=:Female, color=:Category, Guide.xlabel("\# Female Pronouns"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=6pt))
@@ -150,9 +154,9 @@ function barplots_category()
         neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(aux) ])
         total           = sum([ count(x -> true,            aux[language]) for language in names(aux) ])
 
-        push!(results_by_category,[ucfirst(category),"Female",  100 * female_count    / total ])
-        push!(results_by_category,[ucfirst(category),"Male",    100 * male_count      / total ])
-        push!(results_by_category,[ucfirst(category),"Neutral", 100 * neutral_count   / total ])
+        push!(results_by_category,[ucfirst(category),"Female",  round(100 * female_count    / total,3) ])
+        push!(results_by_category,[ucfirst(category),"Male",    round(100 * male_count      / total,3) ])
[…]

### 2018-03-14 Paper/main.tex
Fixed bug in stats

@@
 	\hline		
-	Education / Training / Library 			& $25.207$ 		& $59.917$		& $2.893$		\\ \hline
-	Business / Financial 					& $6.324$ 		& $67.194$		& $7.312$		\\ \hline
-	Office / Administrative					& $10.031$		& $62.382$		& $7.732$		\\ \hline
-	Healthcare support						& $29.545$		& $51.705$		& $7.955$		\\ \hline
-	Management								& $9.486$		& $68.182$		& $3.953$		\\ \hline
-	Installation / Maintenance / Repair		& $3.596$		& $68.332$		& $11.588$		\\ \hline
-	Healthcare practitioners / technical	& $21.564$		& $61.734$		& $6.977$		\\ \hline
-	Community and social service			& $15.584$ 		& $63.636$		& $4.545$		\\ \hline
-	Sales and related						& $8.117$ 		& $69.156$		& $8.766$		\\ \hline
-	Production								& $14.36$ 		& $53.306$		& $13.120$		\\ \hline
-	Architecture and engineering			& $2.194$		& $79.624$		& $5.016$		\\ \hline
-	Life, physical, and social sciences		& $3.743$ 		& $82.888$		& $3.743$		\\ \hline
-	Transportation / material moving		& $7.532$ 		& $66.234$		& $11.688$		\\ \hline
-	Arts / Entertainment					& $7.862$ 		& $71.007$		& $4.423$		\\ \hline
-	Legal									& $11.688$ 		& $67.532$		& $5.195$		\\ \hline
-	Protective service						& $6.818$ 		& $75.000$		& $5.844$		\\ \hline
-	Food prepaparation / serving			& $16.578$ 		& $59.893$		& $10.160$		\\ \hline
-	Farming, fishing, and forestry			& $10.490$ 		& $62.238$		& $8.392$		\\ \hline
-	Computer and mathematical				& $2.841$ 		& $76.136$		& $4.545$		\\ \hline
-	Personal care and service				& $20.937$ 		& $55.096$		& $9.091$		\\ \hline
-	Construction and extraction				& $9.492$ 		& $61.631$		& $11.096$		\\ \hline
-	Cleaning / Maintenance					& $11.818$		& $63.636$		& $8.182$		\\ \hline
+	Education / Training / Library 			& $23.737$ 		& $58.081$		& $3.535$		\\ \hline
+	Business / Financial 					& $6.039$		& $62.319$		& $8.937$		\\ \hline
+	Office / Administrative					& $9.962$		& $56.322$		& $9.451$		\\ \hline
+	Healthcare support						& $27.083$		& $50.000$		& $9.722$		\\ \hline
+	Management								& $11.594$		& $61.111$		& $4.831$		\\ \hline
+	Installation / Maintenance / Repair		& $3.297$		& $62.393$		& $14.164$		\\ \hline
+	Healthcare practitioners / technical	& $18.088$		& $61.499$		& $8.527$		\\ \hline
+	Community and social service			& $14.286$ 		& $60.317$		& $5.556$		\\ \hline
+	Sales and related						& $7.540$		& $64.683$		& $10.714$		\\ \hline
+	Production								& $15.32$		& $45.370$		& $15.825$		\\ \hline
+	Architecture and engineering			& $1.916$		& $75.862$		& $6.130$		\\ \hline
+	Life, physical, and social sciences		& $3.595$		& $80.065$		& $4.575$		\\ \hline
+	Transportation / material moving		& $7.778$		& $60.159$		& $14.286$		\\ \hline
+	Arts / Entertainment					& $7.808$		& $66.366$		& $5.405$		\\ \hline
+	Legal									& $9.524$ 		& $65.079$		& $6.349$		\\ \hline
+	Protective service						& $6.746$		& $71.032$		& $7.143$		\\ \hline
+	Food prepaparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
+	Farming, fishing, and forestry			& $12.821$ 		& $53.846$		& $10.256$		\\ \hline
+	Computer and mathematical				& $3.472$		& $70.833$		& $5.556$		\\ \hline
+	Personal care and service				& $17.845$ 		& $52.862$		& $11.111$		\\ \hline
+	Construction and extraction				& $9.967$		& $54.739$		& $13.562$		\\ \hline
+	Cleaning / Maintenance					& $12.222$		& $57.778$		& $10.000$		\\ \hline
 	\hline
-	Total           						& $11.107$ 		& $63.030$ 		& $9.278$ 		\\ \hline
+	Total           						& $11.078$ 		& $57.398$ 		& $11.286$ 		\\ \hline
 	\end{tabular}
@@
 	\hline
-	Education								& $25.207$				& $59.917$				& $2.893$			\\ \hline
-	Corporate								& $7.955$				& $68.030$				& $6.364$			\\ \hline
-	Service									& $9.221$				& $64.987$				& $9.506$			\\ \hline
-	Healthcare								& $23.729$				& $59.014$				& $7.242$			\\ \hline
-	Production								& $14.360$				& $53.306$				& $13.120$			\\ \hline
-	STEM									& $2.992$				& $80.322$				& $4.373$			\\ \hline
-	Arts / Entertainment					& $7.862$				& $71.007$				& $4.423$			\\ \hline
-	Legal									& $11.688$				& $67.532$				& $5.195$			\\ \hline
-	Farming / Fishing / Forestry			& $10.490$				& $62.238$				& $8.392$			\\ \hline
-	Construction / Extraction				& $9.492$				& $61.631$				& $11.096$			\\ \hline
+	Education								& $23.737$				& $58.081$				& $3.535$			\\ \hline
+	Corporate								& $8.519$				& $62.407$				& $7.778$			\\ \hline
+	Service									& $8.730$				& $59.746$				& $11.619$			\\ \hline
+	Healthcare								& $20.527$				& $58.380$				& $8.851$			\\ \hline
+	Production								& $15.32$				& $45.370$				& $15.825$			\\ \hline
+	STEM									& $2.954$				& $76.653$				& $5.345$			\\ \hline
+	Arts / Entertainment					& $7.808$				& $66.366$				& $5.405$			\\ \hline
+	Legal									& $9.524$				& $65.079$				& $6.349$			\\ \hline
+	Farming / Fishing / Forestry			& $12.821$				& $53.846$				& $10.256$			\\ \hline
+	Construction / Extraction				& $9.967$				& $54.739$				& $13.562$			\\ \hline
 	\hline
-	Total           						& $11.107$ 				& $63.030$ 				& $9.278$ 			\\ \hline
+	Total           						& $11.078$ 				& $57.398$ 				& $11.286$ 			\\ \hline
 	\end{tabular}
@@
 	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
-	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline \hline
-	Total     	& ? 				& ? 				& ? 				\\ \hline
+	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline
 	\end{tabular}
@@
 	\hline
-	Shy     	& 6  		& 2   	& 2  		& 12		\\ \hline
-	Attractive 	& 4  		& 2   	& 4  		& 12		\\ \hline
-	Happy   	& 5  		& 3   	& 2  		& 12		\\ \hline
-	Kind    	& 4  		& 3   	& 1  		& 12		\\ \hline
-	Ashamed 	& 4  		& 5   	& 1  		& 12		\\ \hline
-	Smart   	& 2  		& 5   	& 3  		& 12		\\ \hline
-	Envious 	& 2  		& 6   	& 1  		& 12		\\ \hline
-	Sad     	& 2  		& 6   	& 2  		& 12		\\ \hline
-	Loving  	& 2  		& 6   	& 2  		& 12		\\ \hline
-	Helpless 	& 2  		& 6   	& 2  		& 12		\\ \hline
-	Brave   	& 2  		& 7   	& 1  		& 12		\\ \hline
-	Proud   	& 2  		& 7   	& 1  		& 12		\\ \hline
-	Hateful 	& 1  		& 5   	& 3  		& 12		\\ \hline
-	Dumb    	& 1  		& 6   	& 2  		& 12		\\ \hline
-	Innocent 	& 1  		& 8   	& 1  		& 12		\\ \hline
-	Right   	& 0  		& 7   	& 3  		& 12		\\ \hline
-	Guilty 		& 0 		& 9		& 0			& 12		\\ \hline \hline	
-	Total   	& 43 		& 129 	& 41 		& 264		\\ \hline
+	Happy		& $4$		& $3$	& $2$		& $11$		\\ \hline
+	Wrong		& $0$		& $6$	& $4$		& $11$		\\ \hline
+	Smart		& $2$		& $4$	& $3$		& $11$		\\ \hline
+	Ashamed		& $3$		& $5$	& $1$		& $11$		\\ \hline
+	Envious		& $2$		& $5$	& $1$		& $11$		\\ \hline
+	Modest		& $1$		& $5$	& $3$		& $11$		\\ \hline
+	Innocent	& $1$		& $7$	& $1$		& $11$		\\ \hline
+	Sad			& $2$		& $5$	& $2$		& $11$		\\ \hline
+	Afraid		& $1$		& $5$	& $0$		& $11$		\\ \hline
+	Dumb		& $1$		& $5$	& $2$		& $11$		\\ \hline
+	Kind		& $3$		& $3$	& $1$		& $11$		\\ \hline
+	Loving		& $2$		& $5$	& $2$		& $11$		\\ \hline
+	Arrogant	& $0$		& $8$	& $1$		& $11$		\\ \hline
+	Helpless	& $2$		& $5$	& $2$		& $11$		\\ \hline
+	Right		& $0$		& $6$	& $3$		& $11$		\\ \hline
+	Brave		& $2$		& $6$	& $1$		& $11$		\\ \hline
+	Proud		& $1$		& $7$	& $1$		& $11$		\\ \hline
+	Cruel		& $1$		& $7$	& $2$		& $11$		\\ \hline
+	Hateful		& $1$		& $4$	& $3$		& $11$		\\ \hline
+	Guilty		& $0$		& $8$	& $0$		& $11$		\\ \hline
+	Shy			& $5$		& $2$	& $2$		& $11$		\\ \hline
+	Attractive	& $4$		& $1$	& $4$		& $11$		\\ \hline \hline
+	Total 		& $38$		& $112$	& $41$		& -			\\ \hline 
 	\end{tabular}

### 2018-03-14 stats.jl

@@ -92,15 +92,15 @@ function barplot_adjectives()
     for adjective in dat_adjectives[:Adjective]
         aux = dat_adjectives[ dat_adjectives[:,:Adjective] .== adjective, : ]
         
-        female_count    = sum([ count(x -> x=="Female",     aux[language]) for language in names(dat_adjectives)[1:end] ])
-        male_count      = sum([ count(x -> x=="Male",       aux[language]) for language in names(dat_adjectives)[1:end] ])
-        neutral_count   = sum([ count(x -> x=="Neutral",    aux[language]) for language in names(dat_adjectives)[1:end] ])
-        total           = sum([ count(x -> true,            aux[language]) for language in names(dat_adjectives)[1:end] ])
+        female_count    = count(x -> x=="Female",   convert(Array,aux[:,3:end]))
+        male_count      = count(x -> x=="Male",     convert(Array,aux[:,3:end]))
+        neutral_count   = count(x -> x=="Neutral",  convert(Array,aux[:,3:end]))
+        total           = count(x -> true,          convert(Array,aux[:,3:end]))
         ratio = male_count/female_count
         
-        push!(results_by_adjective, [ucfirst(adjective),"Female",   female_count    /1,   ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Male",     male_count      /1,     ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  neutral_count   /1,  ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Female",   100 * female_count    / total,  ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Male",     100 * male_count      / total,  ratio])
+        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  100 * neutral_count   / total,  ratio])
     end
 
     CSV.write("Results/results-by-adjective.dat",results_by_adjective)
@@ -115,13 +115,13 @@ end
 function histograms_occupations()
     # Compute results table (grouped by occupation, averaged among languages)
     results_by_occupation = DataFrame(Female=[],Male=[],Neutral=[],Category=[])
-    for occupation in dat_jobs[:Occupation]
[…]

### 2018-03-14 stats.jl
Beginning to try to print a barplot histogram. Why does the Julia code take so long to run?

@@ -1,5 +1,20 @@
+#Pkg.add("DataFrames")
+#Pkg.add("CSV")
+#Pkg.add("Gadfly")
+#Pkg.add("Cairo")
+#Pkg.add("Fontconfig")
 
 using DataFrames, CSV, Gadfly
+
+# Read compare data file
+cpm_GT_real = CSV.read("Results/translate-real-histogram-data.tsv",delim='\t',nullable=false)
+function histograms_compare()
+    # Save histograms
+    p = plot(cpm_GT_real, x="Place", color="Type", Guide.xlabel("\# in bin"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=6pt))
+    draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 5.0inch, 3.75inch),p)
+end
+histograms_compare()
+exit()
 
 # Read job-genders.csv into a Julia DataFrame
 dat_jobs = CSV.read("Results/job-genders.tsv",delim='\t',nullable=false)
@@ -216,8 +231,10 @@ function a()
     draw(PDF("test.pdf", 5.0inch, 3.75inch), p)
 end
 
-a()
+
+#a()
 #barplot_adjectives()
 #histograms_occupations()
 #barplots_category()
 #barplots_language()
+

### 2018-03-15 stats.jl
Added all possible tsv formats to plot the compare data with

@@ -7,10 +7,11 @@
 using DataFrames, CSV, Gadfly
 
 # Read compare data file
-cpm_GT_real = CSV.read("Results/translate-real-histogram-data.tsv",delim='\t',nullable=false)
+cpm_GT_real = CSV.read("Results/translate-real-histogram-data-raw.tsv",delim='\t',nullable=false)
 function histograms_compare()
     # Save histograms
-    p = plot(cpm_GT_real, x="Place", color="Type", Guide.xlabel("\# in bin"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=6pt))
+    # "GT_F/A"	"GT_F/(M+F)"	"BLS_Data"
+    p = plot(cpm_GT_real, x="GT_F/A", Geom.histogram(bincount=12), Theme(key_max_columns=1, key_label_font_size=6pt))
     draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 5.0inch, 3.75inch),p)
 end
 histograms_compare()

### 2018-03-15 stats.jl
Updated abstract and conclusion

@@ -1,17 +1,13 @@
-#Pkg.add("DataFrames")
-#Pkg.add("CSV")
-#Pkg.add("Gadfly")
-#Pkg.add("Cairo")
-#Pkg.add("Fontconfig")
 
 using DataFrames, CSV, Gadfly
 
 # Read compare data file
-cpm_GT_real = CSV.read("Results/translate-real-histogram-data-raw.tsv",delim='\t',nullable=false)
+cpm_GT_real = CSV.read("Results/translate-real-histogram-data.tsv",delim='\t',nullable=false)
 function histograms_compare()
     # Save histograms
     # "GT_F/A"	"GT_F/(M+F)"	"BLS_Data"
-    p = plot(cpm_GT_real, x="GT_F/A", Geom.histogram(bincount=12), Theme(key_max_columns=1, key_label_font_size=6pt))
+    cpm_GT_real[:,2] *= 100
+    p = plot(cpm_GT_real, x="12-quantile", y="Frequency (%)", color=:Data, Geom.bar(position=:dodge), Guide.xticks(ticks=[1:12;]), Theme(key_max_columns=1, key_label_font_size=6pt))
     draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 5.0inch, 3.75inch),p)
 end
 histograms_compare()
@@ -231,7 +227,6 @@ function a()
     p = plot(aux2, x=:Count, color=:Gender, Geom.histogram)
     draw(PDF("test.pdf", 5.0inch, 3.75inch), p)
 end
-
 
 #a()
 #barplot_adjectives()

### 2018-03-19 stats.jl
Added figures

@@ -81,7 +81,7 @@ function group_categories(label)
     end
 end
 
-dat_jobs[:Category] = map(x -> rename_category(x), dat_jobs[:Category])
+dat_jobs[:Category] = map(x -> group_categories(x), dat_jobs[:Category])
 
 # Read adjective-genders.csv into a Julia DataFrame
 dat_adjectives = CSV.read("adjective-genders.csv",delim=';',nullable=false)
@@ -109,7 +109,7 @@ function barplot_adjectives()
 
     # Save bar plot
     p = plot(results_by_adjective, x=:Adjective, y=:Count, color=:Gender, Coord.Cartesian(ymin=0,ymax=100), Geom.bar(position=:stack), Theme(background_color="white",grid_color="gray",bar_highlight=colorant"black"), Guide.ylabel("\%"))
-    draw(PDF("Paper/pictures/barplot-adjectives.pdf", 5inch, 3.75inch),p)
+    draw(PDF("Paper/pictures/barplot-adjectives.pdf", 7.50inch, 5.625inch),p)
 end
 
 function histograms_occupations()
@@ -130,14 +130,14 @@ function histograms_occupations()
     CSV.write("Results/results-by-occupation.dat",results_by_occupation)
 
     # Save histograms
-    p = plot(results_by_occupation, x=:Female, color=:Category, Guide.xlabel("\# Female Pronouns"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=6pt))
-    draw(PDF("Paper/pictures/histogram-female-grouped.pdf", 5.0inch, 3.75inch),p)
+    p = plot(results_by_occupation, x=:Female, color=:Category, Guide.xlabel("\# Female Pronouns"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=8pt, plot_padding=[1cm,1cm,1cm,1cm]))
+    draw(PDF("Paper/pictures/histogram-female-grouped.pdf", 7.50inch, 5.625inch),p)
 
-    p = plot(results_by_occupation, x=:Male, color=:Category, Guide.xlabel("\# Male Pronouns"), Geom.histogram, Theme(key_max_columns=1, key_label_font_size=6pt))
-    draw(PDF("Paper/pictures/histogram-male-grouped.pdf", 5.0inch, 3.75inch),p)
[…]

### 2018-03-19 Paper/main.tex
Beginning to try to print a barplot histogram. Why does the Julia code take so long to run?

@@
 \begin{abstract}
-Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices.
-A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas.
-Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI.
-In this paper, we construct sentences with gender neutral pronouns in different languages 
-and translate these sentences into English using the Google Translate API. 
-The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate
-API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
-masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science.
-On the other hand, sentences about artistic fields are less likely to yield male defaults.
-This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
+Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias.
+A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying people of color as gorillas.
+Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.
+
+In this paper, we obtained a comprehensive list of job positions from the U.S.
+Bureau of Labor Statistics and used it to build sentences in the format of \say{He/She is an Engineer} in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc.
+We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output.
+We then show that Google Translate exhibits a strong tendency towards male defaults, which is exaggerated for fields suggested to be troubled with gender stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs.
+We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce the real-world distribution of female workers.
+In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.
+
+We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need for augmented current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
 \end{abstract}
@@
 We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
+
+\section{Assumptions and Preliminaries}
 
@@
 % PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
-These values contrast with LBS' report of gender participation.
-Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
-The variance reported for the translation results is also lower, at $0.0145145752$ in contrast with the report's $0.0666766998$.
-Altogether, the Pearson correlation between the two values is $0.3035612739$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$).
-This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.
+These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section \ref{sec:comparison-women-participation}.
 
@@
 
+\section{Comparison with women participation data across job positions}
+
+A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
+We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects a real-world asymmetry in the workplace.
+In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.
+
+The U.S.
+Bureau of Labor Statistics data summarized in Table \ref{tab:occupations} contains statistics about the percentage of women participation in each occupation category.
+These data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile.
+We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure \ref{fig:histogram-compare-gt-real}.
+The data shows us that Google Translate outputs fail to follow the real-world distribution of female workers across a comprehensive set of job positions.
+The distribution of translated female pronouns is consistently inversely distributed, with female pronouns accumulating in the first 12-quantile.
+By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.
+
+% Histogram with the distirbution of male prononuns for each occupation category
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histogram-compare-gt-real}
+	\caption{Women participation ($\%$) data obtained from the U.S.
+Bureau of Labor Statistics allows us to assess whether the Google Translate bias towards male defaults is at least to some extent explained by small frequencies of female workers in some job positions.
+Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles.
+To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral prononuns).}
+	\label{fig:histogram-compare-gt-real}
+\end{figure}
+
+Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
+The variance reported for the translation results is also lower, at $\approx 0.01451$ in contrast with the report's $\approx 0.06668$.
+Altogether, the Pearson correlation between the two values is $\approx 0.30356$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$).
+This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.
+
 \ifx
@@
 \section{Conclusions}
-In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them.
-As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society.
-By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process.
-Because Google Translate typically uses English as a \emph{lingua franca}
+
+In this paper, we have provided preliminary evidence that statistical translation tools such as Google Translate can exhibit gender biases and a strong tendency towards male defaults.
+Although implicit, these biases possibly stem from the real world data which is used to train them, and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace.
+In this paper, we suggest that and test the hypothesis that statistical translation tools can be probed to yield insights about stereotypical gender roles in our society -- or at least in their training data.
+By translating professional-related sentences such as \say{He/She is an engineer} from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs.
+Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs.
+And Because Google Translate typically uses English as a \emph{lingua franca}
 % PEDRO: reference?
- to translate between other languages, our findings about gender bias in professional occupations possibly generalize to most translations from gender neutral idioms in the same tool, although we have not tested this hypothesis yet.
-Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones).
-We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese.
-Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.
+to translate between other languages (e.g.
+Chinese $\rightarrow$ English $\rightarrow$ Portuguese), our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.
 
-We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potentially harmful effects of the pressing modern concerns of machine bias.
+Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
+Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
+Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese.
+Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.
+
+To solidify our results, we ran our pronominal gender translation statistics against the U.S.
+Bureau of Labor Statistics data on the frequency of women participation for each job position.
+Although Google Translate exhibits male defaults, this phenomenon may merely reflect the unequal distribution of male and female workers in some job positions.
+To test this hypothesis, we compared the distribution of female workers with the frequency of female translations, finding no correlation between said variables.
+Our data shows that Google Translate outputs fail to reflect the real-world distribution of female workers, under-estimating the expected frequency.
+That is to say that even if we do not expect a 50:50 distribution of translated gender pronouns, Google Translate exhibits male defaults in a greater frequency that job occupation data alone would suggest.
+The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.
+
+We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the pressing modern concerns of machine bias.
+We are optimistic that unbiased results can be obtained with little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
 

### 2018-03-19 Paper/main.tex
small changes in paper

@@
 \documentclass[fleqn,10pt]{article}
+
+% Language support
+\usepackage[utf8]{inputenc}
+\usepackage{CJKutf8}
+\DeclareUnicodeCharacter{1EB9}{\d{e}}
 
@@
 \usepackage{array}
-\usepackage{dirtytalk}
 \usepackage{csquotes}
@@
 \usepackage{natbib}
-%\bibliographystyle{agsm}
-\bibliographystyle{named}
+\bibliographystyle{agsm}
+%\bibliographystyle{plain}
 \let\hardvardurl\url
@@
 
+\usepackage{color,soul}
+\usepackage{ifthen}
+\newboolean{mark}
+\setboolean{mark}{false}
+
 \title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
-
 \begin{document}
+\DeclareRobustCommand{\changes}[1]{\ifthenelse{\boolean{mark}}{\hl{#1}}{#1}}
+\newif\ifchanged
+\changedtrue
+\DeclareRobustCommand{\changess}[1]{\ifchanged{\hl{#1}}\else{#1 }\fi}
 
@@
 Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias.
-A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying people of color as gorillas.
+A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying black people as gorillas.
 Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.
 
-In this paper, we obtained a comprehensive list of job positions from the U.S.
-Bureau of Labor Statistics and used it to build sentences in the format of \say{He/She is an Engineer} in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc.
+% PEDRO: TODO: replace say with ``''
+\changess{In this paper, we obtained a comprehensive list of job positions from the U.S.
+Bureau of Labor Statistics and used it to build sentences in the format of ``He/She is an Engineer'' (where ``Engineer'' is replaced by the job position of interest) in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc.
 We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output.
 We then show that Google Translate exhibits a strong tendency towards male defaults, which is exaggerated for fields suggested to be troubled with gender stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs.
-We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce the real-world distribution of female workers.
-In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.
+We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce a real-world distribution of female workers.
+In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.}
 
-We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need for augmented current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
+\changes{We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need to augment current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.}
 \end{abstract}
@@
 
-Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
+Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an ``universal language'' \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
 The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}.
@@
 	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation.
-This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
+This screenshot from Google Translate shows how occupations from traditionally male-dominated fields\cite{WB2014} such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
 	\label{fig:screenshot-gtranslate-hungarian}
@@
 
-As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}.
-Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}.
-Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}.
-This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}.
+\changes{As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily}\cite{Gtranslate200daily2017}\changes{.
+Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative}\cite{TranslateCommunity}\changes{.
+Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as ``sexist''} \cite{AlgorithmGtranslateSexist2018}\changes{.
+This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes}\cite{bolukbasi2016man}\changes{.
 Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique.
-They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms}\cite{hajian2016algorithmic}\changes{.
+The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
 This paper intends to investigate whether it does.
-We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
+We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.}
 
@@
 
+\changes{In this paper we assume that a statistical translation tool should reflect at most, the inequality existent in society -- it is only logical that the tool will poll from examples that the society produced and as such will inevitably retain some of that bias.
+It has been argued that one's language affects its cognition of the world }\cite{kay1984sapir}\changes{, and this leads to the discussion that languages that distinguish between female and male genders gramatically may enforce a bias in the person's perception of the world, with some studies corroborating this, as shown in }\cite{boroditsky2003sex}\changes{, as well some relating this with sexism }\cite{thompson2014linguistic}\changes{ and gender inequalities }\cite{santacreu2013female}.
+
+\changes{With this in mind, one can argue that a move towards gender neutrality in language and communication should be striven as a means to promote more gender equality in the world.
+Thus, in languages where gender neutrality can be achieved -- such as is the case of English -- it would be a valid aim to create translation tools that keeps the gender-neutrality of texts translated to such a language, instead of defaulting to male or female variants.}
+
+\changes{We will thus assume throughout this paper that although the distribution of translated gender pronouns may deviate from 50:50, it should not deviate to the extent of misrepresenting the demographics of job positions.
+That is to say we shall assume that Google Translate incorporates a negative gender bias if the frequency of male defaults super-estimates the (possibly unequal) distribution of male employees per female employee in a given occupation.}
+
 \section{Materials and Methods}
 
-% The hungarian punctuation in 'ő' is not rendered in the PDF. This needs to be fixed somehow. The same problem extends to Table 4 (label tab:templates)
 We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
-Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate.
-As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when \say{nurse} is replaced by \say{engineer}.
+Specifically, we can translate sentences such as the Hungarian ``ő ápolónő'', where ``ápolónő'' translates to ``nurse'' and ``ő'' is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result ``she's a nurse'' on Google Translate.
+As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when ``nurse'' is replaced by ``engineer''.
 The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
 Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
-Also, in order to solidify our results, we have decided to work
-% PEDRO: Removed "with as many gender neutral languages as possible"
- with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures \cite{wals} and other sources.
-Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
+Also, in order to solidify our results, we have decided to work \changes{with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures }\cite{wals}\changes{ and other sources.} Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
 (1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
@@
 To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-This time we resorted to using the Bureau of Labor Statistics' detailed occupations table \cite{BLS2017}, from the United States Department of Labor, for providing such a list.
+\changes{This time we resorted to using the Bureau of Labor Statistics' detailed occupations table }\cite{BLS2017}\changes{, from the United States Department of Labor, for providing such a list.
 The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
 Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers.
-We then separated the curated jobs into categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}.
-Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
-Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.
+We filtered some of these because they were too generic ( ``Computer occupations, all other'', and others) or because they had gender specific words for the profession (``host/hostess'', ``waiter/waitress'').
+We then separated the curated jobs into broader categories (Artistic, Corporate, Theatre, etc.) as shown in Table }\ref{tab:occupations}\changes{.
+Finally, Table }\ref{tab:occupations-examples}\changes{ shows thirty examples of randomly selected occupations from our dataset.
+Finally, we have selected a small list of 21 adjectives, presented in Table }\ref{tab:adjectives}.
 
@@
 	\end{tabular}
-	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+	\caption{Selected gender neutral languages obtained \changes{from WALS and other sources}.
      Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
@@
 
+% PEDRO: TODO highlight table
 % List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
@@
 	\end{tabular}
-	\caption{Selected occupations obtained from the U.S.
-Bureau of Labor Statistics \url{https://???}, grouped by category.
+	\caption{\changes{Selected occupations obtained from the U.S.
+Bureau of Labor Statistics }\url{https://www.bls.gov/cps/cpsaat11.htm}\changes{, grouped by category.
 We obtained a total of $1019$ occupations from $22$ distinct categories.
-We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
+We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}}
 	\label{tab:occupations}
@@
 	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
-	Japanese	& $\langle occupation \rangle$ です			\\ \hline
+	Japanese	& $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}			\\ \hline
 	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
@@
 
+% PEDRO: TODO highlight table
 % Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
@@
 	\end{tabular}
-	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
+	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of \changes{$1019$} different occupations.}
 	\label{tab:occupations-examples}
@@
 
+% PEDRO: TODO highlight section
 \section{Distribution of translated gender pronouns per occupation category}
 
-A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
-What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability
-% Add statistical test here
-than it does either female or gender-neutral pronouns. Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes,
-% Add reference here
-such as life and physical sciences, architecture, engineering, computer science and mathematics.
-Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
-For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).
+\changes{A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
+What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability than it does either female or gender-neutral pronouns.
+Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes, such as life and physical sciences, architecture, engineering, computer science and mathematics }\cite{moss2015can}.\changes{ Table }\ref{tab:gender-by-category}\changes{ summarizes these data, and Table }\ref{tab:gender-by-category-grouped}\changes{ summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
+For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).}
 
+% PEDRO: TODO highlight table
 % Table with our gender bias results grouped by category
@@
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\caption{\changes{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, averaged over all occupations in said category and tested languages detailed in Table }\ref{tab:gender-neutral-languages}\changes{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-category}
@@
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\caption{\changes{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation category, averaged over all occupations in said category and tested languages detailed in Table }\ref{tab:gender-neutral-languages}\changes{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-category-grouped}
@@
 
-Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed.
-The histogram in Figure \ref{fig:histogram-female} (and its coalesced variant in Figure \ref{fig:histogram-female-grouped}) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures \ref{fig:histogram-neutral} and \ref{fig:histogram-neutral-grouped} --, while the same data for male pronouns (shown in Figures \ref{fig:histogram-male}, \ref{fig:histogram-male-grouped}) suggests a skew normal distribution.
-Furthermore we can see both on Figures \ref{fig:histogram-female-grouped} and \ref{fig:histogram-male-grouped} how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.
+\changes{Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed.
+The histogram in Figure }\ref{fig:histogram-female}\changes{ (and its coalesced variant in Figure }\ref{fig:histogram-female-grouped}\changes{) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures }\ref{fig:histogram-neutral}\changes{ and }\ref{fig:histogram-neutral-grouped}\changes{ --, while the same data for male pronouns (shown in Figures }\ref{fig:histogram-male}\changes{, }\ref{fig:histogram-male-grouped}\changes{) suggests a skew normal distribution.
+Furthermore we can see both on Figures }\ref{fig:histogram-female-grouped}\changes{ and }\ref{fig:histogram-male-grouped}\changes{ how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.}
 
-% PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
-These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section \ref{sec:comparison-women-participation}.
+\changes{These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section }\ref{sec:comparison-women-participation}.
 
@@
 	\includegraphics[width=10cm]{pictures/histogram-female}
-	\caption{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
-STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table \ref{tab:gender-by-category}) extend to higher values.}
+	\caption{\changes{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
+STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table }\ref{tab:gender-by-category}\changes{) extend to higher values.}}
 	\label{fig:histogram-female}
@@
 
-% Histogram with the distirbution of female pronouns for each occupation category (but categories are grouped into 'super-categories')
+% Histogram with the distribution of female pronouns for each occupation category (but categories are grouped into 'super-categories')
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure \ref{fig:histogram-female}.}
+	\caption{\changes{Coalescing the data from Table }\ref{tab:gender-by-category}\changes{ by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure }\ref{fig:histogram-female}\changes{.}}
 	\label{fig:histogram-female-grouped}
@@
 
-% Histogram with the distirbution of male prononuns for each occupation category
+% Histogram with the distribution of male prononuns for each occupation category
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-male}
-	\caption{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$. One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}
+	\caption{\changes{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$.
+One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}}
 	\label{fig:histogram-male}
@@
 
-% Histogram with the distirbution of male pronouns for each occupation category (but categories are grouped into 'super-categories')
+% Histogram with the distribution of male pronouns for each occupation category (but categories are grouped into 'super-categories')
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
-	\caption{Coalescing the data from table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure \ref{fig:histogram-female-grouped}).}
+	\caption{\changes{Coalescing the data from table }\ref{tab:gender-by-category}\changes{ by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure }\ref{fig:histogram-female-grouped}\changes{).}}
 	\label{fig:histogram-male-grouped}
@@
 	\includegraphics[width=10cm]{pictures/histogram-neutral}
-	\caption{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure \ref{fig:histogram-female}).
-Once again, STEM fields are predominantly concentrated on $X = 0$.}
+	\caption{\changes{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure }\ref{fig:histogram-female}\changes{).
+Once again, STEM fields are predominantly concentrated on $X = 0$.}}
 	\label{fig:histogram-neutral}
@@
 
-% Histogram with the distirbution of neutral pronouns for each occupation category (but categories are grouped into 'super-categories')
+% Histogram with the distribution of neutral pronouns for each occupation category (but categories are grouped into 'super-categories')
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-Once again, STEM fields are predominantly concentrated at $X = 0$.}
+	\caption{\changes{Coalescing the data from Table }\ref{tab:gender-by-category}\changes{ by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+Once again, STEM fields are predominantly concentrated at $X = 0$.}}
 	\label{fig:histogram-neutral-grouped}
@@
 
-The barplots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
-In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.
+\changes{The barplots in Figure }\ref{fig:gender-by-category}\changes{ help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
+In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.}
 
@@
 	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
-	\caption{Barplots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
+	\caption{\changes{Barplots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table }\ref{tab:gender-by-category-grouped}\changes{) is composed of female, male and neutral terms.
 STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
-Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{fig:gender-by-category}
@@
 
+% PEDRO: TODO highlight section title
 \section{Distribution of translated gender pronouns per language}
 
-We have taken the care of experimenting with a fair amount of different gender neutral languages.
-Because of that, another sensible way of coalescing our data is by language groups, as shown in Table \ref{tab:gender-by-language}.
+\changes{We have taken the care of experimenting with a fair amount of different gender neutral languages.
+Because of that, another sensible way of coalescing our data is by language groups, as shown in Table }\ref{tab:gender-by-language}\changes{.
 This can help us visualize the effect of different cultures in the genesis -- or lack thereof -- of gender bias.
-Nevertheless, the barplots in Figure \ref{fig:gender-by-language} are perhaps most useful to identifying the difficulty of extracting a gender pronoun when translating from certain languages.
-Japanese and Basque are good examples of this difficulty, although the quality of Turkish, Chinese and Yoruba translations are also compromised.
+Nevertheless, the barplots in Figure }\ref{fig:gender-by-language}\changes{ are perhaps most useful to identifying the difficulty of extracting a gender pronoun when translating from certain languages.
+Japanese and Basque are good examples of this difficulty, although the quality of Turkish, Chinese and Yoruba translations are also compromised.}
 
@@
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each language, averaged over all occupations detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\caption{\changes{Percentage} of female, male and neutral gender pronouns \changes{obtained for each language, averaged over all occupations detailed in Table }\ref{tab:gender-neutral-languages}\changes{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-language}
@@
 	\includegraphics[width=10cm]{pictures/gender-by-language}
-	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
+	\caption{\changes{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
 Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
-Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}
+Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}}
 	\label{fig:gender-by-language}
@@
 
+% PEDRO: TODO highlight section title
 \section{Distribution of translated gender pronouns for varied adjectives}
 
-Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
-Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.
+\changes{Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
+Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.}
 
-Once again the data points towards male defaults, but some variation can be observed throughout different adjectives.
-Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).
+\changes{Once again the data points towards male defaults, but some variation can be observed throughout different adjectives.
+Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).}
 
@@
 
-\section{Comparison with women participation data across job positions}
+\section{Comparison with women participation data across job positions}\label{sec:comparison-women-participation}
 
-A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
-We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects a real-world asymmetry in the workplace.
-In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.
+\changes{A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
+We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects the demographics of male-dominated fields (male-dominated fields can be considered those that have less than 25\% of women participation}\cite{WB2014}\changes{, according to the US Department of Labour's Women's Bureau).
+In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.}
 
-The U.S.
-Bureau of Labor Statistics data summarized in Table \ref{tab:occupations} contains statistics about the percentage of women participation in each occupation category.
+\changes{The U.S.
+Bureau of Labor Statistics data summarized in Table }\ref{tab:occupations}\changes{ contains statistics about the percentage of women participation in each occupation category.
 These data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile.
-We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure \ref{fig:histogram-compare-gt-real}.
+We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure }\ref{fig:histogram-compare-gt-real}\changes{.
 The data shows us that Google Translate outputs fail to follow the real-world distribution of female workers across a comprehensive set of job positions.
 The distribution of translated female pronouns is consistently inversely distributed, with female pronouns accumulating in the first 12-quantile.
-By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.
+By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.}
 
-% Histogram with the distirbution of male prononuns for each occupation category
+% Histogram with the distribution of male prononuns for each occupation category
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-compare-gt-real}
-	\caption{Women participation ($\%$) data obtained from the U.S.
+	\caption{\changes{Women participation ($\%$) data obtained from the U.S.
 Bureau of Labor Statistics allows us to assess whether the Google Translate bias towards male defaults is at least to some extent explained by small frequencies of female workers in some job positions.
 Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles.
-To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral prononuns).}
+To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral prononuns).}}
 	\label{fig:histogram-compare-gt-real}
@@
 
-Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
+\changes{Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
 The variance reported for the translation results is also lower, at $\approx 0.01451$ in contrast with the report's $\approx 0.06668$.
 Altogether, the Pearson correlation between the two values is $\approx 0.30356$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$).
-This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.
+This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.}
+
+\section{Conclusions}
+
+In this paper, we have provided \changes{preliminary} evidence that statistical translation tools such as Google Translate \changes{can} exhibit gender biases and a strong tendency towards male defaults.
+\changes{Although implicit, these biases possibly} stem from the real world data which is used to train them, \changes{and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace.
+In this paper}, we suggest that \changes{and test} the hypothesis that \changes{statistical translation} tools can be probed to yield insights about stereotypical gender roles in our society \changes{-- or at least in their training data.} By translating \changes{professional-related sentences such as} ``He/She is an engineer'' from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation \changes{outputs.
+Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs.
+And} because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages (e.g.
+Chinese $\rightarrow$ English $\rightarrow$ Portuguese) \cite{LanguageSupportNMT2018,boitet2010mt}, our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.\changes{    }
+
+\changes{Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
+Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
+Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese.
+Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.}
+
+\changes{To solidify our results, we ran our pronominal gender translation statistics against the U.S.
+Bureau of Labor Statistics data on the frequency of women participation for each job position.
+Although Google Translate exhibits male defaults, this phenomenon may merely reflect the unequal distribution of male and female workers in some job positions.
+To test this hypothesis, we compared the distribution of female workers with the frequency of female translations, finding no correlation between said variables.
+Our data shows that Google Translate outputs fail to reflect the real-world distribution of female workers, under-estimating the expected frequency.
+That is to say that even if we do not expect a 50:50 distribution of translated gender pronouns, Google Translate exhibits male defaults in a greater frequency that job occupation data alone would suggest.
+The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.}
+
+We think this work can shed further light on \changes{what we believe is a} pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the pressing modern concerns of machine bias.
+\changes{We are optimistic that unbiased results can be obtained with little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.}
+
+\begin{thebibliography}{xx}
+
+\harvarditem[Angwin et~al.]{Angwin, Larson, Mattu \harvardand\
+  Kirchner}{2016}{angwin2016machine}
+Angwin, J., Larson, J., Mattu, S. \harvardand\ Kirchner, L.  \harvardyearleft
+  2016\harvardyearright , `Machine bias: {T}here's software used across the
+  country to predict future criminals and it's biased against blacks'.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{https://www.
+  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}
+
+\harvarditem[Bahdanau et~al.]{Bahdanau, Cho \harvardand\
+  Bengio}{2014}{Bahdanau2014}
+Bahdanau, D., Cho, K. \harvardand\ Bengio, Y.  \harvardyearleft
+  2014\harvardyearright , `Neural machine translation by jointly learning to
+  align and translate', {\em CoRR} {\bf abs/1409.0473}.
+\newline\harvardurl{http://arxiv.org/abs/1409.0473}
+
+\harvarditem[Boitet et~al.]{Boitet, Blanchon, Seligman \harvardand\
+  Bellynck}{2010}{boitet2010mt}
+Boitet, C., Blanchon, H., Seligman, M. \harvardand\ Bellynck, V.
+  \harvardyearleft 2010\harvardyearright , Mt on and for the web, {\em in}
+  `Natural Language Processing and Knowledge Engineering (NLP-KE), 2010
+  International Conference on', IEEE, pp.~1--10.
+
+\harvarditem[Bolukbasi et~al.]{Bolukbasi, Chang, Zou, Saligrama \harvardand\
+  Kalai}{2016}{bolukbasi2016man}
+Bolukbasi, T., Chang, K.-W., Zou, J.~Y., Saligrama, V. \harvardand\ Kalai,
+  A.~T.  \harvardyearleft 2016\harvardyearright , Man is to computer programmer
+  as woman is to homemaker? debiasing word embeddings, {\em in} `Advances in
+  Neural Information Processing Systems', pp.~4349--4357.
+
+\harvarditem[Boroditsky et~al.]{Boroditsky, Schmidt \harvardand\
+  Phillips}{2003}{boroditsky2003sex}
+Boroditsky, L., Schmidt, L.~A. \harvardand\ Phillips, W.  \harvardyearleft
+  2003\harvardyearright , `Sex, syntax, and semantics', {\em Language in mind:
+  Advances in the study of language and thought} pp.~61--79.
+
+\harvarditem{{Bureau of Labor Statistics}}{2017}{BLS2017}
+{Bureau of Labor Statistics}  \harvardyearleft 2017\harvardyearright , "table
+  11: Employed persons by detailed occupation, sex, race, and hispanic or
+  latino ethnicity, 2017", Labor force statistics from the current population
+  survey, United States Department of Labor.
+
+\harvarditem{Carl \harvardand\ Way}{2003}{carl2003recent}
+Carl, M. \harvardand\ Way, A.  \harvardyearleft 2003\harvardyearright , {\em
+  Recent advances in example-based machine translation}, Vol.~21, Springer
+  Science \& Business Media.
+
+\harvarditem{Chomsky}{2011}{Chomsky2011}
+Chomsky, N.  \harvardyearleft 2011\harvardyearright , `The golden age: A look
+  at the original roots of artificial intelligence, cognitive science, and
+  neuroscience (partial transcript of an interview with {N}. {C}homsky at
+  {MIT150} {S}ymposia: Brains, minds and machines symposium'.
+\newblock Last visited 2017-12-26.
+\newline\harvardurl{https://chomsky.info/20110616/}
+
+\harvarditem{Dascal}{1982}{dascal1982universal}
+Dascal, M.  \harvardyearleft 1982\harvardyearright , `Universal language
+  schemes in {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson',
+  {\em Studia leibnitiana} {\bf 14}(1),~98--109.
+
+\harvarditem{Dryer \harvardand\ Haspelmath}{2013}{wals}
+Dryer, M.~S. \harvardand\ Haspelmath, M., eds  \harvardyearleft
+  2013\harvardyearright , {\em WALS Online}, Max Planck Institute for
+  Evolutionary Anthropology, Leipzig.
+\newline\harvardurl{http://wals.info/}
+
+\harvarditem[Firat et~al.]{Firat, Cho, Sankaran, Yarman{-}Vural \harvardand\
+  Bengio}{2017}{Firat2017}
+Firat, O., Cho, K., Sankaran, B., Yarman{-}Vural, F.~T. \harvardand\ Bengio, Y.
+   \harvardyearleft 2017\harvardyearright , `Multi-way, multilingual neural
+  machine translation', {\em Computer Speech {\&} Language} {\bf 45},~236--252.
+\newline\harvardurl{https://doi.org/10.1016/j.csl.2016.10.006}
+
+\harvarditem{Garcia}{2016}{garcia2016racist}
+Garcia, M.  \harvardyearleft 2016\harvardyearright , `Racist in the machine:
+  The disturbing implications of algorithmic bias', {\em World Policy Journal}
+  {\bf 33}(4),~111--117.
+
+\harvarditem{Google}{2018}{LanguageSupportNMT2018}
+Google  \harvardyearleft 2018\harvardyearright , `Language support for the
+  neural machine translation model'.
+\newblock Last visited 2018-3-19.
+\newline\harvardurl{https://cloud.google.com/translate/docs/languages\#languages-nmt}
+
+\harvarditem{Gordin}{2015}{gordin2015scientific}
+Gordin, M.~D.  \harvardyearleft 2015\harvardyearright , {\em Scientific Babel:
+  How science was done before and after global English}, University of Chicago
+  Press.
+
+\harvarditem[Hajian et~al.]{Hajian, Bonchi \harvardand\
+  Castillo}{2016}{hajian2016algorithmic}
+Hajian, S., Bonchi, F. \harvardand\ Castillo, C.  \harvardyearleft
+  2016\harvardyearright , Algorithmic bias: From discrimination discovery to
+  fairness-aware data mining, {\em in} `Proceedings of the 22nd ACM SIGKDD
+  international conference on knowledge discovery and data mining', ACM,
+  pp.~2125--2126.
+
+\harvarditem{Hutchins}{1986}{hutchins1986machine}
+Hutchins, W.~J.  \harvardyearleft 1986\harvardyearright , {\em Machine
+  translation: past, present, future}, Ellis Horwood Chichester.
+
+\harvarditem[Johnson et~al.]{Johnson, Schuster, Le, Krikun, Wu, Chen, Thorat,
+  Vi{\'{e}}gas, Wattenberg, Corrado, Hughes \harvardand\
+  Dean}{2017}{wu2016google}
+Johnson, M., Schuster, M., Le, Q.~V., Krikun, M., Wu, Y., Chen, Z., Thorat, N.,
+  Vi{\'{e}}gas, F.~B., Wattenberg, M., Corrado, G., Hughes, M. \harvardand\
+  Dean, J.  \harvardyearleft 2017\harvardyearright , `Google's multilingual
+  neural machine translation system: Enabling zero-shot translation', {\em
+  {TACL}} {\bf 5},~339--351.
+\newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}
+
+\harvarditem{Kay \harvardand\ Kempton}{1984}{kay1984sapir}
+Kay, P. \harvardand\ Kempton, W.  \harvardyearleft 1984\harvardyearright ,
+  `What is the sapir-whorf hypothesis?', {\em American anthropologist} {\bf
+  86}(1),~65--79.
+
+\harvarditem{Kelman}{2014}{TranslateCommunity}
+Kelman, S.  \harvardyearleft 2014\harvardyearright , `Translate community: Help
+  us improve google translate!'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://search.googleblog.com/2014/07/translate-community-help-us-improve.html}
+
+\harvarditem{Kirkpatrick}{2016}{kirkpatrick2016battling}
+Kirkpatrick, K.  \harvardyearleft 2016\harvardyearright , `Battling algorithmic
+  bias: how do we ensure algorithms treat us fairly?', {\em Communications of
+  the ACM} {\bf 59}(10),~16--17.
+
+\harvarditem{Koehn}{2009}{koehn2009statistical}
+Koehn, P.  \harvardyearleft 2009\harvardyearright , {\em Statistical machine
+  translation}, Cambridge University Press.
+
+\harvarditem[Koehn et~al.]{Koehn, Hoang, Birch, Callison{-}Burch, Federico,
+  Bertoldi, Cowan, Shen, Moran, Zens, Dyer, Bojar, Constantin \harvardand\
+  Herbst}{2007}{Moses2007}
+Koehn, P., Hoang, H., Birch, A., Callison{-}Burch, C., Federico, M., Bertoldi,
+  N., Cowan, B., Shen, W., Moran, C., Zens, R., Dyer, C., Bojar, O.,
+  Constantin, A. \harvardand\ Herbst, E.  \harvardyearleft
+  2007\harvardyearright , Moses: Open source toolkit for statistical machine
+  translation, {\em in} `{ACL} 2007, Proceedings of the 45th Annual Meeting of
+  the Association for Computational Linguistics, June 23-30, 2007, Prague,
+  Czech Republic'.
+\newline\harvardurl{http://aclweb.org/anthology/P07-2045}
+
+\harvarditem{Locke \harvardand\ Booth}{1955}{locke1955machine}
+Locke, W.~N. \harvardand\ Booth, A.~D.  \harvardyearleft 1955\harvardyearright
+  , {\em Machine translation of languages: fourteen essays}, Published jointly
+  by Technology Press of the Massachusetts Institute of Technology and Wiley,
+  New York.
+
+\harvarditem{Mills}{2017}{racistsoapdispenser2017}
+Mills, K.-A.  \harvardyearleft 2017\harvardyearright , `'{R}acist' soap
+  dispenser refuses to help dark-skinned man wash his hands - but {T}witter
+  blames 'technology''.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}
+
+\harvarditem[Moss-Racusin et~al.]{Moss-Racusin, Molenda \harvardand\
+  Cramer}{2015}{moss2015can}
+Moss-Racusin, C.~A., Molenda, A.~K. \harvardand\ Cramer, C.~R.
+  \harvardyearleft 2015\harvardyearright , `Can evidence impact attitudes?
+  public reactions to evidence of gender bias in stem fields', {\em Psychology
+  of Women Quarterly} {\bf 39}(2),~194--209.
+
+\harvarditem{Norvig}{2017}{norvig2017chomsky}
+Norvig, P.  \harvardyearleft 2017\harvardyearright , `On {C}homsky and the two
+  cultures of statistical learning'.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://norvig.com/chomsky.html}
+
+\harvarditem{Olson}{2018}{AlgorithmGtranslateSexist2018}
+Olson, P.  \harvardyearleft 2018\harvardyearright , `The algorithm that helped
+  google translate become sexist'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://www.forbes.com/sites/parmyolson/2018/02/15/the-algorithm-that-helped-google-translate-become-sexist/\#1c1122c27daa}
+
+\harvarditem{Papenfuss}{2017}{womanunlockphone2017}
+Papenfuss, M.  \harvardyearleft 2017\harvardyearright , `Woman in {C}hina says
+  colleague's face was able to unlock her i{P}hone {X}'.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double\_us\_5a332cbce4b0ff955ad17d50}
+
+\harvarditem[Santacreu-Vasut et~al.]{Santacreu-Vasut, Shoham \harvardand\
+  Gay}{2013}{santacreu2013female}
+Santacreu-Vasut, E., Shoham, A. \harvardand\ Gay, V.  \harvardyearleft
+  2013\harvardyearright , `Do female/male distinctions in language matter?
+  evidence from gender political quotas', {\em Applied Economics Letters} {\bf
+  20}(5),~495--498.
+
+\harvarditem{Schiebinger}{2014}{schiebinger2014scientific}
+Schiebinger, L.  \harvardyearleft 2014\harvardyearright , `Scientific research
+  must take gender into account', {\em Nature} {\bf 507}(7490),~9.
+
+\harvarditem{Shankland}{2017}{Gtranslate200daily2017}
+Shankland, S.  \harvardyearleft 2017\harvardyearright , `Google translate now
+  serves 200 million people daily'.
+\newblock Last visited 2018-3-12.
+\newline\harvardurl{https://www.cnet.com/news/google-translate-now-serves-200-million-people-daily/}
+
+\harvarditem{Thompson}{2014}{thompson2014linguistic}
+Thompson, A.~J.  \harvardyearleft 2014\harvardyearright , `Linguistic
+  relativity: can gendered languages predict sexist attitudes', {\em
+  Linguistics} .
+
+\harvarditem{Wang \harvardand\ Kosinski}{2017}{wang2017deep}
+Wang, Y. \harvardand\ Kosinski, M.  \harvardyearleft 2017\harvardyearright ,
+  `Deep neural networks are more accurate than humans at detecting sexual
+  orientation from facial images'.
+
+\harvarditem{Weaver}{1955}{weaver1955translation}
+Weaver, W.  \harvardyearleft 1955\harvardyearright , Translation, {\em in}
+  W.~N. Locke \harvardand\ A.~D. Booth, eds, `Machine translation of
+  languages', Vol.~14, Cambridge: Technology Press, MIT, pp.~15--23.
+\newblock Last visited 2017-12-17.
+\newline\harvardurl{http://www.mt-archive.info/Weaver-1949.pdf}
+
+\harvarditem{{Women’s Bureau}}{2014}{WB2014}
+{Women’s Bureau}  \harvardyearleft 2014\harvardyearright , "traditional and
+  nontraditional occupations.", Labor force statistics from the current
+  population survey, United States Department of Labor.
+
+\end{thebibliography}
+
+\end{document}
 
@@
 
-One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
+One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male-dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
 Sentences about occupations from the \emph{Corporate} and \emph{Science} category are also disproportionately translated with male pronouns (sex ratios $9.444$ and $10.5$ respectively), while those containing occupations from the \emph{Dance} category achieve a sex ratio of almost one ($1.064$).
@@
 \fi
-
-\section{Conclusions}
-
-In this paper, we have provided preliminary evidence that statistical translation tools such as Google Translate can exhibit gender biases and a strong tendency towards male defaults.
-Although implicit, these biases possibly stem from the real world data which is used to train them, and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace.
-In this paper, we suggest that and test the hypothesis that statistical translation tools can be probed to yield insights about stereotypical gender roles in our society -- or at least in their training data.
-By translating professional-related sentences such as \say{He/She is an engineer} from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs.
-Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs.
-And Because Google Translate typically uses English as a \emph{lingua franca}
-% PEDRO: reference?
-to translate between other languages (e.g.
-Chinese $\rightarrow$ English $\rightarrow$ Portuguese), our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.
-
-Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
-Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
-Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese.
-Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.
-
-To solidify our results, we ran our pronominal gender translation statistics against the U.S.
-Bureau of Labor Statistics data on the frequency of women participation for each job position.
-Although Google Translate exhibits male defaults, this phenomenon may merely reflect the unequal distribution of male and female workers in some job positions.
-To test this hypothesis, we compared the distribution of female workers with the frequency of female translations, finding no correlation between said variables.
-Our data shows that Google Translate outputs fail to reflect the real-world distribution of female workers, under-estimating the expected frequency.
-That is to say that even if we do not expect a 50:50 distribution of translated gender pronouns, Google Translate exhibits male defaults in a greater frequency that job occupation data alone would suggest.
-The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.
-
-We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the pressing modern concerns of machine bias.
-We are optimistic that unbiased results can be obtained with little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
-
-\bibliography{main}
-
-\end{document}N
-
-\begin{thebibliography}{xx}
-
-\harvarditem[Angwin et~al.]{Angwin, Larson, Mattu \harvardand\
-  Kirchner}{2016}{angwin2016machine}
-Angwin, J., Larson, J., Mattu, S. \harvardand\ Kirchner, L.  \harvardyearleft
-  2016\harvardyearright , `Machine bias: {T}here's software used across the
-  country to predict future criminals and it's biased against blacks'.
-\newblock Last visited 2017-12-17.
-\newline\harvardurl{https://www.
-  propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}
-
-\harvarditem[Bahdanau et~al.]{Bahdanau, Cho \harvardand\
-  Bengio}{2014}{Bahdanau2014}
-Bahdanau, D., Cho, K. \harvardand\ Bengio, Y.  \harvardyearleft
-  2014\harvardyearright , `Neural machine translation by jointly learning to
-  align and translate', {\em CoRR} {\bf abs/1409.0473}.
-\newline\harvardurl{http://arxiv.org/abs/1409.0473}
-
-\harvarditem[Bolukbasi et~al.]{Bolukbasi, Chang, Zou, Saligrama \harvardand\
-  Kalai}{2016}{bolukbasi2016man}
-Bolukbasi, T., Chang, K.-W., Zou, J.~Y., Saligrama, V. \harvardand\ Kalai,
-  A.~T.  \harvardyearleft 2016\harvardyearright , Man is to computer programmer
-  as woman is to homemaker? debiasing word embeddings, {\em in} `Advances in
-  Neural Information Processing Systems', pp.~4349--4357.
-
-\harvarditem{Carl \harvardand\ Way}{2003}{carl2003recent}
-Carl, M. \harvardand\ Way, A.  \harvardyearleft 2003\harvardyearright , {\em
-  Recent advances in example-based machine translation}, Vol.~21, Springer
-  Science \& Business Media.
-
-\harvarditem{Chomsky}{2011}{Chomsky2011}
-Chomsky, N.  \harvardyearleft 2011\harvardyearright , `The golden age: A look
-  at the original roots of artificial intelligence, cognitive science, and
-  neuroscience (partial transcript of an interview with {N}. {C}homsky at
-  {MIT150} {S}ymposia: Brains, minds and machines symposium'.
-\newblock Last visited 2017-12-26.
-\newline\harvardurl{https://chomsky.info/20110616/}
-
-\harvarditem{Dascal}{1982}{dascal1982universal}
-Dascal, M.  \harvardyearleft 1982\harvardyearright , `Universal language
-  schemes in {E}ngland and {F}rance, 1600-1800 comments on {J}ames {K}nowlson',
-  {\em Studia leibnitiana} {\bf 14}(1),~98--109.
-
-\harvarditem[Firat et~al.]{Firat, Cho, Sankaran, Yarman{-}Vural \harvardand\
-  Bengio}{2017}{Firat2017}
-Firat, O., Cho, K., Sankaran, B., Yarman{-}Vural, F.~T. \harvardand\ Bengio, Y.
-   \harvardyearleft 2017\harvardyearright , `Multi-way, multilingual neural
-  machine translation', {\em Computer Speech {\&} Language} {\bf 45},~236--252.
-\newline\harvardurl{https://doi.org/10.1016/j.csl.2016.10.006}
-
-\harvarditem{Garcia}{2016}{garcia2016racist}
-Garcia, M.  \harvardyearleft 2016\harvardyearright , `Racist in the machine:
-  The disturbing implications of algorithmic bias', {\em World Policy Journal}
-  {\bf 33}(4),~111--117.
-
-\harvarditem{Gordin}{2015}{gordin2015scientific}
-Gordin, M.~D.  \harvardyearleft 2015\harvardyearright , {\em Scientific Babel:
-  How science was done before and after global English}, University of Chicago
-  Press.
-
-\harvarditem[Hajian et~al.]{Hajian, Bonchi \harvardand\
-  Castillo}{2016}{hajian2016algorithmic}
-Hajian, S., Bonchi, F. \harvardand\ Castillo, C.  \harvardyearleft
-  2016\harvardyearright , Algorithmic bias: From discrimination discovery to
-  fairness-aware data mining, {\em in} `Proceedings of the 22nd ACM SIGKDD
-  international conference on knowledge discovery and data mining', ACM,
-  pp.~2125--2126.
-
-\harvarditem{Hutchins}{1986}{hutchins1986machine}
-Hutchins, W.~J.  \harvardyearleft 1986\harvardyearright , {\em Machine
-  translation: past, present, future}, Ellis Horwood Chichester.
-
-\harvarditem[Johnson et~al.]{Johnson, Schuster, Le, Krikun, Wu, Chen, Thorat,
-  Vi{\'{e}}gas, Wattenberg, Corrado, Hughes \harvardand\
-  Dean}{2017}{wu2016google}
-Johnson, M., Schuster, M., Le, Q.~V., Krikun, M., Wu, Y., Chen, Z., Thorat, N.,
-  Vi{\'{e}}gas, F.~B., Wattenberg, M., Corrado, G., Hughes, M. \harvardand\
-  Dean, J.  \harvardyearleft 2017\harvardyearright , `Google's multilingual
-  neural machine translation system: Enabling zero-shot translation', {\em
-  {TACL}} {\bf 5},~339--351.
-\newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}
-
-\harvarditem{Kelman}{2014}{TranslateCommunity}
-Kelman, S.  \harvardyearleft 2014\harvardyearright , `Translate community: Help
-  us improve google translate!'.
-\newblock Last visited 2018-3-12.
-\newline\harvardurl{https://search.googleblog.com/2014/07/translate-community-help-us-improve.html}
-
-\harvarditem{Kirkpatrick}{2016}{kirkpatrick2016battling}
-Kirkpatrick, K.  \harvardyearleft 2016\harvardyearright , `Battling algorithmic
-  bias: how do we ensure algorithms treat us fairly?', {\em Communications of
-  the ACM} {\bf 59}(10),~16--17.
-
-\harvarditem{Koehn}{2009}{koehn2009statistical}
-Koehn, P.  \harvardyearleft 2009\harvardyearright , {\em Statistical machine
-  translation}, Cambridge University Press.
-
-\harvarditem[Koehn et~al.]{Koehn, Hoang, Birch, Callison{-}Burch, Federico,
-  Bertoldi, Cowan, Shen, Moran, Zens, Dyer, Bojar, Constantin \harvardand\
-  Herbst}{2007}{Moses2007}
-Koehn, P., Hoang, H., Birch, A., Callison{-}Burch, C., Federico, M., Bertoldi,
-  N., Cowan, B., Shen, W., Moran, C., Zens, R., Dyer, C., Bojar, O.,
-  Constantin, A. \harvardand\ Herbst, E.  \harvardyearleft
-  2007\harvardyearright , Moses: Open source toolkit for statistical machine
-  translation, {\em in} `{ACL} 2007, Proceedings of the 45th Annual Meeting of
-  the Association for Computational Linguistics, June 23-30, 2007, Prague,
-  Czech Republic'.
-\newline\harvardurl{http://aclweb.org/anthology/P07-2045}
-
-\harvarditem{Locke \harvardand\ Booth}{1955}{locke1955machine}
-Locke, W.~N. \harvardand\ Booth, A.~D.  \harvardyearleft 1955\harvardyearright
-  , {\em Machine translation of languages: fourteen essays}, Published jointly
-  by Technology Press of the Massachusetts Institute of Technology and Wiley,
-  New York.
-
-\harvarditem{Mills}{2017}{racistsoapdispenser2017}
-Mills, K.-A.  \harvardyearleft 2017\harvardyearright , `'{R}acist' soap
-  dispenser refuses to help dark-skinned man wash his hands - but {T}witter
-  blames 'technology''.
-\newblock Last visited 2017-12-17.
-\newline\harvardurl{http://www.mirror.co.uk/news/world-news/racist-soap-dispenser-refuses-help-11004385}
-
-\harvarditem{Norvig}{2017}{norvig2017chomsky}
-Norvig, P.  \harvardyearleft 2017\harvardyearright , `On {C}homsky and the two
-  cultures of statistical learning'.
-\newblock Last visited 2017-12-17.
-\newline\harvardurl{http://norvig.com/chomsky.html}
-
-\harvarditem{Olson}{2018}{AlgorithmGtranslateSexist2018}
-Olson, P.  \harvardyearleft 2018\harvardyearright , `The algorithm that helped
-  google translate become sexist'.
-\newblock Last visited 2018-3-12.
-\newline\harvardurl{https://www.forbes.com/sites/parmyolson/2018/02/15/the-algorithm-that-helped-google-translate-become-sexist/\#1c1122c27daa}
-
-\harvarditem{Papenfuss}{2017}{womanunlockphone2017}
-Papenfuss, M.  \harvardyearleft 2017\harvardyearright , `Woman in {C}hina says
-  colleague's face was able to unlock her i{P}hone {X}'.
-\newblock Last visited 2017-12-17.
-\newline\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}
-
-\harvarditem{Schiebinger}{2014}{schiebinger2014scientific}
-Schiebinger, L.  \harvardyearleft 2014\harvardyearright , `Scientific research
-  must take gender into account', {\em Nature} {\bf 507}(7490),~9.
-
-\harvarditem{Shankland}{2017}{Gtranslate200daily2017}
-Shankland, S.  \harvardyearleft 2017\harvardyearright , `Google translate now
-  serves 200 million people daily'.
-\newblock Last visited 2018-3-12.
-\newline\harvardurl{https://www.cnet.com/news/google-translate-now-serves-200-million-people-daily/}
-
-\harvarditem{Wang \harvardand\ Kosinski}{2017}{wang2017deep}
-Wang, Y. \harvardand\ Kosinski, M.  \harvardyearleft 2017\harvardyearright ,
-  `Deep neural networks are more accurate than humans at detecting sexual
-  orientation from facial images'.
-
-\harvarditem{Weaver}{1955}{weaver1955translation}
-Weaver, W.  \harvardyearleft 1955\harvardyearright , Translation, {\em in}
-  W.~N. Locke \harvardand\ A.~D. Booth, eds, `Machine translation of
-  languages', Vol.~14, Cambridge: Technology Press, MIT, pp.~15--23.
-\newblock Last visited 2017-12-17.
-\newline\harvardurl{http://www.mt-archive.info/Weaver-1949.pdf}
-
-\end{thebibliography}

### 2018-03-19 stats.jl

@@ -8,7 +8,7 @@ function histograms_compare()
     # "GT_F/A"	"GT_F/(M+F)"	"BLS_Data"
     cpm_GT_real[:,2] *= 100
     p = plot(cpm_GT_real, x="12-quantile", y="Frequency (%)", color=:Data, Geom.bar(position=:dodge), Guide.xticks(ticks=[1:12;]), Theme(key_max_columns=1, key_label_font_size=6pt))
-    draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 5.0inch, 3.75inch),p)
+    draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 7.50inch, 5.625inch),p)
 end
 histograms_compare()
 exit()
@@ -231,5 +231,5 @@ end
 #a()
 #barplot_adjectives()
 #histograms_occupations()
-barplots_category()
-barplots_language()
+#barplots_category()
+#barplots_language()

### 2018-04-18 Paper/main.tex
Moved previous version of paper to paper/old, included new version, removed all /changess{}

--- 2018-03-19 Paper/main.tex
@@
 
+\usepackage{color,soul}
+\usepackage{ifthen}
+\newboolean{mark}
+
 \title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
+\begin{document}
 
-\begin{document}
+\newif\ifchanged
+\changedtrue
+\setboolean{mark}{true}
+\DeclareRobustCommand{\changess}[1]{\ifchanged{\hl{#1}}\else{#1 }\fi}
+\DeclareRobustCommand{\changess}[1]{\ifthenelse{\boolean{mark}}{\hl{#1}}{#1}}
 
@@
 Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias.
-A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying people of color as gorillas.
+A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying black people as gorillas.
 Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.
 
-In this paper, we obtained a comprehensive list of job positions from the U.S.
-Bureau of Labor Statistics and used it to build sentences in the format of \say{He/She is an Engineer} in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc.
+In this paper,  we start with a comprehensive list of job positions from the U.S.
+Bureau of Labor Statistics (BLS) and used it in order to build sentences in constructions like ``He/She is an Engineer'' (where ``Engineer'' is replaced by the job position of interest) in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, and several others.
 We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output.
-We then show that Google Translate exhibits a strong tendency towards male defaults, which is exaggerated for fields suggested to be troubled with gender stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs.
-We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce the real-world distribution of female workers.
-In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.
+We then show that Google Translate exhibits a strong tendency towards male defaults,  in particular for fields typically associated to unbalanced gender distribution or  stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs.
+We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce a real-world distribution of female workers.
+In summary, we provide experimental evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.
 
-We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need for augmented current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
+We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need to augment current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
 \end{abstract}
@@
 
-Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
-The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}.
+Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an ``universal language'' \citep{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \citep{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation.
+The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \citep{gordin2015scientific}.
 Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Noam Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation.
-As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \cite{hutchins1986machine}.
-Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical,Moses2007}.
-Statistical and example-based machine translation have been on the rise ever since \cite{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily.
+As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \citep{hutchins1986machine}.
+Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \citep{koehn2009statistical,Moses2007}.
+Statistical and example-based machine translation have been on the rise ever since \citep{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \citep{wu2016google}) amounting to over $200$ million users daily.
 
 In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism.
-Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \cite{Chomsky2011,norvig2017chomsky}.
-Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \cite{Chomsky2011}:
+Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \citep{Chomsky2011,norvig2017chomsky}.
+Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \citep{Chomsky2011}:
 
@@
 
-Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \cite{norvig2017chomsky}:
+Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \citep{norvig2017chomsky}:
 
@@
 
-Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models.
+Chomsky and Norvig's debate \citep{norvig2017chomsky} is a microcosm  of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models.
 Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception?
@@
 
-On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}.
+On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \citep{schiebinger2014scientific}.
 Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view.
-Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}.
-Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}.
-\emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal assymetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
-In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.
+Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \citep{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \citep{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \citep{wang2017deep}.
+Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \citep{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \citep{racistsoapdispenser2017}.
+\emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal asymmetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.
+In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \citep{kirkpatrick2016battling}.
 
@@
 	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation.
-This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
+This screenshot from Google Translate shows how occupations from traditionally male-dominated fields\citep{WB2014} such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
 	\label{fig:screenshot-gtranslate-hungarian}
@@
 
-As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}.
-Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}.
-Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}.
-This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}.
+As of 2018, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily\citep{Gtranslate200daily2017}.
+Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative\citep{TranslateCommunity}.
+Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as ``sexist'' \citep{AlgorithmGtranslateSexist2018}.
+This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes\citep{bolukbasi2016man}.
 Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique.
-They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
+They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms\citep{hajian2016algorithmic}.
+The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
 This paper intends to investigate whether it does.
@@
 
+In this paper we assume that a statistical translation tool should reflect at most the inequality existent in society -- it is only logical that a translation tool will poll from examples that  society produced and, as such, will inevitably retain some of that bias.
+It has been argued that one's language affects one's knowledge and cognition about the world \citep{kay1984sapir}, and this leads to the discussion that languages that distinguish between female and male genders grammatically may enforce a bias in the person's perception of the world, with some studies corroborating this, as shown in \citep{boroditsky2003sex}, as well some relating this with sexism \citep{thompson2014linguistic} and gender inequalities \citep{santacreu2013female}.
+
+With this in mind, one can argue that a move towards gender neutrality in language and communication should be striven as a means to promote improved gender equality.
+Thus, in languages where gender neutrality can be achieved -- such as English -- it would be a valid aim to create translation tools that keep the gender-neutrality of texts translated into such a language, instead of defaulting to male or female variants.
+
+We will thus assume throughout this paper that although the distribution of translated gender pronouns may deviate from 50:50, it should not deviate to the extent of misrepresenting the demographics of job positions.
+That is to say we shall assume that Google Translate incorporates a negative gender bias if the frequency of male defaults overestimates the (possibly unequal) distribution of male employees per female employee in a given occupation.
+
 \section{Materials and Methods}
 
-% The hungarian punctuation in 'ő' is not rendered in the PDF. This needs to be fixed somehow. The same problem extends to Table 4 (label tab:templates)
 We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool.
-Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate.
-As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when \say{nurse} is replaced by \say{engineer}.
-The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows.
+Specifically, we can translate sentences such as the Hungarian ``ő ápolónő'', where ``ápolónő'' translates to ``nurse'' and ``ő'' is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result ``she's a nurse'' on Google Translate.
+As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when ``nurse'' is replaced by ``engineer''.
+The same basic template can be ported to all other gender neutral languages, as depicted in Table \ref{tab:templates}.
 Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
-Also, in order to solidify our results, we have decided to work
-% PEDRO: Removed "with as many gender neutral languages as possible"
- with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures \cite{wals} and other sources.
+Also, in order to solidify our results, we have decided to work with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures (WALS) \citep{wals} and other sources.
 Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
@@
 To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-This time we resorted to using the Bureau of Labor Statistics' detailed occupations table \cite{BLS2017}, from the United States Department of Labor, for providing such a list.
+Here, we resorted to using the Bureau of Labor Statistics' detailed occupations table \citep{BLS2017}, from the United States Department of Labor, for providing such a list.
 The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
 Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers.
-We then separated the curated jobs into categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}.
+We filtered some of these because they were too generic ( ``Computer occupations, all other'', and others) or because they had gender specific words for the profession (``host/hostess'', ``waiter/waitress'').
+We then separated the curated jobs into broader categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}.
 Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
+For the occupations that had less than 50 thousand workers, and thus no data about the participation of women, we assumed that its women participation was that of its upper category.
 Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.
+All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
+They are not linked to external sites to avoid authors identification.
 
@@
 \begin{table}[H]
+\centering
 \begin{small}
-	\centering
 	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
@@
 	\end{tabular}
-	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}.
+    \end{small}
+	\caption{Selected gender neutral languages obtained from WALS and other sources.
      Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
@@
 	\label{tab:gender-neutral-languages}
-	\end{small}
 \end{table}
 
+% PEDRO: TODO highlight table
 % List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
@@
 	\end{tabular}
+    \end{small}
 	\caption{Selected occupations obtained from the U.S.
-Bureau of Labor Statistics \url{https://???}, grouped by category.
+Bureau of Labor Statistics \url{https://www.bls.gov/cps/cpsaat11.htm}, grouped by category.
 We obtained a total of $1019$ occupations from $22$ distinct categories.
@@
 	\label{tab:occupations}
-	\end{small}
 \end{table}
@@
 \begin{table}[H]
+\centering
 \small{
-	\centering
 	\begin{tabular}{|c|c|}
@@
 	\end{tabular}
+	}
     \caption{Templates used to infer gender biases in the translation to the English language.}
 	\label{tab:templates}
-	}
 \end{table}
 
+% PEDRO: TODO highlight table
 % Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
 \begin{table}[H]
+\centering
 \small{
-	\centering
 	\begin{tabular}{|c|c|c|c|c|}
@@
 	\end{tabular}
+	}
     \caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
 	\label{tab:occupations-examples}
-	}
 \end{table}
@@
 \begin{table}[H]
+\centering
 \small{
-	\centering
 	\begin{tabular}{|c|c|c|}
@@
 	\end{tabular}
+	}
     \caption{Adjectives}
 	\label{tab:adjectives}
-	}
 \end{table}
 
+% PEDRO: TODO highlight section
 \section{Distribution of translated gender pronouns per occupation category}
@@
 A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
-What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability
-% Add statistical test here
-than it does either female or gender-neutral pronouns. Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes,
-% Add reference here
-such as life and physical sciences, architecture, engineering, computer science and mathematics.
+What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability than it does either female or gender-neutral pronouns.
+Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes, such as life and physical sciences, architecture, engineering, computer science and mathematics \citep{moss2015can}.
 Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
@@
 
+% PEDRO: TODO highlight table
 % Table with our gender bias results grouped by category
 \begin{table}[H]
+\centering
 \small{
-	\centering
 	\begin{tabular}{|c|c|c|c|}
@@
 	Protective service						& $6.746$		& $71.032$		& $7.143$		\\ \hline
-	Food prepaparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
+	Food preparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
 	Farming, fishing, and forestry			& $12.821$ 		& $53.846$		& $10.256$		\\ \hline
@@
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
+	}
+    \caption{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
 Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
 	\label{tab:gender-by-category}
-	}
 \end{table}
@@
 \begin{table}[H]
+\centering
 \small{
-	\centering
 	\begin{tabular}{|c|c|c|c|}
@@
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
+	}
+    \caption{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
 Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
 	\label{tab:gender-by-category-grouped}
-	}
 \end{table}
@@
 
-% PEDRO: Numbers don't match, but shouldn't. One is a average over all translations and the one below is an average over the average:
 These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section \ref{sec:comparison-women-participation}.
@@
 
-% Histogram with the distirbution of male prononuns for each occupation category
+% Histogram with the distribution of male pronouns for each occupation category
 \begin{figure}[H]
@@
 	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
-	\caption{Coalescing the data from table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
 In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure \ref{fig:histogram-female-grouped}).}
@@
 
-The barplots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
+The bar plots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
 In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.
@@
 	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
-	\caption{Barplots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
+	\caption{Bar plots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
 STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
@@
 
+% PEDRO: TODO highlight section title
 \section{Distribution of translated gender pronouns for varied adjectives}
@@
 
-\section{Comparison with women participation data across job positions}
+\section{Comparison with women participation data across job positions}\label{sec:comparison-women-participation}
 
 A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
-We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects a real-world asymmetry in the workplace.
+We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects the demographics of male-dominated fields (male-dominated fields can be considered those that have less than 25\% of women participation\citep{WB2014}, according to the U.S.
+Department of Labor Women's Bureau).
 In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.
@@
 Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles.
-To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral prononuns).}
+To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral pronouns).}
 	\label{fig:histogram-compare-gt-real}
@@
 
-\ifx
-\section{Results}
-
-For each one of the tested 462 occupations (see Tables \ref{tab:occupations}, \ref{tab:occupations-examples}), we used the Python Google Translate API (\url{http://py-googletrans.readthedocs.io/en/latest/}) to translate sentences built with the templates in Table \ref{tab:templates} from each one of the tested languages in Table \ref{tab:gender-neutral-languages} to English.
-The resulting sentences are then classified as \emph{female}, \emph{male} or \emph{neutral} according to their respective pronouns.
-Sentences starting with \say{She/She's/Her} are classified as female, sentences starting with \say{He/He's/His} are classified as male and sentences starting with \say{It/It's/Its/They/They're/Their} are classified as (gender) neutral.
-The results from this analysis, which can be found in \url{https://github.com/marceloprates/Gender-Bias}, are further discussed below.
-
-One can see either in Table \ref{tab:gender-by-category} or Figure \ref{fig:gender-by-category} that not only does Google Translate exhibit a tendency towards male defaults, but also that this tendency is further enhanced for typically male dominated fields such as computer science (with a ratio of $17.857$ male pronouns per female pronoun).
-Sentences about occupations from the \emph{Corporate} and \emph{Science} category are also disproportionately translated with male pronouns (sex ratios $9.444$ and $10.5$ respectively), while those containing occupations from the \emph{Dance} category achieve a sex ratio of almost one ($1.064$).
-Not one category has achieved a balanced sex ratio, neither does any category exhibit more gender neutral than male pronouns.
-In total, female pronouns add up to $16.522\%$ among all categories, while male pronouns add up to $63.083\%$ and gender neutral pronouns to just $7.912\%$, yielding an average sex ratio of $3.818$.
-
-While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each culture possibly has on this issue.
-Table \ref{tab:gender-by-language} shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns.
-Figure \ref{fig:gender-by-language} shows that, when grouping by language, gender neutral pronouns can be more prominent than male pronouns at least in one case: translating sentences from Basque yields 153 neutral vs 50 male and 5 female pronouns.
-Unfortunately this is the exception rather than the rule, with Yoruba following after with 131 neutral vs 304 male and 12 female pronouns.
-In total, female pronouns add up to $18.687\%$ among all categories, while male pronouns add up to $81.626\%$ and gender neutral pronouns to $8.466\%$, yielding an average sex ratio of $4.368$ ($14.405\%$ larger than what we get from grouping among categories).
-It should be noted however that Japanese and Basque, the two languages which stood out from the behavior observed in Figure \ref{fig:gender-by-category}, are precisely the two that Google Translate found hardest to translate.
-These findings should, as a result, be taken with a grain of salt.
-
-% Table with our gender bias results grouped by language
-\begin{table}[H]
-\small{
-	\centering
-	\begin{tabular}{|c|c|c|c|}
-	\hline
-	Language 	& Female ($\%$) 	& Male ($\%$)		& Neutral ($\%$)	\\ \hline
-	\hline
-	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
-	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
-	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
-	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
-	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
-	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
-	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
-	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
-	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
-	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
-	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline \hline
-	Total     	& ? 				& ? 				& ? 				\\ \hline
-	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders per language in the translated sentences.
-The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in some languages such as Chinese, with almost 30 male pronouns for each female one.}
-	\label{tab:gender-by-language}
-	}
-\end{table}
-
-% Barplot with % of female, male and neutral translated pronouns for each language
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/gender-by-language}
-	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively.
-Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
-Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns.}
-	\label{fig:gender-by-language}
-\end{figure}
-
-Figures \ref{fig:histogram-female}, \ref{fig:histogram-male} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns.
-While the number of occurrences of male pronouns tends towards a normal distribution, the figure changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation.
-The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.
-
-Finally, Figure \ref{fig:barplot-adjectives} and Table \ref{tab:gender-by-adjective} show how stereotypical gender roles possibly play a part when simple adjectives are translated.
-We observed that objective statements such as \say{he/she is wrong/right/guilty/innocent} are biased towards male defaults, while statements concerning emotional states (\say{sad},\say{kind},\say{shy}) amass at the opposite extreme of the sex ratio spectrum.
-Unsurprisingly, the statement \say{he/she is attractive} is translated predominately with female pronouns.
-
-% Table with our gender bias results for each adjective
-\begin{table}[H]
-\small{
-	\centering
-	\begin{tabular}{|c|c|c|c|c|c|}
-	\hline
-	Adjective 	& Female & Male & Neutral & Ratio & Total \\ \hline
-	\hline
-	Shy     	& 6  & 2   & 2  & 0.333 	& 12  \\ \hline
-	Attractive 	& 4  & 2   & 4  & 0.500   	& 12  \\ \hline
-	Happy   	& 5  & 3   & 2  & 0.600   	& 12  \\ \hline
-	Kind    	& 4  & 3   & 1  & 0.750 	& 12  \\ \hline
-	Ashamed 	& 4  & 5   & 1  & 1.250 	& 12  \\ \hline
-	Smart   	& 2  & 5   & 3  & 2.500   	& 12  \\ \hline
-	Envious 	& 2  & 6   & 1  & 3.000   	& 12  \\ \hline
-	Sad     	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
-	Loving  	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
-	Helpless 	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
-	Brave   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
-	Proud   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
-	Hateful 	& 1  & 5   & 3  & 5.000   	& 12  \\ \hline
-	Dumb    	& 1  & 6   & 2  & 6.000   	& 12  \\ \hline
-	Innocent 	& 1  & 8   & 1  & 8.000   	& 12  \\ \hline
-	Right   	& 0  & 7   & 3  & -   		& 12  \\ \hline
-	\hline
-	Total   	& 43 & 129 & 41 & 3     	& 264 \\ \hline
-	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
-	\label{tab:gender-by-adjective}
-	}
-\end{table}
-
-% Barplot with % of female, male and neutral translated pronouns for each adjective
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/barplot-adjectives}
-	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
-One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns.
-Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
-	\label{fig:barplot-adjectives}
-\end{figure}
-
-\fi
-
 \section{Conclusions}
@@
 In this paper, we suggest that and test the hypothesis that statistical translation tools can be probed to yield insights about stereotypical gender roles in our society -- or at least in their training data.
-By translating professional-related sentences such as \say{He/She is an engineer} from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs.
+By translating professional-related sentences such as ``He/She is an engineer'' from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs.
 Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs.
-And Because Google Translate typically uses English as a \emph{lingua franca}
-% PEDRO: reference?
-to translate between other languages (e.g.
-Chinese $\rightarrow$ English $\rightarrow$ Portuguese), our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.
+And because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages (e.g.
+Chinese $\rightarrow$ English $\rightarrow$ Portuguese) \citep{LanguageSupportNMT2018,boitet2010mt}, our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.
 
 Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
-Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
+Adjectives such as \emph{Shy} and \emph{Attractive are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
 Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese.
-Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.
+Some languages such as Yoruba and Basque were found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule and these languages also exhibit a high frequency of translation errors.}
 
@@
 
-We think this work can shed further light on what we believe is a pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the pressing modern concerns of machine bias.
-We are optimistic that unbiased results can be obtained with little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
-
-\bibliography{main}
-
-\end{document}N
+We think this work shed new light on a  pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the current concerns about machine bias.
+We are optimistic that unbiased results can be obtained with relatively little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
 
@@
 
+\harvarditem[Boitet et~al.]{Boitet, Blanchon, Seligman \harvardand\
+  Bellynck}{2010}{boitet2010mt}
+Boitet, C., Blanchon, H., Seligman, M. \harvardand\ Bellynck, V.
+  \harvardyearleft 2010\harvardyearright , Mt on and for the web, {\em in}
+  `Natural Language Processing and Knowledge Engineering (NLP-KE), 2010
+  International Conference on', IEEE, pp.~1--10.
+
 \harvarditem[Bolukbasi et~al.]{Bolukbasi, Chang, Zou, Saligrama \harvardand\
@@
   Neural Information Processing Systems', pp.~4349--4357.
+
+\harvarditem[Boroditsky et~al.]{Boroditsky, Schmidt \harvardand\
+  Phillips}{2003}{boroditsky2003sex}
+Boroditsky, L., Schmidt, L.~A. \harvardand\ Phillips, W.  \harvardyearleft
+  2003\harvardyearright , `Sex, syntax, and semantics', {\em Language in mind:
+  Advances in the study of language and thought} pp.~61--79.
+
+\harvarditem{{Bureau of Labor Statistics}}{2017}{BLS2017}
+{Bureau of Labor Statistics}  \harvardyearleft 2017\harvardyearright , "table
+  11: Employed persons by detailed occupation, sex, race, and hispanic or
+  latino ethnicity, 2017", Labor force statistics from the current population
+  survey, United States Department of Labor.
+  \newblock Last visited 2018-3-13.
+  \newline\harvardurl{https://www.bls.gov/cps/cpsaat11.htm}
 
@@
 
+\harvarditem{Dryer \harvardand\ Haspelmath}{2013}{wals}
+Dryer, M.~S. \harvardand\ Haspelmath, M., eds  \harvardyearleft
+  2013\harvardyearright , {\em WALS Online}, Max Planck Institute for
+  Evolutionary Anthropology, Leipzig.
+\newline\harvardurl{http://wals.info/}
+
 \harvarditem[Firat et~al.]{Firat, Cho, Sankaran, Yarman{-}Vural \harvardand\
@@
   {\bf 33}(4),~111--117.
+
+\harvarditem{Google}{2018}{LanguageSupportNMT2018}
+Google  \harvardyearleft 2018\harvardyearright , `Language support for the
+  neural machine translation model'.
+\newblock Last visited 2018-3-19.
+\newline\harvardurl{https://cloud.google.com/translate/docs/languages\#languages-nmt}
 
@@
 \newline\harvardurl{https://transacl.org/ojs/index.php/tacl/article/view/1081}
+
+\harvarditem{Kay \harvardand\ Kempton}{1984}{kay1984sapir}
+Kay, P. \harvardand\ Kempton, W.  \harvardyearleft 1984\harvardyearright ,
+  `What is the sapir-whorf hypothesis?', {\em American anthropologist} {\bf
+  86}(1),~65--79.
 
@@
 
+\harvarditem[Moss-Racusin et~al.]{Moss-Racusin, Molenda \harvardand\
+  Cramer}{2015}{moss2015can}
+Moss-Racusin, C.~A., Molenda, A.~K. \harvardand\ Cramer, C.~R.
+  \harvardyearleft 2015\harvardyearright , `Can evidence impact attitudes?
+  public reactions to evidence of gender bias in stem fields', {\em Psychology
+  of Women Quarterly} {\bf 39}(2),~194--209.
+
 \harvarditem{Norvig}{2017}{norvig2017chomsky}
@@
 \newblock Last visited 2017-12-17.
-\newline\url{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double_us_5a332cbce4b0ff955ad17d50}
+\newline\harvardurl{http://www.huffpostbrasil.com/entry/iphone-face-recognition-double\_us\_5a332cbce4b0ff955ad17d50}
+
+\harvarditem[Santacreu-Vasut et~al.]{Santacreu-Vasut, Shoham \harvardand\
+  Gay}{2013}{santacreu2013female}
+Santacreu-Vasut, E., Shoham, A. \harvardand\ Gay, V.  \harvardyearleft
+  2013\harvardyearright , `Do female/male distinctions in language matter?
+  evidence from gender political quotas', {\em Applied Economics Letters} {\bf
+  20}(5),~495--498.
 
@@
 \newline\harvardurl{https://www.cnet.com/news/google-translate-now-serves-200-million-people-daily/}
+
+\harvarditem{Thompson}{2014}{thompson2014linguistic}
+Thompson, A.~J.  \harvardyearleft 2014\harvardyearright , `Linguistic
+  relativity: can gendered languages predict sexist attitudes', {\em
+  Linguistics} .
 
@@
 
+\harvarditem{{Women’s Bureau}}{2014}{WB2014}
+{Women’s Bureau}  \harvardyearleft 2014\harvardyearright , "traditional and
+  nontraditional occupations.", Labor force statistics from the current
+  population survey, United States Department of Labor.
+  \newblock Last visited 2018-3-17.
+  \newline\harvardurl{https://www.dol.gov/wb/stats/nontra\_traditional\_occupations.htm}
+
 \end{thebibliography}
+
+\section{Data Availability}
+All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
+They are not linked to external sites to avoid authors identification.
+
+\section{Competing interests}
+The authors declare that they have no competing financial interests.
+
+\end{document}

### 2018-04-18 Paper/old/main.tex
Moved previous version of paper to paper/old, included new version, removed all /changess{}

--- 2018-04-18 Paper/main.tex
@@
 \changedtrue
-\setboolean{mark}{true}
+\setboolean{mark}{false}
 \DeclareRobustCommand{\changess}[1]{\ifchanged{\hl{#1}}\else{#1 }\fi}
@@
 
-In this paper,  we start with a comprehensive list of job positions from the U.S.
+% PEDRO: TODO: replace say with ``''
+\changess{In this paper,  we start with a comprehensive list of job positions from the U.S.
 Bureau of Labor Statistics (BLS) and used it in order to build sentences in constructions like ``He/She is an Engineer'' (where ``Engineer'' is replaced by the job position of interest) in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, and several others.
@@
 We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce a real-world distribution of female workers.
-In summary, we provide experimental evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.
+In summary, we provide experimental evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.}
 
-We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need to augment current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
+\changess{We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need to augment current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.}
 \end{abstract}
@@
 	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation.
-This screenshot from Google Translate shows how occupations from traditionally male-dominated fields\citep{WB2014} such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
+This screenshot from Google Translate shows how occupations from traditionally male-dominated fields \citep{WB2014} such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
 	\label{fig:screenshot-gtranslate-hungarian}
@@
 
-As of 2018, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily\citep{Gtranslate200daily2017}.
-Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative\citep{TranslateCommunity}.
-Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as ``sexist'' \citep{AlgorithmGtranslateSexist2018}.
-This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes\citep{bolukbasi2016man}.
+\changess{As of 2018, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily}\citep{Gtranslate200daily2017}\changess{.
+Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative}\citep{TranslateCommunity}\changess{.
+Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as ``sexist''} \citep{AlgorithmGtranslateSexist2018}\changess{.
+This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes}\citep{bolukbasi2016man}\changess{.
 Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique.
-They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms\citep{hajian2016algorithmic}.
+They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms}\citep{hajian2016algorithmic}\changess{.
 The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist.
 This paper intends to investigate whether it does.
-We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.
+We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.}
 
@@
 
-In this paper we assume that a statistical translation tool should reflect at most the inequality existent in society -- it is only logical that a translation tool will poll from examples that  society produced and, as such, will inevitably retain some of that bias.
-It has been argued that one's language affects one's knowledge and cognition about the world \citep{kay1984sapir}, and this leads to the discussion that languages that distinguish between female and male genders grammatically may enforce a bias in the person's perception of the world, with some studies corroborating this, as shown in \citep{boroditsky2003sex}, as well some relating this with sexism \citep{thompson2014linguistic} and gender inequalities \citep{santacreu2013female}.
+\changess{In this paper we assume that a statistical translation tool should reflect at most the inequality existent in society -- it is only logical that a translation tool will poll from examples that  society produced and, as such, will inevitably retain some of that bias.
+It has been argued that one's language affects one's knowledge and cognition about the world }\citep{kay1984sapir}\changess{, and this leads to the discussion that languages that distinguish between female and male genders grammatically may enforce a bias in the person's perception of the world, with some studies corroborating this, as shown in }\citep{boroditsky2003sex}\changess{, as well some relating this with sexism }\citep{thompson2014linguistic}\changess{ and gender inequalities }\citep{santacreu2013female}.
 
-With this in mind, one can argue that a move towards gender neutrality in language and communication should be striven as a means to promote improved gender equality.
-Thus, in languages where gender neutrality can be achieved -- such as English -- it would be a valid aim to create translation tools that keep the gender-neutrality of texts translated into such a language, instead of defaulting to male or female variants.
+\changess{With this in mind, one can argue that a move towards gender neutrality in language and communication should be striven as a means to promote improved gender equality.
+Thus, in languages where gender neutrality can be achieved -- such as English -- it would be a valid aim to create translation tools that keep the gender-neutrality of texts translated into such a language, instead of defaulting to male or female variants.}
 
-We will thus assume throughout this paper that although the distribution of translated gender pronouns may deviate from 50:50, it should not deviate to the extent of misrepresenting the demographics of job positions.
-That is to say we shall assume that Google Translate incorporates a negative gender bias if the frequency of male defaults overestimates the (possibly unequal) distribution of male employees per female employee in a given occupation.
+\changess{We will thus assume throughout this paper that although the distribution of translated gender pronouns may deviate from 50:50, it should not deviate to the extent of misrepresenting the demographics of job positions.
+That is to say we shall assume that Google Translate incorporates a negative gender bias if the frequency of male defaults overestimates the (possibly unequal) distribution of male employees per female employee in a given occupation.}
 
@@
 Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
-Also, in order to solidify our results, we have decided to work with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures (WALS) \citep{wals} and other sources.
-Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
+Also, in order to solidify our results, we have decided to work \changess{with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures (WALS) }\citep{wals}\changess{ and other sources.} Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
 (1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.
@@
 To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-Here, we resorted to using the Bureau of Labor Statistics' detailed occupations table \citep{BLS2017}, from the United States Department of Labor, for providing such a list.
+\changess{Here, we resorted to using the Bureau of Labor Statistics' detailed occupations Table }\citep{BLS2017}\changess{, from the United States Department of Labor, for providing such a list.
 The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
@@
 We filtered some of these because they were too generic ( ``Computer occupations, all other'', and others) or because they had gender specific words for the profession (``host/hostess'', ``waiter/waitress'').
-We then separated the curated jobs into broader categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}.
-Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.
+We then separated the curated jobs into broader categories (Artistic, Corporate, Theatre, etc.) as shown in Table }\ref{tab:occupations}\changess{.
+Finally, Table }\ref{tab:occupations-examples}\changess{ shows thirty examples of randomly selected occupations from our dataset.
 For the occupations that had less than 50 thousand workers, and thus no data about the participation of women, we assumed that its women participation was that of its upper category.
-Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.
-All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
-They are not linked to external sites to avoid authors identification.
+Finally, we have selected a small list of 21 adjectives, presented in Table }\ref{tab:adjectives}.
 
 % Table containing a list of gender neutral languages with information about pronominal gender systems and GTranslate support
-\begin{table}[H]
-\centering
-\begin{small}
-	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
-	\hline
-	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \hline \hline
-	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
-									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \hline
-	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
-									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \hline
-	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
-									& English 				& \checkmark 				& \checkmark 	& \xmark 		\\
-									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \hline
-	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
-									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
-									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \hline
-	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Korean 				& O 						& \checkmark 	& \xmark		\\
-									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
-									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
-									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
-									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \hline
-	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
-									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
-									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
-									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
-	\end{tabular}
-    \end{small}
-	\caption{Selected gender neutral languages obtained from WALS and other sources.
-     Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
-For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded.
-Languages lacking support from Google Translate have been discarded.}
-	\label{tab:gender-neutral-languages}
-\end{table}
-
-% PEDRO: TODO highlight table
-% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
-\begin{table}[H]
-\begin{small}
-	\centering
-	\begin{tabular}{|m{3.5cm}|c|c|c|}
-	\hline	
-	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
-	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
-	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
-	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
-	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
-	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
-	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
-	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
-	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
-	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
-	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
-	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
-	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
-	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
-	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
-	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
-	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
-	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
-	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
-	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
-	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
-	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
-	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
-	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
-	\end{tabular}
-    \end{small}
-	\caption{Selected occupations obtained from the U.S.
-Bureau of Labor Statistics \url{https://www.bls.gov/cps/cpsaat11.htm}, grouped by category.
+%\begin{table}[H]
+%\centering
+%\begin{small}
+%	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
+%	\hline
+%	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \%hline \hline
+%	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
+%									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \%hline
+%	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
+%									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \%hline
+%	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
+%									& English 				& \checkmark 				& \checkmark 	& \xmark 		\\
+%									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \%hline
+%	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
+%									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
+%									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \%hline
+%	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Korean 				& O 						& \checkmark 	& \xmark		\\
+%									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
+%									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
+%									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
+%									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
+%									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \%hline
+%	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
+%									& Ido 					& O 						& \xmark 		& \xmark 		\\ \%cline{2-2}
+%									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \%cline{2-2}
+%									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \%hline
+%	\end{tabular}
+%    \end{small}
+%	\caption{Selected gender neutral languages obtained \changess{from WALS and other sources}.
+%     Languages are grouped according to language families and classified according to whether they exhibit pronominal gender %system (\checkmark: yes, \xmark:~no, O: it is optional).
+For the purposes of this work, we have decided to work only %with languages lacking such a system and, as such, Persian and Esperanto have been discarded.
+Languages lacking support %from Google Translate have been discarded.}
+%	\label{tab:gender-neutral-languages}
+%\end{table}
+%
+%% PEDRO: TODO highlight table
+%% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and %female participation (%) in each category
+%\begin{table}[H]
+%\begin{small}
+%	\centering
+%	\begin{tabular}{|m{3.5cm}|c|c|c|}
+%	\hline	
+%	Category 										& Group 						& \# Occupations 	& Female %Participation	\\ \hline \hline
+%	Education, training, and library 				& Education						& $22$ 				& $73.0\%$			%	\\ \hline
+%	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$			%	\\ \hline
+%	Office and administrative support				& Service						& $87$ 				& $72.2\%$			%	\\ \hline
+%	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$			%	\\ \hline
+%	Management 										& Corporate						& $46$ 				& $39.8\%$			%	\\ \hline
+%	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 			%	\\ \hline
+%	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$			%	\\ \hline
+%	Community and social service 					& Service						& $14$ 				& $66.1\%$			%	\\ \hline
+%	Sales and related 								& Corporate						& $28$ 				& $49.1\%$			%	\\ \hline
+%	Production 										& Production					& $264$ 			& $28.9\%$			%	\\ \hline
+%	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$			%	\\ \hline
+%	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$			%	\\ \hline
+%	Transportation and material moving				& Service						& $70$ 				& $17.3\%$			%	\\ \hline
+%	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$			%	\\ \hline
+%	Legal											& Legal 						& $7$				& $52.8\%$			%	\\ \hline
+%	Protective Service 								& Service 						& $28$ 				& $22.3\%$			%	\\ \hline
+%	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$			%	\\ \hline
+%	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$			%	\\ \hline
+%	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$			%	\\ \hline
+%	Personal care and service						& Service 						& $33$ 				& $76.1\%$			%	\\ \hline
+%	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 			%	\\ \hline
+%	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$			%	\\ \hline \hline
+%	Total											& - 							& $1019$			& $41.3\%$			%	\\ \hline
+%	\end{tabular}
+%    \end{small}
+%	\caption{\changess{Selected occupations obtained from the U.S.
+Bureau of Labor Statistics }\url{https://www.bls.gov/cps/%cpsaat11.htm}\changess{, grouped by category.
 We obtained a total of $1019$ occupations from $22$ distinct categories.
-We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
-	\label{tab:occupations}
-\end{table}
-
-% List of sentence templates for assessing gender bias in the translation from source language to English
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|}
-	\hline
-	Language 	& Sentence template 						\\ \hline \hline
-	Malay		& dia adalah $\langle occupation \rangle$ 	\\ \hline
-	Estonian	& ta on $\langle occupation \rangle$ 		\\ \hline
-	Finnish		& hän on $\langle occupation \rangle$ 		\\ \hline
-	Hungarian	& ő $\langle occupation \rangle$ 			\\ \hline
-	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
-	Japanese	& $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}			\\ \hline
-	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
-	Yoruba		& o jẹ $\langle occupation \rangle$ 		\\ \hline
-	Basque		& $\langle occupation \rangle$ da 			\\ \hline
-	Swahili		& yeye ni $\langle occupation \rangle$ 		\\ \hline
-	Chinese		& ta $\langle occupation \rangle$ 			\\ \hline
-	\end{tabular}
-	}
-    \caption{Templates used to infer gender biases in the translation to the English language.}
-	\label{tab:templates}
-\end{table}
-
-% PEDRO: TODO highlight table
-% Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|c|c|c|}
-	\hline
-	Insurance sales agent 	& Editor 						& Rancher 					\\ \hline
-	Ticket taker 			& Pile-driver operator 			& Tool maker 				\\ \hline
-	Jeweler 				& Judicial law clerk 			& Auditing clerk 			\\ \hline
-	Physician 				& Embalmer 						& Door-to-door salesperson 	\\ \hline
-	Packer 					& Bookkeeping clerk 			& Community health worker 	\\ \hline
-	Sales worker 			& Floor finisher 				& Social science technician \\ \hline
-	Probation officer 		& Paper goods machine setter 	& Heating installer 		\\ \hline
-	Animal breeder 			& Instructor 					& Teacher assistant 		\\ \hline
-	Statistical assistant 	& Shipping clerk 				& Trapper 					\\ \hline
-	Pharmacy aide 			& Sewing machine operator 		& Service unit operator 	\\ \hline
-	\end{tabular}
-	}
-    \caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of $1019$ different occupations.}
-	\label{tab:occupations-examples}
-\end{table}
-
-% Table with all adjectives we experimented with
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|c|}
-	\hline
-	Happy 		& Sad 		& Right 	\\ \hline
-	Wrong 		& Afraid	& Brave 	\\ \hline
-	Smart		& Dumb		& Proud 	\\ \hline
-	Ashamed		& Kind		& Cruel 	\\ \hline
-	Envious		& Loving	& Hateful 	\\ \hline
-	Modest 		& Arrogant	& Guilty	\\ \hline
-	Innocent	& Helpless	& Shy		\\ \hline
-	\end{tabular}
-	}
-    \caption{Adjectives}
-	\label{tab:adjectives}
-\end{table}
+%We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}}
+%	\label{tab:occupations}
+%\end{table}
+%
+%% List of sentence templates for assessing gender bias in the translation from source language to English
+%\begin{table}[H]
+%\centering
+%\small{
+%	\begin{tabular}{|c|c|}
+%	\hline
+%	Language 	& Sentence template 						\\ \hline \hline
+%	Malay		& dia adalah $\langle occupation \rangle$ 	\\ \hline
+%	Estonian	& ta on $\langle occupation \rangle$ 		\\ \hline
+%	Finnish		& hän on $\langle occupation \rangle$ 		\\ \hline
+%	Hungarian	& ő $\langle occupation \rangle$ 			\\ \hline
+%	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
+%	Japanese	& $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}			\\ \hline
+%	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
+%	Yoruba		& o jẹ $\langle occupation \rangle$ 		\\ \hline
+%	Basque		& $\langle occupation \rangle$ da 			\\ \hline
+%	Swahili		& yeye ni $\langle occupation \rangle$ 		\\ \hline
+%	Chinese		& ta $\langle occupation \rangle$ 			\\ \hline
+%	\end{tabular}
+%	}
+%    \caption{Templates used to infer gender biases in the translation to the English language.}
+%	\label{tab:templates}
+%\end{table}
+%
+%% PEDRO: TODO highlight table
+%% Table with a random sample of 30 occupations obtained from the U.S. Bureau of Labor Statistics
+%\begin{table}[H]
+%\centering
+%\small{
+%	\begin{tabular}{|c|c|c|c|c|}
+%	\hline
+%	Insurance sales agent 	& Editor 						& Rancher 					\\ \hline
+%	Ticket taker 			& Pile-driver operator 			& Tool maker 				\\ \hline
+%	Jeweler 				& Judicial law clerk 			& Auditing clerk 			\\ \hline
+%	Physician 				& Embalmer 						& Door-to-door salesperson 	\\ \hline
+%	Packer 					& Bookkeeping clerk 			& Community health worker 	\\ \hline
+%	Sales worker 			& Floor finisher 				& Social science technician \\ \hline
+%	Probation officer 		& Paper goods machine setter 	& Heating installer 		\\ \hline
+%	Animal breeder 			& Instructor 					& Teacher assistant 		\\ \hline
+%	Statistical assistant 	& Shipping clerk 				& Trapper 					\\ \hline
+%	Pharmacy aide 			& Sewing machine operator 		& Service unit operator 	\\ \hline
+%	\end{tabular}
+%	}
+%    \caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of \changess{$%1019$} different occupations.}
+%	\label{tab:occupations-examples}
+%\end{table}
+%
+%% Table with all adjectives we experimented with
+%\begin{table}[H]
+%\centering
+%\small{
+%	\begin{tabular}{|c|c|c|}
+%	\hline
+%	Happy 		& Sad 		& Right 	\\ \hline
+%	Wrong 		& Afraid	& Brave 	\\ \hline
+%	Smart		& Dumb		& Proud 	\\ \hline
+%	Ashamed		& Kind		& Cruel 	\\ \hline
+%	Envious		& Loving	& Hateful 	\\ \hline
+%	Modest 		& Arrogant	& Guilty	\\ \hline
+%	Innocent	& Helpless	& Shy		\\ \hline
+%	\end{tabular}
+%	}
+%    \caption{Adjectives}
+%	\label{tab:adjectives}
+%\end{table}
 
@@
 
-A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
+\changess{A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
 What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability than it does either female or gender-neutral pronouns.
-Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes, such as life and physical sciences, architecture, engineering, computer science and mathematics \citep{moss2015can}.
-Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
-For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).
+Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes, such as life and physical sciences, architecture, engineering, computer science and mathematics }\citep{moss2015can}.\changess{ Table }\ref{tab:gender-by-category}\changess{ summarizes these data, and Table }\ref{tab:gender-by-category-grouped}\changess{ summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
+For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).}
 
@@
 % Table with our gender bias results grouped by category
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|c|c|}
-	\hline
-	Category 								& Female ($\%$)	& Male ($\%$)	& Neutral ($\%$) \\ \hline
-	\hline		
-	Education / Training / Library 			& $23.737$ 		& $58.081$		& $3.535$		\\ \hline
-	Business / Financial 					& $6.039$		& $62.319$		& $8.937$		\\ \hline
-	Office / Administrative					& $9.962$		& $56.322$		& $9.451$		\\ \hline
-	Healthcare support						& $27.083$		& $50.000$		& $9.722$		\\ \hline
-	Management								& $11.594$		& $61.111$		& $4.831$		\\ \hline
-	Installation / Maintenance / Repair		& $3.297$		& $62.393$		& $14.164$		\\ \hline
-	Healthcare practitioners / technical	& $18.088$		& $61.499$		& $8.527$		\\ \hline
-	Community and social service			& $14.286$ 		& $60.317$		& $5.556$		\\ \hline
-	Sales and related						& $7.540$		& $64.683$		& $10.714$		\\ \hline
-	Production								& $15.32$		& $45.370$		& $15.825$		\\ \hline
-	Architecture and engineering			& $1.916$		& $75.862$		& $6.130$		\\ \hline
-	Life, physical, and social sciences		& $3.595$		& $80.065$		& $4.575$		\\ \hline
-	Transportation / material moving		& $7.778$		& $60.159$		& $14.286$		\\ \hline
-	Arts / Entertainment					& $7.808$		& $66.366$		& $5.405$		\\ \hline
-	Legal									& $9.524$ 		& $65.079$		& $6.349$		\\ \hline
-	Protective service						& $6.746$		& $71.032$		& $7.143$		\\ \hline
-	Food preparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
-	Farming, fishing, and forestry			& $12.821$ 		& $53.846$		& $10.256$		\\ \hline
-	Computer and mathematical				& $3.472$		& $70.833$		& $5.556$		\\ \hline
-	Personal care and service				& $17.845$ 		& $52.862$		& $11.111$		\\ \hline
-	Construction and extraction				& $9.967$		& $54.739$		& $13.562$		\\ \hline
-	Cleaning / Maintenance					& $12.222$		& $57.778$		& $10.000$		\\ \hline
-	\hline
-	Total           						& $11.078$ 		& $57.398$ 		& $11.286$ 		\\ \hline
-	\end{tabular}
-	}
-    \caption{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
-	\label{tab:gender-by-category}
-\end{table}
+%\begin{table}[H]
+%\centering
+%\small{
+%	\begin{tabular}{|c|c|c|c|}
+%	\hline
+%	Category 								& Female ($\%$)	& Male ($\%$)	& Neutral ($\%$) \\ \hline
+%	\hline		
+%	Education / Training / Library 			& $23.737$ 		& $58.081$		& $3.535$		\\ \hline
+%	Business / Financial 					& $6.039$		& $62.319$		& $8.937$		\\ \hline
+%	Office / Administrative					& $9.962$		& $56.322$		& $9.451$		\\ \hline
+%	Healthcare support						& $27.083$		& $50.000$		& $9.722$		\\ \hline
+%	Management								& $11.594$		& $61.111$		& $4.831$		\\ \hline
+%	Installation / Maintenance / Repair		& $3.297$		& $62.393$		& $14.164$		\\ \hline
+%	Healthcare practitioners / technical	& $18.088$		& $61.499$		& $8.527$		\\ \hline
+%	Community and social service			& $14.286$ 		& $60.317$		& $5.556$		\\ \hline
+%	Sales and related						& $7.540$		& $64.683$		& $10.714$		\\ \hline
+%	Production								& $15.32$		& $45.370$		& $15.825$		\\ \hline
+%	Architecture and engineering			& $1.916$		& $75.862$		& $6.130$		\\ \hline
+%	Life, physical, and social sciences		& $3.595$		& $80.065$		& $4.575$		\\ \hline
+%	Transportation / material moving		& $7.778$		& $60.159$		& $14.286$		\\ \hline
+%	Arts / Entertainment					& $7.808$		& $66.366$		& $5.405$		\\ \hline
+%	Legal									& $9.524$ 		& $65.079$		& $6.349$		\\ \hline
+%	Protective service						& $6.746$		& $71.032$		& $7.143$		\\ \hline
+%	Food preparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
+%	Farming, fishing, and forestry			& $12.821$ 		& $53.846$		& $10.256$		\\ \hline
+%	Computer and mathematical				& $3.472$		& $70.833$		& $5.556$		\\ \hline
+%	Personal care and service				& $17.845$ 		& $52.862$		& $11.111$		\\ \hline
+%	Construction and extraction				& $9.967$		& $54.739$		& $13.562$		\\ \hline
+%	Cleaning / Maintenance					& $12.222$		& $57.778$		& $10.000$		\\ \hline
+%	\hline
+%	Total           						& $11.078$ 		& $57.398$ 		& $11.286$ 		\\ \hline
+%	\end{tabular}
+%	}
+%    \caption{\changess{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, %averaged over all occupations in said category and tested languages detailed in Table }\ref{tab:gender-neutral-languages}%\changess{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for %which we cannot obtain a gender pronoun.}}
+%	\label{tab:gender-by-category}
+%\end{table}
+%
+%% Table with our gender bias results grouped by category (but categories are further grouped into "super-categories")
+%\begin{table}[H]
+%\centering
+%\small{
+%	\begin{tabular}{|c|c|c|c|}
+%	\hline
+%	Category 								& Female ($\%$)			& Male ($\%$)			& Neutral ($\%$)	\\ \hline
+%	\hline
+%	Education								& $23.737$				& $58.081$				& $3.535$			\\ \hline
+%	Corporate								& $8.519$				& $62.407$				& $7.778$			\\ \hline
+%	Service									& $8.730$				& $59.746$				& $11.619$			\\ \hline
+%	Healthcare								& $20.527$				& $58.380$				& $8.851$			\\ \hline
+%	Production								& $15.32$				& $45.370$				& $15.825$			\\ \hline
+%	STEM									& $2.954$				& $76.653$				& $5.345$			\\ \hline
+%	Arts / Entertainment					& $7.808$				& $66.366$				& $5.405$			\\ \hline
+%	Legal									& $9.524$				& $65.079$				& $6.349$			\\ \hline
+%	Farming / Fishing / Forestry			& $12.821$				& $53.846$				& $10.256$			\\ \hline
+%	Construction / Extraction				& $9.967$				& $54.739$				& $13.562$			\\ \hline
+%	\hline
+%	Total           						& $11.078$ 				& $57.398$ 				& $11.286$ 			\\ \hline
+%	\end{tabular}
+%	}
+%    \caption{\changess{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation %category, averaged over all occupations in said category and tested languages detailed in Table }\ref{%tab:gender-neutral-languages}\changess{.
+Note that columns do not in general add up to $100\%$, as there is a fair %amount of translated sentences for which we cannot obtain a gender pronoun.}}
+%	\label{tab:gender-by-category-grouped}
+%\end{table}
 
-% Table with our gender bias results grouped by category (but categories are further grouped into "super-categories")
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|c|c|}
-	\hline
-	Category 								& Female ($\%$)			& Male ($\%$)			& Neutral ($\%$)	\\ \hline
-	\hline
-	Education								& $23.737$				& $58.081$				& $3.535$			\\ \hline
-	Corporate								& $8.519$				& $62.407$				& $7.778$			\\ \hline
-	Service									& $8.730$				& $59.746$				& $11.619$			\\ \hline
-	Healthcare								& $20.527$				& $58.380$				& $8.851$			\\ \hline
-	Production								& $15.32$				& $45.370$				& $15.825$			\\ \hline
-	STEM									& $2.954$				& $76.653$				& $5.345$			\\ \hline
-	Arts / Entertainment					& $7.808$				& $66.366$				& $5.405$			\\ \hline
-	Legal									& $9.524$				& $65.079$				& $6.349$			\\ \hline
-	Farming / Fishing / Forestry			& $12.821$				& $53.846$				& $10.256$			\\ \hline
-	Construction / Extraction				& $9.967$				& $54.739$				& $13.562$			\\ \hline
-	\hline
-	Total           						& $11.078$ 				& $57.398$ 				& $11.286$ 			\\ \hline
-	\end{tabular}
-	}
-    \caption{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
-	\label{tab:gender-by-category-grouped}
-\end{table}
+\changess{Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed.
+The histogram in Figure }\ref{fig:histogram-female}\changess{ (and its coalesced variant in Figure }\ref{fig:histogram-female-grouped}\changess{) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures }\ref{fig:histogram-neutral}\changess{ and }\ref{fig:histogram-neutral-grouped}\changess{ --, while the same data for male pronouns (shown in Figures }\ref{fig:histogram-male}\changess{, }\ref{fig:histogram-male-grouped}\changess{) suggests a skew normal distribution.
+Furthermore we can see both on Figures }\ref{fig:histogram-female-grouped}\changess{ and }\ref{fig:histogram-male-grouped}\changess{ how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.}
 
-Plotting histograms for the number of gender pronouns per occupation category sheds further light on how female, male and gender-neutral pronouns are differently distributed.
-The histogram in Figure \ref{fig:histogram-female} (and its coalesced variant in Figure \ref{fig:histogram-female-grouped}) suggests that the number of female pronouns is inversely distributed -- which is mirrored in the data for gender-neutral pronouns in Figures \ref{fig:histogram-neutral} and \ref{fig:histogram-neutral-grouped} --, while the same data for male pronouns (shown in Figures \ref{fig:histogram-male}, \ref{fig:histogram-male-grouped}) suggests a skew normal distribution.
-Furthermore we can see both on Figures \ref{fig:histogram-female-grouped} and \ref{fig:histogram-male-grouped} how STEM fields (labeled in red) exhibit predominantly male defaults -- amounting predominantly near $X = 0$ in the female histogram although much to the right in the male histogram.
+\changess{These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section }\ref{sec:comparison-women-participation}.
 
-These values contrast with LBS' report of gender participation, which will be discussed in more detail in Section \ref{sec:comparison-women-participation}.
+%% Histogram with the distribution of female pronouns for each occupation category
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-female}
+%	\caption{\changess{The data for the total number $X$ of female pronouns per occupation category suggests an inverse %distribution.
+STEM fields such as life, social, computer and physical sciences, architecture, engineering and %mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and %healthcare (see Table }\ref{tab:gender-by-category}\changess{) extend to higher values.}}
+%	\label{fig:histogram-female}
+%\end{figure}
+%
+%% Histogram with the distribution of female pronouns for each occupation category (but categories are grouped into %'super-categories')
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
+%	\caption{\changess{Coalescing the data from Table }\ref{tab:gender-by-category}\changess{ by grouping categories into %broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, %Education, Arts / Entertainment, etc.
+In this context it becomes even clearer that STEM fields concentrate at $X = 0$, %as discussed in Figure }\ref{fig:histogram-female}\changess{.}}
+%	\label{fig:histogram-female-grouped}
+%\end{figure}
+%
+%% Histogram with the distribution of male pronouns for each occupation category
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-male}
+%	\caption{\changess{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at %$X = 6$.
+One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}}
+%	\label{fig:histogram-male}
+%\end{figure}
+%
+%% Histogram with the distribution of male pronouns for each occupation category (but categories are grouped into %'super-categories')
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
+%	\caption{\changess{Coalescing the data from table }\ref{tab:gender-by-category}\changess{ by grouping categories into %broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, %Education, Arts / Entertainment, etc.
+In this context it becomes clearer that STEM fields concentrate mainly to the %right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure }\ref{fig:histogram-female-grouped}\%changess{).}}
+%	\label{fig:histogram-male-grouped}
+%\end{figure}
+%
+%% Histogram with the distribution of neutral pronouns for each occupation category
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-neutral}
+%	\caption{\changess{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the %rarity even of female pronouns (see Figure }\ref{fig:histogram-female}\changess{).
+Once again, STEM fields are %predominantly concentrated on $X = 0$.}}
+%	\label{fig:histogram-neutral}
+%\end{figure}
+%
+%% Histogram with the distribution of neutral pronouns for each occupation category (but categories are grouped into %'super-categories')
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
+%	\caption{\changess{Coalescing the data from Table }\ref{tab:gender-by-category}\changess{ by grouping categories into %broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, %Education, Arts / Entertainment, etc.
+Once again, STEM fields are predominantly concentrated at $X = 0$.}}
+%	\label{fig:histogram-neutral-grouped}
+%\end{figure}
+%
+%\changess{The bar plots in Figure }\ref{fig:gender-by-category}\changess{ help us visualize how much of the distribution of %each occupation category is composed of female, male and gender-neutral pronouns.
+In this context, STEM fields, which show a %predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female %pronouns.}
+%
+%% Barplot with % of female, male and neutral translated pronouns for each occupation category
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
+%	\caption{\changess{Bar plots show how much of the distribution of translated gender pronouns for each occupation %category (grouped as in Table }\ref{tab:gender-by-category-grouped}\changess{) is composed of female, male and neutral %terms.
+STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger %proportion of female and neutral pronouns.
+Note that in general the bars do not add up to $100\%$, as there is a fair %amount of translated sentences for which we cannot obtain a gender pronoun.}}
+%	\label{fig:gender-by-category}
+%\end{figure}
 
-% Histogram with the distribution of female pronouns for each occupation category
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female}
-	\caption{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
-STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table \ref{tab:gender-by-category}) extend to higher values.}
-	\label{fig:histogram-female}
-\end{figure}
-
-% Histogram with the distribution of female pronouns for each occupation category (but categories are grouped into 'super-categories')
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure \ref{fig:histogram-female}.}
-	\label{fig:histogram-female-grouped}
-\end{figure}
-
-% Histogram with the distribution of male pronouns for each occupation category
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male}
-	\caption{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$. One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}
-	\label{fig:histogram-male}
-\end{figure}
-
-% Histogram with the distribution of male pronouns for each occupation category (but categories are grouped into 'super-categories')
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure \ref{fig:histogram-female-grouped}).}
-	\label{fig:histogram-male-grouped}
-\end{figure}
-
-% Histogram with the distribution of neutral pronouns for each occupation category
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral}
-	\caption{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure \ref{fig:histogram-female}).
-Once again, STEM fields are predominantly concentrated on $X = 0$.}
-	\label{fig:histogram-neutral}
-\end{figure}
-
-% Histogram with the distribution of neutral pronouns for each occupation category (but categories are grouped into 'super-categories')
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-Once again, STEM fields are predominantly concentrated at $X = 0$.}
-	\label{fig:histogram-neutral-grouped}
-\end{figure}
-
-The bar plots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
-In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.
-
-% Barplot with % of female, male and neutral translated pronouns for each occupation category
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
-	\caption{Bar plots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
-STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
-Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
-	\label{fig:gender-by-category}
-\end{figure}
-
+% PEDRO: TODO highlight section title
 \section{Distribution of translated gender pronouns per language}
 
-We have taken the care of experimenting with a fair amount of different gender neutral languages.
-Because of that, another sensible way of coalescing our data is by language groups, as shown in Table \ref{tab:gender-by-language}.
+\changess{We have taken the care of experimenting with a fair amount of different gender neutral languages.
+Because of that, another sensible way of coalescing our data is by language groups, as shown in Table }\ref{tab:gender-by-language}\changess{.
 This can help us visualize the effect of different cultures in the genesis -- or lack thereof -- of gender bias.
-Nevertheless, the barplots in Figure \ref{fig:gender-by-language} are perhaps most useful to identifying the difficulty of extracting a gender pronoun when translating from certain languages.
-Japanese and Basque are good examples of this difficulty, although the quality of Turkish, Chinese and Yoruba translations are also compromised.
+Nevertheless, the barplots in Figure }\ref{fig:gender-by-language}\changess{ are perhaps most useful to identifying the difficulty of extracting a gender pronoun when translating from certain languages.
+Japanese and Basque are good examples of this difficulty, although the quality of Turkish, Chinese and Yoruba translations are also compromised.}
 
-% Table with our gender bias results grouped by language
-\begin{table}[H]
-\small{
-	\centering
-	\begin{tabular}{|c|c|c|c|}
-	\hline
-	Language 	& Female ($\%$) 	& Male ($\%$)		& Neutral ($\%$)	\\ \hline
-	\hline
-	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
-	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
-	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
-	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
-	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
-	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
-	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
-	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
-	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
-	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
-	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline
-	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each language, averaged over all occupations detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
-	\label{tab:gender-by-language}
-	}
-\end{table}
+%% Table with our gender bias results grouped by language
+%\begin{table}[H]
+%\small{
+%	\centering
+%	\begin{tabular}{|c|c|c|c|}
+%	\hline
+%	Language 	& Female ($\%$) 	& Male ($\%$)		& Neutral ($\%$)	\\ \hline
+%	\hline
+%	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
+%	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
+%	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
+%	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
+%	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
+%	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
+%	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
+%	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
+%	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
+%	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
+%	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline
+%	\end{tabular}
+%	\caption{\changess{Percentage} of female, male and neutral gender pronouns \changess{obtained for each language, %averaged over all occupations detailed in Table }\ref{tab:gender-neutral-languages}\changess{.
+Note that columns do not %in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender %pronoun.}}
+%	\label{tab:gender-by-language}
+%	}
+%\end{table}
 
@@
 	\includegraphics[width=10cm]{pictures/gender-by-language}
-	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
+	\caption{\changess{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
 Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
-Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}
+Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}}
 	\label{fig:gender-by-language}
@@
 
-Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
-Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.
+\changess{Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
+Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.}
 
-Once again the data points towards male defaults, but some variation can be observed throughout different adjectives.
-Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).
+\changess{Once again the data points towards male defaults, but some variation can be observed throughout different adjectives.
+Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).}
 
 % Table with our gender bias results for each adjective
-\begin{table}[H]
-\small{
-	\centering
-	\begin{tabular}{|c|c|c|c|c|}
-	\hline
-	Adjective 	& Female 	& Male 	& Neutral 	& Total 	\\ \hline
-	\hline
-	Happy		& $4$		& $3$	& $2$		& $11$		\\ \hline
-	Wrong		& $0$		& $6$	& $4$		& $11$		\\ \hline
-	Smart		& $2$		& $4$	& $3$		& $11$		\\ \hline
-	Ashamed		& $3$		& $5$	& $1$		& $11$		\\ \hline
-	Envious		& $2$		& $5$	& $1$		& $11$		\\ \hline
-	Modest		& $1$		& $5$	& $3$		& $11$		\\ \hline
-	Innocent	& $1$		& $7$	& $1$		& $11$		\\ \hline
-	Sad			& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Afraid		& $1$		& $5$	& $0$		& $11$		\\ \hline
-	Dumb		& $1$		& $5$	& $2$		& $11$		\\ \hline
-	Kind		& $3$		& $3$	& $1$		& $11$		\\ \hline
-	Loving		& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Arrogant	& $0$		& $8$	& $1$		& $11$		\\ \hline
-	Helpless	& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Right		& $0$		& $6$	& $3$		& $11$		\\ \hline
-	Brave		& $2$		& $6$	& $1$		& $11$		\\ \hline
-	Proud		& $1$		& $7$	& $1$		& $11$		\\ \hline
-	Cruel		& $1$		& $7$	& $2$		& $11$		\\ \hline
-	Hateful		& $1$		& $4$	& $3$		& $11$		\\ \hline
-	Guilty		& $0$		& $8$	& $0$		& $11$		\\ \hline
-	Shy			& $5$		& $2$	& $2$		& $11$		\\ \hline
-	Attractive	& $4$		& $1$	& $4$		& $11$		\\ \hline \hline
-	Total 		& $38$		& $112$	& $41$		& -			\\ \hline 
-	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
-	\label{tab:gender-by-adjective}
-	}
-\end{table}
+%\begin{table}[H]
+%\small{
+%	\centering
+%	\begin{tabular}{|c|c|c|c|c|}
+%	\hline
+%	Adjective 	& Female 	& Male 	& Neutral 	& Total 	\\ \hline
+%	\hline
+%	Happy		& $4$		& $3$	& $2$		& $11$		\\ \hline
+%	Wrong		& $0$		& $6$	& $4$		& $11$		\\ \hline
+%	Smart		& $2$		& $4$	& $3$		& $11$		\\ \hline
+%	Ashamed		& $3$		& $5$	& $1$		& $11$		\\ \hline
+%	Envious		& $2$		& $5$	& $1$		& $11$		\\ \hline
+%	Modest		& $1$		& $5$	& $3$		& $11$		\\ \hline
+%	Innocent	& $1$		& $7$	& $1$		& $11$		\\ \hline
+%	Sad			& $2$		& $5$	& $2$		& $11$		\\ \hline
+%	Afraid		& $1$		& $5$	& $0$		& $11$		\\ \hline
+%	Dumb		& $1$		& $5$	& $2$		& $11$		\\ \hline
+%	Kind		& $3$		& $3$	& $1$		& $11$		\\ \hline
+%	Loving		& $2$		& $5$	& $2$		& $11$		\\ \hline
+%	Arrogant	& $0$		& $8$	& $1$		& $11$		\\ \hline
+%	Helpless	& $2$		& $5$	& $2$		& $11$		\\ \hline
+%	Right		& $0$		& $6$	& $3$		& $11$		\\ \hline
+%	Brave		& $2$		& $6$	& $1$		& $11$		\\ \hline
+%	Proud		& $1$		& $7$	& $1$		& $11$		\\ \hline
+%	Cruel		& $1$		& $7$	& $2$		& $11$		\\ \hline
+%	Hateful		& $1$		& $4$	& $3$		& $11$		\\ \hline
+%	Guilty		& $0$		& $8$	& $0$		& $11$		\\ \hline
+%	Shy			& $5$		& $2$	& $2$		& $11$		\\ \hline
+%	Attractive	& $4$		& $1$	& $4$		& $11$		\\ \hline \hline
+%	Total 		& $38$		& $112$	& $41$		& -			\\ \hline 
+%	\end{tabular}
+%	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
+%	\label{tab:gender-by-adjective}
+%	}
+%\end{table}
 
@@
 
-A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
-We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects the demographics of male-dominated fields (male-dominated fields can be considered those that have less than 25\% of women participation\citep{WB2014}, according to the U.S.
+\changess{A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
+We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects the demographics of male-dominated fields (male-dominated fields can be considered those that have less than 25\% of women participation}\citep{WB2014}\changess{, according to the U.S.
 Department of Labor Women's Bureau).
-In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.
+In this context, the argument in favor of a critical revision of statistic translation algorithms weakens considerably, and possibly shifts the blame away from these tools.}
 
-The U.S.
-Bureau of Labor Statistics data summarized in Table \ref{tab:occupations} contains statistics about the percentage of women participation in each occupation category.
+\changess{The U.S.
+Bureau of Labor Statistics data summarized in Table }\ref{tab:occupations}\changess{ contains statistics about the percentage of women participation in each occupation category.
 These data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile.
-We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure \ref{fig:histogram-compare-gt-real}.
+We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure }\ref{fig:histogram-compare-gt-real}\changess{.
 The data shows us that Google Translate outputs fail to follow the real-world distribution of female workers across a comprehensive set of job positions.
 The distribution of translated female pronouns is consistently inversely distributed, with female pronouns accumulating in the first 12-quantile.
-By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.
+By contrast, BLS data shows that female participation peaks in the fourth 12-quantile and remains significant throughout the next ones.}
 
@@
 	\includegraphics[width=10cm]{pictures/histogram-compare-gt-real}
-	\caption{Women participation ($\%$) data obtained from the U.S.
+	\caption{\changess{Women participation ($\%$) data obtained from the U.S.
 Bureau of Labor Statistics allows us to assess whether the Google Translate bias towards male defaults is at least to some extent explained by small frequencies of female workers in some job positions.
 Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles.
-To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral pronouns).}
+To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral pronouns).}}
 	\label{fig:histogram-compare-gt-real}
@@
 
-Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
+\changess{Running a t-test on the percentage of female translations over all translations for every occupation against the percentage of female participation in an occupation (surrogating with the LBS class value for the profession, in case it was missing), we see a mean of $11.11\%$ gender participation in the translations and a mean of $35.99\%$ in the LBS report.
 The variance reported for the translation results is also lower, at $\approx 0.01451$ in contrast with the report's $\approx 0.06668$.
 Altogether, the Pearson correlation between the two values is $\approx 0.30356$, with the p-values for one and two tail both being lower than the alpha value chosen ($\alpha = 0.05$).
-This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.
+This result enforces the fact that the translation exhibits a tendency towards male defaults, even when the female participation in that job is higher than expected.}
 
@@
 
-In this paper, we have provided preliminary evidence that statistical translation tools such as Google Translate can exhibit gender biases and a strong tendency towards male defaults.
-Although implicit, these biases possibly stem from the real world data which is used to train them, and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace.
-In this paper, we suggest that and test the hypothesis that statistical translation tools can be probed to yield insights about stereotypical gender roles in our society -- or at least in their training data.
-By translating professional-related sentences such as ``He/She is an engineer'' from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation outputs.
+In this paper, we have provided \changess{preliminary} evidence that statistical translation tools such as Google Translate \changess{can} exhibit gender biases and a strong tendency towards male defaults.
+\changess{Although implicit, these biases possibly} stem from the real world data which is used to train them, \changess{and in this context possibly provide a window into the way our society talks (and writes) about women in the workplace.
+In this paper}, we suggest that \changess{and test} the hypothesis that \changess{statistical translation} tools can be probed to yield insights about stereotypical gender roles in our society \changess{-- or at least in their training data.} By translating \changess{professional-related sentences such as} ``He/She is an engineer'' from gender neutral languages such as Hungarian and Chinese into  English, we were able to collect statistics bout the asymmetry between female and male pronominal genders in the translation \changess{outputs.
 Our results show that male defaults are not only prominent but exaggerated in fields suggested to be troubled with gender stereotypes, such as STEM (Science, Technology, Engineering and Mathematics) jobs.
-And because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages (e.g.
-Chinese $\rightarrow$ English $\rightarrow$ Portuguese) \citep{LanguageSupportNMT2018,boitet2010mt}, our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.
+And} because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages (e.g.
+Chinese $\rightarrow$ English $\rightarrow$ Portuguese) \citep{LanguageSupportNMT2018,boitet2010mt}, our findings possibly extend to translations between gender neutral languages and non-gender neutral languages (apart from English) in general, although we have not tested this hypothesis.\changess{    }
 
-Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
-Adjectives such as \emph{Shy} and \emph{Attractive are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
+\changess{Although not conclusive, our results seem to suggest that this phenomenon extends beyond the scope of the workplace, with the proportion of female pronouns varying significantly according to adjectives used to describe a person.
+Adjectives such as \emph{Shy} and \emph{Attractive} are predominantly translated with female pronouns, while \emph{Guilty} and \emph{Cruel} are almost exclusively translated with male pronouns.
 Different languages also seemingly have a significant impact in machine gender bias, with Hungarian exhibiting a better equilibrium between male and female pronouns than for example Chinese.
@@
 
-To solidify our results, we ran our pronominal gender translation statistics against the U.S.
+\changess{To solidify our results, we ran our pronominal gender translation statistics against the U.S.
 Bureau of Labor Statistics data on the frequency of women participation for each job position.
@@
 That is to say that even if we do not expect a 50:50 distribution of translated gender pronouns, Google Translate exhibits male defaults in a greater frequency that job occupation data alone would suggest.
-The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.
+The prominence of male defaults in Google Translate is therefore to the best of our knowledge yet lacking a clear justification.}
 
 We think this work shed new light on a  pressing ethical difficulty arising from modern statistical machine translation, and hope that it will lead to discussions about the role of AI engineers on minimizing potential harmful effects of the current concerns about machine bias.
-We are optimistic that unbiased results can be obtained with relatively little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.
+\changess{We are optimistic that unbiased results can be obtained with relatively little effort and marginal cost to the performance of current methods, to which current \emph{debiasing} algorithms in the scientific literature are a testament.}
 
+\ifx
 \begin{thebibliography}{xx}
@@
 \end{thebibliography}
-
-\section{Data Availability}
-All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
-They are not linked to external sites to avoid authors identification.
-
-\section{Competing interests}
-The authors declare that they have no competing financial interests.
+\fi
 

### 2018-04-18 Paper/main.tex
Changes to paper, gender-bias.py and new results for job genders in Results/

@@
 \begin{small}
-	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
+	\begin{tabular}{|c|m{1.5cm}|m{1.5cm}|m{2.2cm}|m{1.5cm}|}
 	\hline
-	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \hline \hline
-	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
-									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \hline
-	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
-									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \hline
-	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
+	Language Family 				& Language & Supported 	& Phrases have male/female markers & Tested \\ \hline \hline
+	\multirow{2}{*}{Austronesian} 	& Malay 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Tagalog 				& \xmark	 	& - 			& \xmark		\\ \hline
+	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Finnish 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Hungarian 			& \checkmark 	& \xmark 		& \checkmark	\\ \hline
+	\multirow{4}{*}{Indo-European} 	& Armenian 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Bengali				& \checkmark 	& O 			& \checkmark	\\
 									& English 				& \checkmark 	& \checkmark 	& \xmark 		\\
-									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \hline
-	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
-									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
-									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \hline
-	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Korean 				& O 						& \checkmark 	& \xmark		\\
-									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
-									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
-									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
-									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
-									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \hline
+									& {\color{red}Persian}	& \checkmark 	& \xmark 		& \checkmark	\\ \hline
+	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 	 	& - 			& \xmark 		\\
+									& {\color{red}Nepali}	& \checkmark 	& O 	 		& \checkmark	\\
+									& Oriya 				& \xmark 	 	& - 			& \xmark 		\\ \hline
+	\multirow{10}{*}{} 				& Japanese 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& {\color{red}Korean}	& \checkmark 	& \checkmark	& \xmark		\\
+									& Turkish 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Yoruba 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Basque 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Swahili 				& \checkmark 	& \xmark 		& \checkmark	\\
+									& Chinese 				& \checkmark 	& O 			& \checkmark	\\
+									& Cantonese 			& \xmark	 	& - 			& \xmark		\\
+									& Pipil 				& \xmark	 	& - 			& \xmark		\\
+									& Quechuan 				& \xmark	 	& - 			& \xmark		\\ \hline
 	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 	& \checkmark 	& \xmark		\\
-									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
-									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
-									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
+									& Ido 					& \xmark 	 	& - 			& \xmark 		\\ \cline{2-2}
+									& Lingua Franca Nova 	& \xmark 	 	& - 			& \xmark		\\ \cline{2-2}
+									& Interlingua 			& \xmark 	 	& - 			& \xmark		\\ \hline
 	\end{tabular}
     \end{small}
-	\caption{Selected gender neutral languages obtained from WALS and other sources.
-     Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional).
-For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded.
-Languages lacking support from Google Translate have been discarded.}
+	\caption{\changess{Selected gender neutral languages obtained from WALS and other sources.
+Languages are grouped according to language families and classified according to whether enforce any kind of mandatory gender (male/female) demarcation on simple phrases (\checkmark: yes, \xmark:~no, O: it is optional).
+For the purposes of this work, we have decided to work only with languages lacking such demarcation and as such Koran and Esperanto have been discarded.
+Languages lacking support from Google Translate have been discarded.
+Languages colored in red have been omitted in other reasons.
+See Section} \ref{ssec:rationale_for_language_exceptions} \changess{for further explanation.}}
 	\label{tab:gender-neutral-languages}
 \end{table}
+
+\subsection{Rationale for language exceptions}
+\label{ssec:rationale_for_language_exceptions}
+
+\changess{
+While it is possible to construct gender neutral sentences in two of the languages omitted in our experiments (namely Korean and Nepali), we have chosen to omit them for the following reasons:}
+
+\begin{enumerate}
+\item \changess{We faced technical difficulties to form templates and automatically translate sentences with the right-to-left, top-to-bottom nature of the Persian script, and as such have decided not to include it in our experiments.}
+\item \changess{Due to Nepali having a rather complex grammar, with possible male/female gender demarcations on the phrases and due to none of the authors being fluent or able to reach someone fluent in the language, we were not confident enough in our ability to produce the required templates.
+Bengali was almost discarded under the same rationale, but we have decided to keep it because our sentence template for Bengali has a simple grammatical structure which does not require any kind of inflection.}
+\item \changess{While technically, one can construct gender neutral phrases in Korean by omitting the gender pronoun, in fact, this is default procedure.
+However, the expressiveness of this omission depends on the context of the sentence being clear, which is not possible in the way we frame phrases.}
+\end{enumerate}
 
@@
 	\centering
-	\begin{tabular}{|m{3.5cm}|c|c|c|}
+	\begin{tabular}{|m{3.5cm}|m{2.0cm}|m{2.5cm}|m{2.5cm}|}
 	\hline	
@@
 	Finnish		& hän on $\langle occupation \rangle$ 									\\ \hline
-	Hungarian	& ő $\langle occupation \rangle$ 			\\ \hline
+	Hungarian	& ő egy $\langle occupation \rangle$ 									\\ \hline
 	Armenian	& na $\langle occupation \rangle$ e 									\\ \hline
+	\multirow{6}{*}{Bengali}	& Ē ē ēkajana $\langle occupation \rangle$				\\
+								& Yini yini ēkajana $\langle occupation \rangle$		\\
+								& Ō ō ēkajana $\langle occupation \rangle$				\\
+								& Uni uni ēkajana $\langle occupation \rangle$			\\
+								& Sē sē ēkajana $\langle occupation \rangle$			\\
+								& Tini tini ēkajana $\langle occupation \rangle$		\\ \hline
 	Japanese	& $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
@@
 	Smart		& Dumb		 & Proud 	\\ \hline
-	Ashamed		& Kind		& Cruel 	\\ \hline
-	Envious		& Loving	& Hateful 	\\ \hline
-	Modest 		& Arrogant	& Guilty	\\ \hline
-	Innocent	& Helpless	& Shy		\\ \hline
+	STRONG		& POLITE	 & Cruel 	\\ \hline
+	DESIRABLE	& Loving	 & SYMPHATETIC 	\\ \hline
+	Modest 		& SUCCESSFUL & Guilty	\\ \hline
+	Innocent	& MATURE	 & Shy		\\ \hline
 	\end{tabular}
@@
 
-Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
-Bureau of Labor Statistics, we have also selected a small and for all purposes not representative subset of adjectives, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.
+We queried the 1000 most frequently used adjectives in English, as classified in the COCA corpus [https://corpus.byu.edu/coca/], but since not all of them were readily appliable to the sentence template we used, we filtered the N adjectives that would fit the templates and made sense for describing a human being.
+The list of adjectives extracted from the corpus will be present on the github repository after publishing along with the rationale for the exclusion of each word.
+
+\changess{Apart from occupations, which we have exhaustively examined by collecting labor data from the U.S.
+Bureau of Labor Statistics, we have also selected a small subset of adjectives from the Corpus of Contemporary American English (COCA)} \url{https://corpus.byu.edu/coca/}\changess{, in an attempt to provide preliminary evidence that the phenomenon of gender bias may extend beyond the professional context examined in this paper.
+Because a large number of adjectives are not appliable to human subjects, we manually curated a reasonable subset of such words.
+The template used for adjectives is identical to that used for occupations, but is provided again for reference in Table} \ref{tab:templates-adjectives}.
 
@@
 Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).
+
+% List of sentence templates for assessing gender bias in the translation from source language to English
+\begin{table}[H]
+\centering
+\small{
+	\begin{tabular}{|c|c|}
+	\hline
+	Language 	& Sentence template 													\\ \hline \hline
+	Malay		& dia adalah $\langle adjective \rangle$ 								\\ \hline
+	Estonian	& ta on $\langle adjective \rangle$ 									\\ \hline
+	Finnish		& hän on $\langle adjective \rangle$ 									\\ \hline
+	Hungarian	& ő egy $\langle adjective \rangle$ 									\\ \hline
+	Armenian	& na $\langle adjective \rangle$ e 									\\ \hline
+	\multirow{6}{*}{Bengali}	& Ē ē ēkajana $\langle adjective \rangle$				\\
+								& Yini yini ēkajana $\langle adjective \rangle$		\\
+								& Ō ō ēkajana $\langle adjective \rangle$				\\
+								& Uni uni ēkajana $\langle adjective \rangle$			\\
+								& Sē sē ēkajana $\langle adjective \rangle$			\\
+								& Tini tini ēkajana $\langle adjective \rangle$		\\ \hline
+	Japanese	& $\langle adjective \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
+	Turkish		& o bir $\langle adjective \rangle$ 									\\ \hline
+	Yoruba		& o jẹ $\langle adjective \rangle$ 									\\ \hline
+	Basque		& $\langle adjective \rangle$ da 										\\ \hline
+	Swahili		& yeye ni $\langle adjective \rangle$ 									\\ \hline
+	Chinese		& ta $\langle adjective \rangle$ 										\\ \hline
+	\end{tabular}
+	}
+	\caption{Templates used to infer gender biases in the translation to the English language.}
+	\label{tab:templates-adjectives}
+\end{table}
 
@@
 	Smart			& $2$		& $4$	& $3$		& $11$		\\ \hline
-	Ashamed		& $3$		& $5$	& $1$		& $11$		\\ \hline
-	Envious		& $2$		& $5$	& $1$		& $11$		\\ \hline
 	Modest			& $1$		& $5$	& $3$		& $11$		\\ \hline
@@
 	Dumb			& $1$		& $5$	& $2$		& $11$		\\ \hline
-	Kind		& $3$		& $3$	& $1$		& $11$		\\ \hline
 	Loving			& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Arrogant	& $0$		& $8$	& $1$		& $11$		\\ \hline
-	Helpless	& $2$		& $5$	& $2$		& $11$		\\ \hline
 	Right			& $0$		& $6$	& $3$		& $11$		\\ \hline
@@
 	Cruel			& $1$		& $7$	& $2$		& $11$		\\ \hline
-	Hateful		& $1$		& $4$	& $3$		& $11$		\\ \hline
 	Guilty			& $0$		& $8$	& $0$		& $11$		\\ \hline
 	Shy				& $5$		& $2$	& $2$		& $11$		\\ \hline
-	Attractive	& $4$		& $1$	& $4$		& $11$		\\ \hline \hline
+	Attractive		& $4$		& $1$	& $4$		& $11$		\\ \hline
+	STRONG			&			&		&			&			\\ \hline
+	DESIRABLE		&			&		&			&			\\ \hline
+	SYMPHATETIC		&			&		&			&			\\ \hline
+	POLITE			&			&		&			&			\\ \hline
+	SUCCESSFUL		&			&		&			&			\\ \hline
+	MATURE			&			&		&			&			\\ \hline \hline
 	Total 			& $38$		& $112$	& $41$		& -			\\ \hline 
 	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
+	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.
+Adjectives marked with an asterisk did not make up to the top $1000$ most frequently used adjectives in the COCA corpus but we have decided to include them because }
 	\label{tab:gender-by-adjective}
@@
 Bureau of Labor Statistics data summarized in Table \ref{tab:occupations} contains statistics about the percentage of women participation in each occupation category.
-These data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile.
+This data is also available for each individual occupation, which allows us to compute the frequency of women participation for each 12-quantile.
 We carried the same computation in the context of frequencies of translated female pronouns, and the resulting histograms are plotted side-by-side in Figure \ref{fig:histogram-compare-gt-real}.

### 2018-04-18 gender-bias.py

@@ -18,6 +18,9 @@ with open('languages.csv', 'r') as f:
     		languages.append(language)
 #end with
 
+# Define list of discarded languages
+discarded_languages = ['Tagalog','English','Maithili','Oriya','Korean','Cantonese','Pipil','Quechuan','Esperanto','Ido','Lingua Franca Nova','Interlingua']
+
 # Read job occupations list into table
 table = np.array(list(csv.reader(open('jobs/bureau_of_labor_statistics_profession_list_gender_filtered_expanded.tsv','r'), delimiter='\t')))
 
@@ -79,7 +82,7 @@ if False:
 translated_occupations = list(csv.reader(open('Results/jobs-translations.tsv','r'), delimiter='\t'))
 
 # Define function to obtain the translated gender of an occupation in a given language (through Google Translate)
-def get_gender(occupation,language):
+def get_gender(occupation,language,case=None):
 
 	translation = ''
 
@@ -90,11 +93,26 @@ def get_gender(occupation,language):
 	elif(language == 'Finnish'):
 		translation = translator.translate('hän on %s' % occupation, src=language, dest='en').text
 	elif(language == 'Hungarian'):
-		translation = translator.translate('ő %s' % occupation, src=language, dest='en').text
+		translation = translator.translate('ő egy %s' % occupation, src=language, dest='en').text
 	elif(language == 'Armenian'):
 		translation = translator.translate('նա %s է' % occupation, src=language, dest='en').text
+	
+	elif(language == 'Bengali'):
+		if(case == 'HF'):
+			translation = translator.translate('এ এ একজন %s' % occupation, src=language, dest='en').text
+		elif(case == 'HP'):
+			translation = translator.translate('যিনি যিনি একজন %s' % occupation, src=language, dest='en').text
+		elif(case == 'TF'):
+			translation = translator.translate('ও ও একজন %s' % occupation, src=language, dest='en').text
+		elif(case == 'TP'):
+			translation = translator.translate('উনি উনি একজন %s' % occupation, src=language, dest='en').text
+		elif(case == 'EF'):
+			translation = translator.translate('সে সে একজন %s' % occupation, src=language, dest='en').text
+		elif(case == 'EP'):
+			translation = translator.translate('তিনি তিনি একজন %s' % occupation, src=language, dest='en').text
+
 	elif(language == 'Japanese'):
[…]

### 2018-04-18 gender-bias.py
Small change to chinese occupation template

@@ -122,7 +122,7 @@ def get_gender(occupation,language,case=None):
 	elif(language == 'Swahili'):
 		translation = translator.translate('yeye ni %s' % occupation, src=language, dest='en').text
 	elif(language == 'Chinese'):
-		translation = translator.translate('ta %s' % p.get_pinyin(occupation,''), src='zh-cn', dest='en').text
+		translation = translator.translate('ta shi %s' % p.get_pinyin(occupation,''), src='zh-cn', dest='en').text
 
 	translation = translation.lower()
 

### 2018-04-19 gender-bias.py
Fixed some wrong templates and added a comment at the end with the templates for adjectives.

@@ -99,17 +99,17 @@ def get_gender(occupation,language,case=None):
 	
 	elif(language == 'Bengali'):
 		if(case == 'HF'):
-			translation = translator.translate('এ এ একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('এ একজন %s' % occupation, src=language, dest='en').text
 		elif(case == 'HP'):
-			translation = translator.translate('যিনি যিনি একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('যিনি একজন %s' % occupation, src=language, dest='en').text
 		elif(case == 'TF'):
-			translation = translator.translate('ও ও একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('ও একজন %s' % occupation, src=language, dest='en').text
 		elif(case == 'TP'):
-			translation = translator.translate('উনি উনি একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('উনি একজন %s' % occupation, src=language, dest='en').text
 		elif(case == 'EF'):
-			translation = translator.translate('সে সে একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('সে একজন %s' % occupation, src=language, dest='en').text
 		elif(case == 'EP'):
-			translation = translator.translate('তিনি তিনি একজন %s' % occupation, src=language, dest='en').text
+			translation = translator.translate('তিনি একজন %s' % occupation, src=language, dest='en').text
 
 	elif(language == 'Japanese'):
 		translation = translator.translate('あの人は%sです' % occupation, src=language, dest='en').text
@@ -118,7 +118,7 @@ def get_gender(occupation,language,case=None):
 	elif(language == 'Yoruba'):
 		translation = translator.translate('o jẹ %s' % occupation, src=language, dest='en').text
 	elif(language == 'Basque'):
-		translation = translator.translate('%s da' % occupation, src=language, dest='en').text
+		translation = translator.translate('%s bat da' % occupation, src=language, dest='en').text
 	elif(language == 'Swahili'):
 		translation = translator.translate('yeye ni %s' % occupation, src=language, dest='en').text
 	elif(language == 'Chinese'):
@@ -185,3 +185,49 @@ with open('Results/job-genders.tsv','w') as output:
 	#end for
 
 #end with
+
+
+
+"""
+For adjectives use the following templates:
+
+	if(language == 'Malay'):
[…]

### 2018-04-19 gender-bias.py
minor changes

@@ -126,7 +126,7 @@ def get_gender(occupation,language,case=None):
 
 	translation = translation.lower()
 
-	if(translation[0:4].find("she") != -1 or translation[0:4].find("she's") != -1 or translation[0:4].find("her") != -1 or translation[0:10].find("that woman") != -1):
+	if(translation[0:3].find("she") != -1 or translation[0:4].find("she's") != -1 or translation[0:4].find("her") != -1 or translation[0:10].find("that woman") != -1):
 		return 'Female'
 	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1 or translation[0:4].find("his") != -1 or translation[0:8].find("that man") != -1):
 		return 'Male'

### 2018-04-19 gender-bias.py
Changed syntax for checking for markers

@@ -125,13 +125,19 @@ def get_gender(occupation,language,case=None):
 		translation = translator.translate('ta shi %s' % p.get_pinyin(occupation,''), src='zh-cn', dest='en').text
 
 	translation = translation.lower()
+	
+	female_markers = ["she", "she's", "her"]
+	male_markers = ["he", "he's", "his"]
+	neuter_markers = ["it","it's","its","they","they're","them","who","this","that"]
+	
+	has_any = lambda markers, translation: any( [ marker.lower() in translation.lower().split() for marker in markers ] )
 
-	if(translation[0:4].find("she") != -1 or translation[0:4].find("she's") != -1 or translation[0:4].find("her") != -1 or translation[0:10].find("that woman") != -1):
-		return 'Female'
-	elif(translation[0:4].find("he") != -1 or translation[0:4].find("he's") != -1 or translation[0:4].find("his") != -1 or translation[0:8].find("that man") != -1):
-		return 'Male'
-	elif(translation[0:4].find("it") != -1 or translation[0:4].find("it's") != -1 or translation[0:4].find("its") != -1 or translation[0:7].find("they") != -1 or translation[0:7].find("they're") != -1 or translation[0:4].find("them") != -1 or translation[0:3].find("who") != -1 or translation[0:4].find("this") != -1 or translation[0:4].find("that") != -1):
-		return 'Neutral'
+	if( has_any(female_markers, translation) or translation[0:10].find("that woman") != -1):
+		return 'Female' # Suggestion: (1,0,0)
+	elif( has_any(male_markers, translation) or translation[0:8].find("that man") != -1):
+		return 'Male' # Suggestion: (0,1,0)
+	elif( has_any(neuter_markers, translation) ):
+		return 'Neutral' # Suggestion: (0,0,1)
 	else:
 		return '?'
 #end def

### 2018-04-19 gender-bias.py
Updated gender-bias code

@@ -130,9 +130,9 @@ def get_gender(language, occupation=None, adjective=None, case=None):
 		#end if
 	elif adjective is not None:
 		if language == 'Bengali':
-			phrase = occupation_dict['Bengali-%s' % case] % occupation
+			phrase = adjective_dict['Bengali-%s' % case] % adjective
 		else:
-			phrase = occupation_dict[language] % occupation
+			phrase = adjective_dict[language] % adjective
 		#end if
 	else:
 		raise Exception("Neither and occupation nor an adjective has been provided")

### 2018-04-20 gender-bias.py
Added adjective sub-part of the experiment as a possible workflow in the python script. Now one has to send "occ" or "adj" as an argument to the script to choose which types of data collectino will be done.

@@ -1,4 +1,4 @@
-
+import sys
 from googletrans import Translator
 from googletrans import LANGUAGES
 from xpinyin import Pinyin
@@ -7,6 +7,10 @@ import numpy as np
 
 translator = Translator()
 p = Pinyin()
+
+do_occ = "occ" in sys.argv
+do_adj = "adj" in sys.argv
+
 
 # Get language list
 languages = []
@@ -21,23 +25,27 @@ with open('languages.csv', 'r') as f:
 # Define list of discarded languages
 discarded_languages = ['Nepali','Tagalog','English','Maithili','Oriya','Korean','Cantonese','Pipil','Quechuan','Esperanto','Ido','Lingua Franca Nova','Interlingua']
 
-# Read job occupations list into table
-table = np.array(list(csv.reader(open('jobs/bureau_of_labor_statistics_profession_list_gender_filtered_expanded.tsv','r'), delimiter='\t')))
+if do_occ:
+	# Read job occupations list into table
+	occupation_table = np.array(list(csv.reader(open('jobs/bureau_of_labor_statistics_profession_list_gender_filtered_expanded.tsv','r'), delimiter='\t')))
+	# Compute list of categories
+	categories = list(set(occupation_table[1:,-3]))
+	# Get dictionary of category -> jobs (jobs per category)
+	categories_dict = dict([ (category, occupation_table[occupation_table[:,-3] == category,0]) for category in categories ])
+#end if
 
-# Compute list of categories
-categories = list(set(table[1:,-3]))
-
-# Get dictionary of category -> jobs (jobs per category)
-categories_dict = dict([ (category, table[table[:,-3] == category,0]) for category in categories ])
+if do_adj:
+	adjectives = np.array( [ "happy", "sad", "right", "wrong", "afraid", "brave", "smart", "dumb", "proud", "strong", "polite", "cruel", "desirable", "loving", "sympathetic", "modest", "successful", "guilty", "innocent", "mature", "shy" ] ) # Could've read from file, but since the file is so small pasted it directly here.
+#end if
 
 """
-	Create a table with the translated version (provided by Google Translate) of each job,
+	Create a occupation_table with the translated version (provided by Google Translate) of each job,
 	in the following structure:
 		1. Each line corresponds to a single occupation
 		2. The first column is the occupation category ("Computer and mathematical occupations", "Healthcare practitioners and technical occupations", etc.)
[…]

### 2018-04-20 Paper/main.tex
Added paper
Major changes

@@
 \usepackage{pifont}
+\usepackage[caption=false]{subfig}
 
@@
 As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when ``nurse'' is replaced by ``engineer''.
-The same basic template can be ported to all other gender neutral languages, as depicted in Table \ref{tab:templates}.
+The same basic template can be ported to all other gender neutral languages, as depicted in Table \ref{tab:templates-occupations}.
 Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias.
@@
 
-There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}.
-To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias.
-Here, we resorted to using the Bureau of Labor Statistics' detailed occupations table \citep{BLS2017}, from the United States Department of Labor, for providing such a list.
-The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
+\changess{There is a prohibitively large class of nouns and adjectives that could in principle be substituted into our templates.
+To simplify our dataset, we have decided to focus our work on job positions -- which, we believe, are an interesting window into the nature of gender bias --, and were able to obtain a comprehensive list of professional occupations from the Bureau of Labor Statistics' detailed occupations table} \citep{BLS2017}\changess{, from the United States Department of Labor.} The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones.
 Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers.
@@
 For the occupations that had less than 50 thousand workers, and thus no data about the participation of women, we assumed that its women participation was that of its upper category.
-Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.
-All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
+\changess{Finally, as complementary evidence we have decided to include a small subset of 21 adjectives in our study.
+All adjectives were obtained from the top one thousand most frequent words in this category as featured in the Corpus of Contemporary American English (COCA)} \url{https://corpus.byu.edu/coca/}\changess{, but it was necessary to manually curate them because a substantial fraction of these adjectives cannot be applied to human subjects.
+Also because the sentiment associated with each adjective is not as easily accessible as for example the occupation category of each job position, we performed a manual selection of a subset of such words which we believe to be meaningful to this study.
+These words are presented in Table} \ref{tab:adjectives}.
+All code and data used to generate and{} compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request.
 They are not linked to external sites to avoid authors identification.
@@
 Languages are grouped according to language families and classified according to whether enforce any kind of mandatory gender (male/female) demarcation on simple phrases (\checkmark: yes, \xmark:~no, O: it is optional).
-For the purposes of this work, we have decided to work only with languages lacking such demarcation and as such Koran and Esperanto have been discarded.
+For the purposes of this work, we have decided to work only with languages lacking such demarcation such as Esperanto have been discarded.
 Languages lacking support from Google Translate have been discarded.
@@
 	Armenian	& na $\langle occupation \rangle$ e 									\\ \hline
-	\multirow{6}{*}{Bengali}	& Ē ē ēkajana $\langle occupation \rangle$				\\
-								& Yini yini ēkajana $\langle occupation \rangle$		\\
-								& Ō ō ēkajana $\langle occupation \rangle$				\\
-								& Uni uni ēkajana $\langle occupation \rangle$			\\
-								& Sē sē ēkajana $\langle occupation \rangle$			\\
-								& Tini tini ēkajana $\langle occupation \rangle$		\\ \hline
-	Japanese	& $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
+	\multirow{6}{*}{Bengali}	& Ē ēkajana $\langle occupation \rangle$				\\
+								& Yini ēkajana $\langle occupation \rangle$		\\
+								& Ō ēkajana $\langle occupation \rangle$				\\
+								& Uni ēkajana $\langle occupation \rangle$			\\
+								& Sē ēkajana $\langle occupation \rangle$			\\
+								& Tini ēkajana $\langle occupation \rangle$		\\ \hline
+	Japanese	& \begin{CJK*}{UTF8}{min}あの人は\end{CJK*} $\langle occupation \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
 	Turkish		& o bir $\langle occupation \rangle$ 									\\ \hline
 	Yoruba		& o jẹ $\langle occupation \rangle$ 									\\ \hline
-	Basque		& $\langle occupation \rangle$ da 										\\ \hline
+	Basque		& $\langle occupation \rangle$ bat da 										\\ \hline
 	Swahili		& yeye ni $\langle occupation \rangle$ 									\\ \hline
-	Chinese		& ta $\langle occupation \rangle$ 										\\ \hline
+	Chinese		& ta shi $\langle occupation \rangle$ 										\\ \hline
 	\end{tabular}
 	}
-	\caption{Templates used to infer gender biases in the translation to the English language.}
-	\label{tab:templates}
+	\caption{\changess{Templates used to infer gender biases in the translation of job occupations to the English language.}}
+	\label{tab:templates-occupations}
+\end{table}
+
+\begin{table}[H]
+\centering
+\small{
+	\begin{tabular}{|c|c|}
+	\hline
+	Language 	& Sentence template 													\\ \hline \hline
+	Malay		& dia $\langle adjective \rangle$ 								\\ \hline
+	Estonian	& ta on $\langle adjective \rangle$ 									\\ \hline
+	Finnish		& hän on $\langle adjective \rangle$ 									\\ \hline
+	Hungarian	& ő $\langle adjective \rangle$ 									\\ \hline
+	Armenian	& na $\langle adjective \rangle$ e 									\\ \hline
+	\multirow{6}{*}{Bengali}	& Ē $\langle adjective \rangle$				\\
+								& Yini $\langle adjective \rangle$		\\
+								& Ō $\langle adjective \rangle$				\\
+								& Uni $\langle adjective \rangle$			\\
+								& Sē $\langle adjective \rangle$			\\
+								& Tini $\langle adjective \rangle$		\\ \hline
+	Japanese	& \begin{CJK*}{UTF8}{min}あの人は\end{CJK*} $\langle adjective \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
+	Turkish		& o $\langle adjective \rangle$ 									\\ \hline
+	Yoruba		& o jẹ $\langle adjective \rangle$ 									\\ \hline
+	Basque		& $\langle adjective \rangle$ da 										\\ \hline
+	Swahili		& yeye ni $\langle adjective \rangle$ 									\\ \hline
+	Chinese		& ta hen $\langle adjective \rangle$ 										\\ \hline
+	\end{tabular}
+	}
+	\caption{\changess{Templates used to infer gender biases in the translation of adjectives to the English language.}}
+	\label{tab:templates-adjectives}
 \end{table}
@@
 	Smart		& Dumb		 & Proud 		\\ \hline
-	STRONG		& POLITE	 & Cruel 	\\ \hline
-	DESIRABLE	& Loving	 & SYMPHATETIC 	\\ \hline
-	Modest 		& SUCCESSFUL & Guilty	\\ \hline
-	Innocent	& MATURE	 & Shy		\\ \hline
+	Strong		& Polite	 & Cruel 		\\ \hline
+	Desirable	& Loving	 & Sympathetic 	\\ \hline
+	Modest 		& Successful & Guilty		\\ \hline
+	Innocent	& Mature	 & Shy			\\ \hline
 	\end{tabular}
 	}
-    \caption{Adjectives}
+    \caption{\changess{Curated list of 21 adjectives obtained from the top one thousand most frequent words in this category in the Corpus of Contemporary American English (COCA)} \url{https://corpus.byu.edu/coca/}.}
 	\label{tab:adjectives}
@@
 	\hline
-	Education / Training / Library 			& $23.737$ 		& $58.081$		& $3.535$		\\ \hline
-	Business / Financial 					& $6.039$		& $62.319$		& $8.937$		\\ \hline
-	Office / Administrative					& $9.962$		& $56.322$		& $9.451$		\\ \hline
-	Healthcare support						& $27.083$		& $50.000$		& $9.722$		\\ \hline
-	Management								& $11.594$		& $61.111$		& $4.831$		\\ \hline
-	Installation / Maintenance / Repair		& $3.297$		& $62.393$		& $14.164$		\\ \hline
-	Healthcare practitioners / technical	& $18.088$		& $61.499$		& $8.527$		\\ \hline
-	Community and social service			& $14.286$ 		& $60.317$		& $5.556$		\\ \hline
-	Sales and related						& $7.540$		& $64.683$		& $10.714$		\\ \hline
-	Production								& $15.32$		& $45.370$		& $15.825$		\\ \hline
-	Architecture and engineering			& $1.916$		& $75.862$		& $6.130$		\\ \hline
-	Life, physical, and social sciences		& $3.595$		& $80.065$		& $4.575$		\\ \hline
-	Transportation / material moving		& $7.778$		& $60.159$		& $14.286$		\\ \hline
-	Arts / Entertainment					& $7.808$		& $66.366$		& $5.405$		\\ \hline
-	Legal									& $9.524$ 		& $65.079$		& $6.349$		\\ \hline
-	Protective service						& $6.746$		& $71.032$		& $7.143$		\\ \hline
-	Food preparation / serving			& $14.379$ 		& $56.863$		& $12.418$		\\ \hline
-	Farming, fishing, and forestry			& $12.821$ 		& $53.846$		& $10.256$		\\ \hline
-	Computer and mathematical				& $3.472$		& $70.833$		& $5.556$		\\ \hline
-	Personal care and service				& $17.845$ 		& $52.862$		& $11.111$		\\ \hline
-	Construction and extraction				& $9.967$		& $54.739$		& $13.562$		\\ \hline
-	Cleaning / Maintenance					& $12.222$		& $57.778$		& $10.000$		\\ \hline
-	\hline
-	Total           						& $11.078$ 		& $57.398$ 		& $11.286$ 		\\ \hline
+	Office and administrative support & 11.015 & 58.812 & 16.954 \\ \hline
+	Architecture and engineering & 2.299 & 72.701 & 10.92 \\ \hline
+	Farming, fishing, and forestry & 12.179 & 62.179 & 14.744 \\ \hline
+	Management & 11.232 & 66.667 & 12.681 \\ \hline
+	Community and social service & 20.238 & 62.5 & 10.119 \\ \hline
+	Healthcare support & 25.0 & 43.75 & 17.188 \\ \hline
+	Sales and related & 8.929 & 62.202 & 16.964 \\ \hline
+	Installation, maintenance, and repair & 5.22 & 58.333 & 17.125 \\ \hline
+	Transportation and material moving & 8.81 & 62.976 & 17.5 \\ \hline
+	Legal & 11.905 & 72.619 & 10.714 \\ \hline
+	Business and financial operations & 7.065 & 67.935 & 15.58 \\ \hline
+	Life, physical, and social science & 5.882 & 73.284 & 10.049 \\ \hline
+	Arts, design, entertainment, sports, and media & 10.36 & 67.342 & 11.486 \\ \hline
+	Education, training, and library & 23.485 & 53.03 & 9.091 \\ \hline
+	Building and grounds cleaning and maintenance & 12.5 & 68.333 & 11.667 \\ \hline
+	Personal care and service & 18.939 & 49.747 & 18.434 \\ \hline
+	Healthcare practitioners and technical & 22.674 & 51.744 & 15.116 \\ \hline
+	Production & 14.331 & 51.199 & 18.245 \\ \hline
+	Computer and mathematical & 4.167 & 66.146 & 14.062 \\ \hline
+	Construction and extraction & 8.578 & 61.887 & 17.525 \\ \hline
+	Protective service & 8.631 & 65.179 & 12.5 \\ \hline
+	Food preparation and serving related & 21.078 & 58.333 & 17.647 \\ \hline \hline
+	Total & 11.76 & 58.93 & 15.939 \\ \hline
+
 	\end{tabular}
 	}
-    \caption{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+    \caption{\changess{Percentage of female, male and neutral gender pronouns obtained for each BLS occupation category, averaged over all occupations in said category and tested languages detailed in Table} \ref{tab:gender-neutral-languages}\changess{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-category}
@@
 	\hline
-	Education								& $23.737$				& $58.081$				& $3.535$			\\ \hline
-	Corporate								& $8.519$				& $62.407$				& $7.778$			\\ \hline
-	Service									& $8.730$				& $59.746$				& $11.619$			\\ \hline
-	Healthcare								& $20.527$				& $58.380$				& $8.851$			\\ \hline
-	Production								& $15.32$				& $45.370$				& $15.825$			\\ \hline
-	STEM									& $2.954$				& $76.653$				& $5.345$			\\ \hline
-	Arts / Entertainment					& $7.808$				& $66.366$				& $5.405$			\\ \hline
-	Legal									& $9.524$				& $65.079$				& $6.349$			\\ \hline
-	Farming / Fishing / Forestry			& $12.821$				& $53.846$				& $10.256$			\\ \hline
-	Construction / Extraction				& $9.967$				& $54.739$				& $13.562$			\\ \hline
-	\hline
-	Total           						& $11.078$ 				& $57.398$ 				& $11.286$ 			\\ \hline
+	Service & 10.5 & 59.548 & 16.476 \\ \hline
+	STEM & 4.219 & 71.624 & 11.181 \\ \hline
+	Farming / Fishing / Forestry & 12.179 & 62.179 & 14.744 \\ \hline
+	Corporate & 9.167 & 66.042 & 14.861 \\ \hline
+	Healthcare & 23.305 & 49.576 & 15.537 \\ \hline
+	Legal & 11.905 & 72.619 & 10.714 \\ \hline
+	Arts / Entertainment & 10.36 & 67.342 & 11.486 \\ \hline
+	Education & 23.485 & 53.03 & 9.091 \\ \hline
+	Production & 14.331 & 51.199 & 18.245 \\ \hline
+	Construction / Extraction & 8.578 & 61.887 & 17.525 \\ \hline \hline
+	Total & 11.76 & 58.93 & 15.939 \\ \hline
+
 	\end{tabular}
 	}
-    \caption{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation category, averaged over all occupations in said category and tested languages detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+    \caption{\changess{Percentage of female, male and neutral gender pronouns obtained for each of the merged occupation category, averaged over all occupations in said category and tested languages detailed in Table} \ref{tab:gender-neutral-languages}\changess{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-category-grouped}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female}
-	\caption{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
-STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table \ref{tab:gender-by-category}) extend to higher values.}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Female}
+	\caption{\changess{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
+STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table} \ref{tab:gender-by-category}\changess{) extend to higher values.}}
 	\label{fig:histogram-female}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-female-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure \ref{fig:histogram-female}.}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Female-grouped}
+	\caption{\changess{Coalescing the data from Table} \ref{tab:gender-by-category} \changess{by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes even clearer that STEM fields concentrate at $X = 0$, as discussed in Figure} \ref{fig:histogram-female}.}
 	\label{fig:histogram-female-grouped}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male}
-	\caption{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$. One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Male}
+	\caption{\changess{In contrast to female pronouns, male pronouns are seemingly skew normally distributed, with a peak at $X = 6$.
+One can see that STEM fields concentrate mainly to the right ($X \geq 6$).}}
 	\label{fig:histogram-male}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-male-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure \ref{fig:histogram-female-grouped}).}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Male-grouped}
+	\caption{\changess{Coalescing the data from Table} \ref{tab:gender-by-category} \changess{by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+In this context it becomes clearer that STEM fields concentrate mainly to the right ($X \geq 6$), in contrast to what we observe for female pronouns (see Figure} \ref{fig:histogram-female-grouped}).}
 	\label{fig:histogram-male-grouped}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral}
-	\caption{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure \ref{fig:histogram-female}).
-Once again, STEM fields are predominantly concentrated on $X = 0$.}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Neutral}
+	\caption{\changess{The scarcity of gender-neutral pronouns is manifest in their histogram, showing how they exceed the rarity even of female pronouns (see Figure} \ref{fig:histogram-female}\changess{).
+Once again, STEM fields are predominantly concentrated on $X = 0$.}}
 	\label{fig:histogram-neutral}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/histogram-neutral-grouped}
-	\caption{Coalescing the data from Table \ref{tab:gender-by-category} by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
-Once again, STEM fields are predominantly concentrated at $X = 0$.}
+	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Neutral-grouped}
+	\caption{\changess{Coalescing the data from Table} \ref{tab:gender-by-category} \changess{by grouping categories into broader groups helps us see how translated gender pronouns are distributed for some fields of interest such as STEM, Education, Arts / Entertainment, etc.
+Once again, STEM fields are predominantly concentrated at $X = 0$.}}
 	\label{fig:histogram-neutral-grouped}
+\end{figure}
+
+AAAAAAAA
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histograms/categories/gender-dodged-STEM}
+	\caption{TODO STEM}
+	\label{fig:histogram-dodged-STEM}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/histograms/categories/gender-dodged-Healthcare}
+	\caption{TODO Healthcare}
+	\label{fig:histogram-dodged-Healthcare}
+\end{figure}
+
+BBBBBVBBV
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Female}
+	\caption{TODO heatmap female}
+	\label{fig:heatmap-female}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Male}
+	\caption{TODO heatmap male}
+	\label{fig:heatmap-male}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Neutral}
+	\caption{TODO heatmap neutral}
+	\label{fig:heatmap-neutral}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/gender-by-category-grouped}
-	\caption{Bar plots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table \ref{tab:gender-by-category-grouped}) is composed of female, male and neutral terms.
+	\includegraphics[width=10cm]{pictures/barplot-gender-by-category-grouped}
+	\caption{\changess{Bar plots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table} \ref{tab:gender-by-category-grouped}\changess{) is composed of female, male and neutral terms.
 STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
-Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{fig:gender-by-category}
@@
 	\hline
-	Malay     	& $4.122$ 			& $95.780$  		& $0.000$			\\ \hline
-	Estonian  	& $18.351$			& $80.962$  		& $0.491$			\\ \hline
-	Finnish   	& $33.562$			& $66.143$  		& $0.000$			\\ \hline
-	Hungarian 	& $36.016$			& $63.297$  		& $0.393$			\\ \hline
-	Armenian  	& $9.814$			& $81.256$  		& $0.589$			\\ \hline
-	Japanese  	& $0.196$			& $26.791$  		& $1.079$			\\ \hline
-	Turkish   	& $2.355$			& $63.199$  		& $18.449$			\\ \hline
-	Yoruba    	& $1.178$			& $48.577$  		& $36.997$			\\ \hline
-	Basque    	& $0.393$			& $12.365$  		& $34.838$			\\ \hline
-	Swahili   	& $14.328$ 			& $85.574$  		& $0.000$			\\ \hline
-	Chinese   	& $1.865$ 			& $69.382$  		& $9.225$			\\ \hline
+	Malay & 3.827 & 88.420 & 0.000 \\ \hline
+	Estonian & 17.370 & 72.228 & 0.491 \\ \hline
+	Finnish & 34.446 & 56.624 & 0.000 \\ \hline
+	Hungarian & 34.347 & 58.292 & 0.000 \\ \hline
+	Armenian & 10.010 & 82.041 & 0.687 \\ \hline
+	Bengali & 16.765 & 37.782 & 25.63 \\ \hline
+	Japanese & 0.000 & 66.928 & 24.436 \\ \hline
+	Turkish & 2.748 & 62.807 & 18.744 \\ \hline
+	Yoruba & 1.178 & 48.184 & 38.371 \\ \hline
+	Basque & 0.393 & 5.496 & 58.587 \\ \hline
+	Swahili & 14.033 & 76.644 & 0.000 \\ \hline
+	Chinese & 5.986 & 51.717 & 24.338 \\ \hline
 	\end{tabular}
-	\caption{Percentage of female, male and neutral gender pronouns obtained for each language, averaged over all occupations detailed in Table \ref{tab:gender-neutral-languages}.
-Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}
+	\caption{\changess{Percentage of female, male and neutral gender pronouns obtained for each language, averaged over all occupations detailed in Table} \ref{tab:gender-neutral-languages}\changess{.
+Note that columns do not in general add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
 	\label{tab:gender-by-language}
@@
 Because a large number of adjectives are not appliable to human subjects, we manually curated a reasonable subset of such words.
-The template used for adjectives is identical to that used for occupations, but is provided again for reference in Table} \ref{tab:templates-adjectives}.
+The template used for adjectives is similar to that used for occupations, and is provided again for reference in Table} \ref{tab:templates-adjectives}.
 
@@
 Sentences containing the words \emph{Shy}, \emph{Attractive}, \emph{Happy}, \emph{Kind} and \emph{Ashamed} are predominantly female translated (\emph{Attractive} is translated as female and gender-neutral in equal parts), while \emph{Arrogant}, \emph{Cruel} and \emph{Guilty} are disproportionately translated with male pronouns (\emph{Guilty} is in fact never translated with female or neutral pronouns).
-
-% List of sentence templates for assessing gender bias in the translation from source language to English
-\begin{table}[H]
-\centering
-\small{
-	\begin{tabular}{|c|c|}
-	\hline
-	Language 	& Sentence template 													\\ \hline \hline
-	Malay		& dia adalah $\langle adjective \rangle$ 								\\ \hline
-	Estonian	& ta on $\langle adjective \rangle$ 									\\ \hline
-	Finnish		& hän on $\langle adjective \rangle$ 									\\ \hline
-	Hungarian	& ő egy $\langle adjective \rangle$ 									\\ \hline
-	Armenian	& na $\langle adjective \rangle$ e 									\\ \hline
-	\multirow{6}{*}{Bengali}	& Ē ē ēkajana $\langle adjective \rangle$				\\
-								& Yini yini ēkajana $\langle adjective \rangle$		\\
-								& Ō ō ēkajana $\langle adjective \rangle$				\\
-								& Uni uni ēkajana $\langle adjective \rangle$			\\
-								& Sē sē ēkajana $\langle adjective \rangle$			\\
-								& Tini tini ēkajana $\langle adjective \rangle$		\\ \hline
-	Japanese	& $\langle adjective \rangle$ \begin{CJK*}{UTF8}{min}です\end{CJK*}		\\ \hline
-	Turkish		& o bir $\langle adjective \rangle$ 									\\ \hline
-	Yoruba		& o jẹ $\langle adjective \rangle$ 									\\ \hline
-	Basque		& $\langle adjective \rangle$ da 										\\ \hline
-	Swahili		& yeye ni $\langle adjective \rangle$ 									\\ \hline
-	Chinese		& ta $\langle adjective \rangle$ 										\\ \hline
-	\end{tabular}
-	}
-	\caption{Templates used to infer gender biases in the translation to the English language.}
-	\label{tab:templates-adjectives}
-\end{table}
 
@@
 	\centering
-	\begin{tabular}{|c|c|c|c|c|}
+	\begin{tabular}{|c|c|c|c|}
 	\hline
-	Adjective 	& Female 	& Male 	& Neutral 	& Total 	\\ \hline
-	\hline
-	Happy			& $4$		& $3$	& $2$		& $11$		\\ \hline
-	Wrong			& $0$		& $6$	& $4$		& $11$		\\ \hline
-	Smart			& $2$		& $4$	& $3$		& $11$		\\ \hline
-	Modest			& $1$		& $5$	& $3$		& $11$		\\ \hline
-	Innocent		& $1$		& $7$	& $1$		& $11$		\\ \hline
-	Sad				& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Afraid			& $1$		& $5$	& $0$		& $11$		\\ \hline
-	Dumb			& $1$		& $5$	& $2$		& $11$		\\ \hline
-	Loving			& $2$		& $5$	& $2$		& $11$		\\ \hline
-	Right			& $0$		& $6$	& $3$		& $11$		\\ \hline
-	Brave			& $2$		& $6$	& $1$		& $11$		\\ \hline
-	Proud			& $1$		& $7$	& $1$		& $11$		\\ \hline
-	Cruel			& $1$		& $7$	& $2$		& $11$		\\ \hline
-	Guilty			& $0$		& $8$	& $0$		& $11$		\\ \hline
-	Shy				& $5$		& $2$	& $2$		& $11$		\\ \hline
-	Attractive		& $4$		& $1$	& $4$		& $11$		\\ \hline
-	STRONG			&			&		&			&			\\ \hline
-	DESIRABLE		&			&		&			&			\\ \hline
-	SYMPHATETIC		&			&		&			&			\\ \hline
-	POLITE			&			&		&			&			\\ \hline
-	SUCCESSFUL		&			&		&			&			\\ \hline
-	MATURE			&			&		&			&			\\ \hline \hline
-	Total 			& $38$		& $112$	& $41$		& -			\\ \hline 
+	Adjective & Female & Male & Neutral \\ \hline \hline
+	Happy & 36.364 & 27.273 & 18.182 \\ \hline
+	Sad & 18.182 & 36.364 & 18.182 \\ \hline
+	Right & 0.000 & 63.636 & 27.273 \\ \hline
+	Wrong & 0.000 & 54.545 & 36.364 \\ \hline
+	Afraid & 9.091 & 54.545 & 0.000 \\ \hline
+	Brave & 9.091 & 63.636 & 18.182 \\ \hline
+	Smart & 18.182 & 45.455 & 18.182 \\ \hline
+	Dumb & 18.182 & 36.364 & 18.182 \\ \hline
+	Proud & 9.091 & 72.727 & 9.091 \\ \hline
+	Strong & 9.091 & 54.545 & 18.182 \\ \hline
+	Polite & 18.182 & 45.455 & 18.182 \\ \hline
+	Cruel & 9.091 & 63.636 & 18.182 \\ \hline
+	Desirable & 9.091 & 36.364 & 45.455 \\ \hline
+	Loving & 18.182 & 45.455 & 27.273 \\ \hline
+	Sympathetic & 18.182 & 45.455 & 18.182 \\ \hline
+	Modest & 18.182 & 45.455 & 27.273 \\ \hline
+	Successful & 9.091 & 54.545 & 27.273 \\ \hline
+	Guilty & 0.000 & 72.727 & 0.000 \\ \hline
+	Innocent & 9.091 & 54.545 & 9.091 \\ \hline
+	Mature & 36.364 & 36.364 & 9.091 \\ \hline
+	Shy & 36.364 & 27.273 & 27.273 \\ \hline \hline
+	Total & 0.303 & 0.981 & 0.417 \\ \hline
+
 	\end{tabular}
-	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.
-Adjectives marked with an asterisk did not make up to the top $1000$ most frequently used adjectives in the COCA corpus but we have decided to include them because }
+	\caption{\changess{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}}
 	\label{tab:gender-by-adjective}
@@
 	\includegraphics[width=10cm]{pictures/barplot-adjectives}
-	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
+	\caption{\changess{The distribution of pronominal genders for each word in Table} \ref{tab:adjectives} \changess{shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
 One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns.
-Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
+Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}}
 	\label{fig:barplot-adjectives}
@@
 	\includegraphics[width=10cm]{pictures/histogram-compare-gt-real}
-	\caption{Women participation ($\%$) data obtained from the U.S.
+	\caption{\changess{Women participation ($\%$) data obtained from the U.S.
 Bureau of Labor Statistics allows us to assess whether the Google Translate bias towards male defaults is at least to some extent explained by small frequencies of female workers in some job positions.
-Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile (which is consistently inversely distributed) does not seem to respond to the higher proportion of female workers (in red) in the last quantiles.
-To remove doubt we have also plotted in yellow the frequency of translated female pronouns among male and female pronouns (i.e., excluding eventual translation errors and gender-neutral pronouns).}
+Our data does not make a very good case for that hypothesis: the total frequency of translated female pronouns (in blue) for each 12-quantile does not seem to respond to the higher proportion of female workers (in yellow) in the last quantiles.}}
 	\label{fig:histogram-compare-gt-real}

### 2018-04-20 gender-bias.py

@@ -5,120 +5,6 @@ from xpinyin import Pinyin
 import csv
 import numpy as np
 
-translator = Translator()
-p = Pinyin()
-
-do_occ = "occ" in sys.argv
-do_adj = "adj" in sys.argv
-
-
-# Get language list
-languages = []
-with open('languages.csv', 'r') as f:
-    f.readline()
-    for line in f:
-    	language,family,pronomial_gender_system,supported = line.split(";")
-    	if(pronomial_gender_system != 'Yes' and supported.rstrip() == 'Yes'):
-    		languages.append(language)
-#end with
-
-# Define list of discarded languages
-discarded_languages = ['Nepali','Tagalog','English','Maithili','Oriya','Korean','Cantonese','Pipil','Quechuan','Esperanto','Ido','Lingua Franca Nova','Interlingua']
-
-if do_occ:
-	# Read job occupations list into table
-	occupation_table = np.array(list(csv.reader(open('jobs/bureau_of_labor_statistics_profession_list_gender_filtered_expanded.tsv','r'), delimiter='\t')))
-	# Compute list of categories
-	categories = list(set(occupation_table[1:,-3]))
-	# Get dictionary of category -> jobs (jobs per category)
-	categories_dict = dict([ (category, occupation_table[occupation_table[:,-3] == category,0]) for category in categories ])
-#end if
-
-if do_adj:
-	adjectives = np.array( [ "happy", "sad", "right", "wrong", "afraid", "brave", "smart", "dumb", "proud", "strong", "polite", "cruel", "desirable", "loving", "sympathetic", "modest", "successful", "guilty", "innocent", "mature", "shy" ] ) # Could've read from file, but since the file is so small pasted it directly here.
-#end if
-
-"""
-	Create a occupation_table with the translated version (provided by Google Translate) of each job,
-	in the following structure:
-		1. Each line corresponds to a single occupation
-		2. The first column is the occupation category ("Computer and mathematical occupations", "Healthcare practitioners and technical occupations", etc.)
-		3. The following N columns give a translated version of that occupation for each language (of N) in our list
-"""
-if do_occ:
-	with open("Results/jobs-translations.tsv","w") as output:
-	
-		# Write header
-		# First column is category
-		output.write("Category")
-		# Then follows one column per language
-		output.write("\tEnglish")
-		for language in languages:
-			output.write('\t' + language)
-		#end for
-		output.write('\n')
-	
-		# Not iterate over all categories
-		for category in categories:
[…]

### 2018-04-20 stats.jl

@@ -1,23 +1,229 @@
 
-using DataFrames, CSV, Gadfly
+using DataFrames, CSV, Gadfly, Colors
 
-# Read compare data file
-cpm_GT_real = CSV.read("Results/translate-real-histogram-data.tsv",delim='\t',nullable=false)
+# Function to count the # of occurences of a gender (among languages) within a table
+function count_gender(gender, table)
+    
+    return count(x -> x==gender, convert(Array,table[:,3:7])) + round((1/6)*count(x -> x==gender, convert(Array,table[:,8:13]))) + count(x -> x==gender, convert(Array,table[:,14:end]))
+end
+
+function count_total(table)
+
+    return count(x -> true, convert(Array,table[:,3:7])) + round((1/6)*count(x -> true, convert(Array,table[:,8:13]))) + count(x -> true, convert(Array,table[:,14:end]))
+end
+
+function get_tables()
+
+    # Group results by category
+
+    # Get a list of categories
+    categories = convert(Array{String,1},unique(dat_jobs[:,:Category]))
+
+    results_by_category = DataFrame(Category=[], Female=[], Male=[], Neutral=[])
+    for category in categories
+
+        # Query the table rows relative to this category
+        category_filtered = dat_jobs[dat_jobs[:,:Category] .== category, :]
+
+        female_count    = count_gender("Female",   category_filtered)
+        male_count      = count_gender("Male",     category_filtered)
+        neutral_count   = count_gender("Neutral",  category_filtered)
+        total           = count_total(category_filtered)
+
+        push!(results_by_category, [category, round(100*female_count/total,3), round(100*male_count/total,3), round(100*neutral_count/total,3)])
+    end
+
+    # Get "totals" row
+    female_count    = count_gender("Female",   dat_jobs)
+    male_count      = count_gender("Male",     dat_jobs)
+    neutral_count   = count_gender("Neutral",  dat_jobs)
+    total           = count_total(dat_jobs)
+    push!(results_by_category, ["Total", round(100*female_count/total,3), round(100*male_count/total,3), round(100*neutral_count/total,3)])
+
+    CSV.write("Results/results-by-category.dat",delim='\t',results_by_category)
+
+    # Get a list of languages
+    languages = vcat( names(dat_jobs)[3:7], "Bengali", names(dat_jobs)[14:end] )
+    
+    results_by_language = DataFrame(Language=[],Female=[],Male=[],Neutral=[])
+    
[…]

### 2018-04-21 Paper/main.tex
Changes to paper and stats.jl

@@
 \usepackage[caption=false]{subfig}
+\usepackage[table]{xcolor}
 
@@
 
-A sensible way to group translation data is to coalesce occupations in the same category and collect statistics about how prominent male defaults are in each field.
-What we have found is that Google Translate does indeed translate sentences with male pronouns with greater probability than it does either female or gender-neutral pronouns.
+A sensible way to group translation data is to coalesce occupations in the same category and collect statistics \changess{among languages} about how prominent male defaults are in each field.
+What we have found is that \changess{Google Translate does indeed translate sentences with male pronouns with greater probability than it does either female or gender-neutral pronouns, in general}.
 Furthermore, this bias is seemingly aggravated for fields suggested to be troubled by male stereotypes, such as life and physical sciences, architecture, engineering, computer science and mathematics \citep{moss2015can}.
 Table \ref{tab:gender-by-category} summarizes these data, and Table \ref{tab:gender-by-category-grouped} summarizes it even further by coalescing occupation categories into broader groups to ease interpretation.
-For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields ($80.322\%$ of male defaults) to that of more evenly distributed fields such as healthcare ($59.014\%$).
+For instance, STEM (Science, Technology, Engineering and Mathematics) fields are grouped into a single category, which helps us compare the large asymmetry between gender pronouns in these fields (\changess{$72\%$} of male defaults) to that of more evenly distributed fields such as healthcare (\changess{$50\%$}).
 
-% PEDRO: TODO highlight table
 % Table with our gender bias results grouped by category
@@
 	\includegraphics[width=10cm]{pictures/histograms/all-categories-stacked-Female}
-	\caption{\changess{The data for the total number $X$ of female pronouns per occupation category suggests an inverse distribution.
+	\caption{\changess{The data for the number of translated female pronouns per occupation category totaled among languages suggests an inverse distribution.
 STEM fields such as life, social, computer and physical sciences, architecture, engineering and mathematics are nearly exclusively concentrated at $X=0$, while more evenly distributed fields such as production and healthcare (see Table} \ref{tab:gender-by-category}\changess{) extend to higher values.}}
@@
 
-AAAAAAAA
+\changess{We can also visualize male, female, and gender neutral histograms side by side, in which context is useful to compare the dissimilar distributions of translated STEM and Healthcare occupations (Figures }\ref{fig:histogram-dodged-STEM} \changess{and} \ref{fig:histogram-dodged-Healthcare} \changess{respectively).
+The number of translated female pronouns among languages is not normally distributed for any of the individual categories in Table} \ref{tab:occupations}\changess{, but Healthcare is in many ways the most balanced category, which can be seen in comparison with STEM -- in which male defaults are most prominent.}
 
@@
 	\includegraphics[width=10cm]{pictures/histograms/categories/gender-dodged-STEM}
-	\caption{TODO STEM}
+	\caption{\changess{Histograms for the distribution of the number of translated female, male and gender neutral pronouns totaled among languages are plotted side by side for job occupations in the STEM (Science, Technology, Engineering and Mathematics) field, in which male defaults are most prominent.}}
 	\label{fig:histogram-dodged-STEM}
@@
 	\includegraphics[width=10cm]{pictures/histograms/categories/gender-dodged-Healthcare}
-	\caption{TODO Healthcare}
+	\caption{\changess{Histograms for the distribution of the number of translated female, male and gender neutral pronouns totaled among languages are plotted side by side for job occupations in the Healthcare field, in which male defaults are least prominent.}}
 	\label{fig:histogram-dodged-Healthcare}
@@
 
-BBBBBVBBV
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Female}
-	\caption{TODO heatmap female}
-	\label{fig:heatmap-female}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Male}
-	\caption{TODO heatmap male}
-	\label{fig:heatmap-male}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Neutral}
-	\caption{TODO heatmap neutral}
-	\label{fig:heatmap-neutral}
-\end{figure}
-
 The bar plots in Figure \ref{fig:gender-by-category} help us visualize how much of the distribution of each occupation category is composed of female, male and gender-neutral pronouns.
-In this context, STEM fields, which show a predominance of male defaults, are contrasted with healthcare and educations, which show a larger proportion of female pronouns.
+In this context, STEM fields, which show a predominance of male defaults, are contrasted with Healthcare and educations, which show a larger proportion of female pronouns.
 
@@
 \end{figure}
+
+\changess{Although computing our statistics over the set of all languages has practical value, this may erase subtleties characteristic to each individual idiom.
+In this context, it is also important to visualize how each language translates job occupations in each category.
+The heatmaps in Figures } \ref{fig:heatmap-female}, \ref{fig:heatmap-male} \changess{and} \ref{fig:heatmap-neutral} \changess{ show the translation probabilities into female, male and neutral pronouns, respectively, for each pair of language and category (blue is $0\%$ and red is $100\%$).
+Both axes are sorted in these Figures, which helps us visualize both languages and categories in an spectrum of increasing male / female / neutral translation tendencies.
+Agreeing with suggested stereotypes,} \citep{moss2015can} \changess{ STEM fields are consistently translated with more male defaults than any other category.
+They are followed by Legal, Arts \& Entertainment and Corporate, in this order, while Healthcare, Production and Education lie on the opposite end of the spectrum.}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Female}
+	\caption{Heatmap for the translation probability into female pronouns for each pair of language and occupation category.
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+	\label{fig:heatmap-female}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Male}
+	\caption{Heatmap for the translation probability into male pronouns for each pair of language and occupation category.
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+	\label{fig:heatmap-male}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=10cm]{pictures/heatmap-languages-categories-Neutral}
+	\caption{Heatmap for the translation probability into gender neutral pronouns for each pair of language and occupation category.
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+	\label{fig:heatmap-neutral}
+\end{figure}
+
+\changess{Our analysis is not truly complete without tests for statistical significant differences in the translation tendencies among female, male and gender neutral pronouns.
+We want to know for which languages and categories does Google Translate translate sentences with significantly more male than female, or male than neutral, or neutral than female, pronouns.
+We ran one-sided t-tests to assess this question for each pair of language and category and also totaled among either languages or categories.
+The corresponding p-values are presented in Tables} \ref{tab:pvalues-MF}, \ref{tab:pvalues-MN}, \ref{tab:pvalues-NF} \changess{respectively.
+Language-Category pairs for which the null hypothesis was not rejected for a confidence level of $\alpha = .005$ are highlighted in blue.
+Although there is a decent level of variation among languages and categories, the null hypothesis that male pronouns are not significantly more frequent than female ones was consistently rejected for all languages and all categories examined.
+The same is true for the null hypothesis that male pronouns are not significantly more frequent than gender neutral pronouns, with the one exception of the Basque language (which exhibits a rather strong tendency towards neutral pronouns).
+The null hypothesis that neutral pronouns are not significantly more frequent than female ones is accepted with much more frequency, namely for the languages Malay, Estonian, Finnish, Hungarian, Armenian and for the categories Farming \& Fishing \& Forestry, Healthcare, Legal, Arts \& Entertainment, Education.}
+
+\begin{table}[H]
+\small
+\begin{tabular}{|m{1.75cm}|cccccccccccc|c|}
+\hline
+& Mal. & Est. & Fin. & Hun. & Arm. & Ben. & Jap. & Tur. & Yor. & Bas. & Swa. & Chi. & Total \\ \hline
+Service & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+STEM & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1} 
+Farming, Fishing, Forestry & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.603$ & \cellcolor{blue!25}$.786$ & $<\alpha$ & \cellcolor{blue!25}$.019$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $*$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Corporate & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.024$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.012$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Healthcare & $<\alpha$ & \cellcolor{blue!25}$.938$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.999$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.012$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Legal & $<\alpha$ & \cellcolor{blue!25}$.368$ & \cellcolor{blue!25}$.632$ & \cellcolor{blue!25}$.368$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.086$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Arts, Entertainment & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.08$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Education & $<\alpha$ & \cellcolor{blue!25}$.808$ & \cellcolor{blue!25}$.333$ & \cellcolor{blue!25}$.263$ & \cellcolor{blue!25}$.588$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.417$ & $<\alpha$ & \cellcolor{blue!25}$.052$ & \cellcolor{blue!25}$.023$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Production & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.01$ & \cellcolor{blue!25}$.5$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.159$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Construction, Extraction & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.16$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1} \hline
+Total & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & -  \\ \hline  
+\end{tabular}
+\caption{\changess{Computed p-values relative to the null hypothesis that the number of translated male pronouns is greater than that of female pronouns, organized for each language and each occupation category.
+A significance level of $\alpha =.005$ was adopted.
+Asterisks indicate cases in which all pronouns are translated with gender neutral pronouns.
+Cells corresponding to the acceptance of the null hypothesis are marked in} \colorbox{blue!25}{blue}.}
+\label{tab:pvalues-MF}
+\end{table}
+
+\begin{table}[H]
+\small
+\begin{tabular}{|m{1.75cm}|cccccccccccc|c|}
+\hline
+& Mal. & Est. & Fin. & Hun. & Arm. & Ben. & Jap. & Tur. & Yor. & Bas. & Swa. & Chi. & Total \\ \hline
+Service & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.01$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+STEM & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.984$ & $<\alpha$ & \cellcolor{blue!25}$.07$ & $<\alpha$  \\ \cline{1-1}
+Farming 
+ Fishing 
+ Forestry & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.009$ & $<\alpha$ & \cellcolor{blue!25}$.135$ & $<\alpha$ & \cellcolor{blue!25}$.014$ & \cellcolor{blue!25}$.068$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & \cellcolor{blue!25}$.027$ & $<\alpha$  \\ \cline{1-1}
+Corporate & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1} 
+Healthcare & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.39$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & \cellcolor{blue!25}$.088$ & $<\alpha$  \\ \cline{1-1}
+Legal & $<\alpha$ & \cellcolor{blue!25}$.015$ & \cellcolor{blue!25}$.039$ & \cellcolor{blue!25}$.015$ & $<\alpha$ & \cellcolor{blue!25}$.007$ & \cellcolor{blue!25}$.145$ & $<\alpha$ & \cellcolor{blue!25}$.023$ & \cellcolor{blue!25}$.771$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Arts 
+ Entertainment & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.07$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Education & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.017$ & \cellcolor{blue!25}$.093$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.5$ & $<\alpha$ & \cellcolor{blue!25}$.068$ & $<\alpha$  \\ \cline{1-1}
+Production & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.412$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Construction 
+ Extraction & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.92$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1} \hline
+Total & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & -  \\ \hline
+\end{tabular}
+\caption{\changess{Computed p-values relative to the null hypothesis that the number of translated male pronouns is greater than that of neutral pronouns, organized for each language and each occupation category.
+A significance level of $\alpha =.005$ was adopted.
+Asterisks indicate cases in which all pronouns are translated with female pronouns.
+Cells corresponding to the acceptance of the null hypothesis are marked in} \colorbox{blue!25}{blue}.}
+\label{tab:pvalues-MN}
+\end{table}
+
+\begin{table}[H]
+\small
+\begin{tabular}{|m{1.75cm}|cccccccccccc|c|}
+\hline
+& Mal. & Est. & Fin. & Hun. & Arm. & Ben. & Jap. & Tur. & Yor. & Bas. & Swa. & Chi. & Total \\ \hline
+Service & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.981$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+STEM & \cellcolor{blue!25}$.84$ & \cellcolor{blue!25}$.978$ & \cellcolor{blue!25}$.998$ & \cellcolor{blue!25}$.993$ & \cellcolor{blue!25}$.84$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.079$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.84$ & \cellcolor{blue!25}$.009$ & $<\alpha$  \\ \cline{1-1}
+Farming 
+ Fishing 
+ Forestry & $*$ & $*$ & \cellcolor{blue!25}$.999$ & \cellcolor{blue!25}$1.0$ & $*$ & \cellcolor{blue!25}$.167$ & \cellcolor{blue!25}$.169$ & \cellcolor{blue!25}$.292$ & \cellcolor{blue!25}$.041$ & $<\alpha$ & $*$ & \cellcolor{blue!25}$.083$ & \cellcolor{blue!25}$.147$  \\ \cline{1-1}
+Corporate & $*$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.996$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$.977$ & \cellcolor{blue!25}$.005$ & $<\alpha$  \\ \cline{1-1}
+Healthcare & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.086$ & $<\alpha$ & \cellcolor{blue!25}$.87$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & \cellcolor{blue!25}$.977$  \\ \cline{1-1}
+Legal & $*$ & \cellcolor{blue!25}$.961$ & \cellcolor{blue!25}$.985$ & \cellcolor{blue!25}$.961$ & $*$ & $<\alpha$ & \cellcolor{blue!25}$.086$ & $*$ & \cellcolor{blue!25}$.178$ & \cellcolor{blue!25}$.015$ & $*$ & $*$ & \cellcolor{blue!25}$.072$  \\ \cline{1-1}
+Arts 
+ Entertainment & \cellcolor{blue!25}$.92$ & \cellcolor{blue!25}$.994$ & \cellcolor{blue!25}$.999$ & \cellcolor{blue!25}$.998$ & \cellcolor{blue!25}$.998$ & \cellcolor{blue!25}$.067$ & \cellcolor{blue!25}$.006$ & \cellcolor{blue!25}$.042$ & \cellcolor{blue!25}$.042$ & $<\alpha$ & \cellcolor{blue!25}$.92$ & \cellcolor{blue!25}$.162$ & \cellcolor{blue!25}$.097$  \\ \cline{1-1}
+Education & $*$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.999$ & \cellcolor{blue!25}$.999$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.058$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.164$ & \cellcolor{blue!25}$.052$ & \cellcolor{blue!25}$.995$ & \cellcolor{blue!25}$.052$ & \cellcolor{blue!25}$.992$  \\ \cline{1-1}
+Production & \cellcolor{blue!25}$.996$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$.113$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1}
+Construction 
+ Extraction & \cellcolor{blue!25}$.84$ & \cellcolor{blue!25}$.996$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & $*$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$  \\ \cline{1-1} \hline
+Total & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & $<\alpha$ & \cellcolor{blue!25}$1.0$ & $<\alpha$ & -  \\ \hline
+\end{tabular}
+\caption{\changess{Computed p-values relative to the null hypothesis that the number of translated gender neutral pronouns is greater than that of female pronouns, organized for each language and each occupation category.
+A significance level of $\alpha =.005$ was adopted.
+Asterisks indicate cases in which all pronouns are translated with male pronouns.
+Cells corresponding to the acceptance of the null hypothesis are marked in} \colorbox{blue!25}{blue}.}
+\label{tab:pvalues-NF}
+\end{table}
 

### 2018-04-21 stats.jl

@@ -1,5 +1,5 @@
 
-using DataFrames, CSV, Gadfly, Colors
+using DataFrames, CSV, Gadfly, Colors, HypothesisTests
 
 # Function to count the # of occurences of a gender (among languages) within a table
 function count_gender(gender, table)
@@ -532,6 +532,142 @@ function heatmap_languages_categories()
     end
 end
 
+function statistical_tests()
+
+    # Get a list of languages
+    languages = vcat( names(dat_jobs)[3:7], Symbol("Bengali"), names(dat_jobs)[14:end] )
+
+    # Get list of occupation categories
+    categories = unique(dat_jobs[:Category])
+
+    pvalues = DataFrame(Language=[], Category=[], MF=[], MN=[], NF=[])
+
+    for language in languages
+        for category in categories
+
+            if language == :Bengali
+                filtered = convert(Array{String,2}, dat_jobs[ dat_jobs[:,:Category] .== category ,8:13] )
+                filtered = reshape(filtered, size(filtered)[1]*size(filtered)[2])
+            else
+                filtered = convert(Array{String,1}, dat_jobs[ dat_jobs[:,:Category] .== category ,language])
+            end
+
+            male    = map(x -> convert(Int64,x=="Male"), filtered)
+            female  = map(x -> convert(Int64,x=="Female"), filtered)
+            neutral = map(x -> convert(Int64,x=="Neutral"), filtered)
+
+            p_MF = pvalue(OneSampleTTest(male,female,0),tail=:right)
+            p_MN = pvalue(OneSampleTTest(male,neutral,0),tail=:right)
+            p_NF = pvalue(OneSampleTTest(neutral,female,0),tail=:right)
+
+            push!(pvalues,[language,category,p_MF,p_MN,p_NF])
+        end
+
+        if language == :Bengali
+            filtered = convert(Array{String,2}, dat_jobs[:,8:13] )
+            filtered = reshape(filtered, size(filtered)[1]*size(filtered)[2])
+        else
+            filtered = convert(Array{String,1}, dat_jobs[:,language])
+        end
+
+        male    = map(x -> convert(Int64,x=="Male"), filtered)
+        female  = map(x -> convert(Int64,x=="Female"), filtered)
+        neutral = map(x -> convert(Int64,x=="Neutral"), filtered)
+
+        p_MF = pvalue(OneSampleTTest(male,female,0),tail=:right)
[…]

### 2018-04-21 Paper/main.tex
Small change to paper

@@
 	\hline
-	Adjective & Female & Male & Neutral \\ \hline \hline
+	Adjective & Female ($\%$) & Male ($\%$) & Neutral ($\%$) \\ \hline \hline
 	Happy & 36.364 & 27.273 & 18.182 \\ \hline
@@
 	Shy & 36.364 & 27.273 & 27.273 \\ \hline \hline
-	Total & 0.303 & 0.981 & 0.417 \\ \hline
+	Total & 30.3 & 98.1 & 41.7 \\ \hline
 

### 2018-04-22 Paper/main.tex
Some changes to the paper and figures generation

@@
 
-\subsection{Rationale for language exceptions}
-\label{ssec:rationale_for_language_exceptions}
-
-\changess{
-While it is possible to construct gender neutral sentences in two of the languages omitted in our experiments (namely Korean and Nepali), we have chosen to omit them for the following reasons:}
-
-\begin{enumerate}
-\item \changess{We faced technical difficulties to form templates and automatically translate sentences with the right-to-left, top-to-bottom nature of the Persian script, and as such have decided not to include it in our experiments.}
-\item \changess{Due to Nepali having a rather complex grammar, with possible male/female gender demarcations on the phrases and due to none of the authors being fluent or able to reach someone fluent in the language, we were not confident enough in our ability to produce the required templates.
-Bengali was almost discarded under the same rationale, but we have decided to keep it because our sentence template for Bengali has a simple grammatical structure which does not require any kind of inflection.}
-\item \changess{While technically, one can construct gender neutral phrases in Korean by omitting the gender pronoun, in fact, this is default procedure.
-However, the expressiveness of this omission depends on the context of the sentence being clear, which is not possible in the way we frame phrases.}
-\end{enumerate}
-
-% PEDRO: TODO highlight table
 % List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
@@
 
-% PEDRO: TODO highlight section
+\subsection{Rationale for language exceptions}
+\label{ssec:rationale_for_language_exceptions}
+
+\changess{
+While it is possible to construct gender neutral sentences in two of the languages omitted in our experiments (namely Korean and Nepali), we have chosen to omit them for the following reasons:}
+
+\begin{enumerate}
+\item \changess{We faced technical difficulties to form templates and automatically translate sentences with the right-to-left, top-to-bottom nature of the Persian script, and as such have decided not to include it in our experiments.}
+\item \changess{Due to Nepali having a rather complex grammar, with possible male/female gender demarcations on the phrases and due to none of the authors being fluent or able to reach someone fluent in the language, we were not confident enough in our ability to produce the required templates.
+Bengali was almost discarded under the same rationale, but we have decided to keep it because our sentence template for Bengali has a simple grammatical structure which does not require any kind of inflection.}
+\item \changess{One can construct gender neutral phrases in Korean by omitting the gender pronoun, in fact, this is default procedure.
+However, the expressiveness of this omission depends on the context of the sentence being clear, which is not possible in the way we frame phrases.}
+\end{enumerate}
+
 \section{Distribution of translated gender pronouns per occupation category}
@@
 	\includegraphics[width=10cm]{pictures/histograms/categories/gender-dodged-STEM}
-	\caption{\changess{Histograms for the distribution of the number of translated female, male and gender neutral pronouns totaled among languages are plotted side by side for job occupations in the STEM (Science, Technology, Engineering and Mathematics) field, in which male defaults are most prominent.}}
+	\caption{\changess{Histograms for the distribution of the number of translated female, male and gender neutral pronouns totaled among languages are plotted side by side for job occupations in the STEM (Science, Technology, Engineering and Mathematics) field, in which male defaults are the second-to-most prominent (after Legal).}}
 	\label{fig:histogram-dodged-STEM}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/barplot-gender-by-category-grouped}
+	\includegraphics[width=10cm]{pictures/barplot-gender-by-category}
 	\caption{\changess{Bar plots show how much of the distribution of translated gender pronouns for each occupation category (grouped as in Table} \ref{tab:gender-by-category-grouped}\changess{) is composed of female, male and neutral terms.
-STEM fields exhibit a predominance of male defaults and contrast with healthcare and education, with a larger proportion of female and neutral pronouns.
-Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.}}
+Legal and STEM fields exhibit a predominance of male defaults and contrast with Healthcare and Education, with a larger proportion of female and neutral pronouns.
+Note that in general the bars do not add up to $100\%$, as there is a fair amount of translated sentences for which we cannot obtain a gender pronoun.
+Categories are sorted with respect to the proportions of male, female and neutral translated pronouns respectively.}}
 	\label{fig:gender-by-category}
@@
 Both axes are sorted in these Figures, which helps us visualize both languages and categories in an spectrum of increasing male / female / neutral translation tendencies.
-Agreeing with suggested stereotypes,} \citep{moss2015can} \changess{ STEM fields are consistently translated with more male defaults than any other category.
-They are followed by Legal, Arts \& Entertainment and Corporate, in this order, while Healthcare, Production and Education lie on the opposite end of the spectrum.}
+Agreeing with suggested stereotypes,} \citep{moss2015can} \changess{ STEM fields are second only to Legal ones in the prominence of male defaults.
+These two are followed by Arts \& Entertainment and Corporate, in this order, while Healthcare, Production and Education lie on the opposite end of the spectrum.}
 
@@
 	\caption{Heatmap for the translation probability into female pronouns for each pair of language and occupation category.
-Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom right corner.}
 	\label{fig:heatmap-female}
@@
 	\caption{Heatmap for the translation probability into male pronouns for each pair of language and occupation category.
-Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom right corner.}
 	\label{fig:heatmap-male}
@@
 	\caption{Heatmap for the translation probability into gender neutral pronouns for each pair of language and occupation category.
-Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom left corner.}
+Probabilities range from $0\%$ (blue) to $100\%$ (red), and both axes are sorted in such a way that higher probabilities concentrate on the bottom right corner.}
 	\label{fig:heatmap-neutral}
@@
 	\centering
-	\includegraphics[width=10cm]{pictures/gender-by-language}
+	\includegraphics[width=10cm]{pictures/barplot-gender-by-language}
 	\caption{\changess{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.196\%$ and $1.865\%$ for Japanese and Chinese respectively.
 Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque.
-Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.}}
+Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns, with Yoruba and Turkish following after in this order.
+Languages are sorted with respect to the proportions of male, female and neutral translated pronouns respectively.}}
 	\label{fig:gender-by-language}
@@
 	\caption{\changess{The distribution of pronominal genders for each word in Table} \ref{tab:adjectives} \changess{shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives.
-One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns.
-Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}}
+One can see that adjectives such as \emph{Shy} and \emph{Desirable}, \emph{Sad} and \emph{Dumb} amass at the female side of the spectrum, contrasting with \emph{Proud}, \emph{Guilty}, \emph{Cruel} and \emph{Brave} which are almost exclusively translated with male pronouns.
+}}
 	\label{fig:barplot-adjectives}
@@
 
-A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that (possibly) female participation in some job positions is itself low.
+A sensible objection to the conclusions we draw from our study is that the perceived gender bias in Google Translate results stems from the fact that possibly female participation in some job positions is itself low.
 We must account for the possibility that the statistics of gender pronouns in Google Translate outputs merely reflects the demographics of male-dominated fields (male-dominated fields can be considered those that have less than 25\% of women participation\citep{WB2014}, according to the U.S.

### 2018-04-22 stats.jl

@@ -299,38 +299,6 @@ function group_categories(label)
     end
 end
 
-function barplots_adjectives()
-    results_by_adjective = DataFrame(Adjective=[],Gender=[],Count=[],Ratio=[])
-
-    for adjective in dat_adjectives[:Adjective]
-        aux = dat_adjectives[ dat_adjectives[:,:Adjective] .== adjective, : ]
-        
-        female_count    = count_gender("Female",    aux)
-        male_count      = count_gender("Male",      aux)
-        neutral_count   = count_gender("Neutral",   aux)
-        total           = count_total(aux)
-        ratio = male_count/female_count
-        
-        push!(results_by_adjective, [ucfirst(adjective),"Female",   100 * female_count    / total,  ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Male",     100 * male_count      / total,  ratio])
-        push!(results_by_adjective, [ucfirst(adjective),"Neutral",  100 * neutral_count   / total,  ratio])
-    end
-
-    CSV.write("Results/results-by-adjective.dat",results_by_adjective)
-
-    sort!(results_by_adjective, cols = [order(:Ratio)])
-
-    # Save bar plot
-    p = plot(
-        results_by_adjective,
-        x=:Adjective, y=:Count, color=:Gender,
-        Coord.Cartesian(ymin=0,ymax=100),
-        Geom.bar(position=:stack),
-        Theme(background_color="white",grid_color="gray",bar_highlight=colorant"black",bar_spacing=1mm),
-        Guide.ylabel("\%"))
-    draw(PDF("Paper/pictures/barplot-adjectives.pdf", 7.50inch, 5.625inch),p)
-end
-
 function histograms_occupations_per_category()
 
     # Get the top 10 most frequent categories
@@ -376,20 +344,26 @@ function barplots_category()
     # Compute results table (grouped by category)
     # Get list of occupation categories
     categories = unique(dat_jobs[:Category])
-    results_by_category = DataFrame(Category=[],Gender=[],Count=[])
+    results_by_category = DataFrame(Category=[],Gender=[],Count=[],Male=[],Female=[],Neutral=[])
     for category in categories
 
         aux = dat_jobs[dat_jobs[:,:Category] .== category, 3:end]
 
-        female_count    = count(x -> x=="Female",   convert(Array,aux[:,3:end]))
-        male_count      = count(x -> x=="Male",     convert(Array,aux[:,3:end]))
[…]

### 2018-04-23 stats.jl
New stats

@@ -580,16 +580,27 @@ function statistical_tests()
     # Get list of occupation categories
     categories = unique(dat_jobs[:Category])
 
-    pvalues = DataFrame(Language=[], Category=[], MF=[], MN=[], NF=[])
+    pvalues = DataFrame(Language=[], Category=[], MF=[], FM=[], MN=[], NM=[], NF=[], FN=[])
 
-    for language in languages
-        for category in categories
+    for language in vcat(languages, [Symbol("Total")])
+        for category in vcat(categories, [Symbol("Total")])
 
-            if language == :Bengali
-                filtered = convert(Array{String,2}, dat_jobs[ dat_jobs[:,:Category] .== category ,8:13] )
+            # Filter by category (or not, if total)
+            if category == :Total
+                filtered = dat_jobs[:,:]
+            else
+                filtered = dat_jobs[dat_jobs[:,:Category].==category,:]
+            end
+
+            # Filter by language (or not, if total)
+            if language == :Total
+                filtered = convert(Array{String,2}, filtered[:,:])
+                filtered = reshape(filtered, size(filtered)[1]*size(filtered)[2])
+            elseif language == :Bengali
+                filtered = convert(Array{String,2}, filtered[:,8:13])
                 filtered = reshape(filtered, size(filtered)[1]*size(filtered)[2])
             else
-                filtered = convert(Array{String,1}, dat_jobs[ dat_jobs[:,:Category] .== category ,language])
+                filtered = convert(Array{String,1}, filtered[:,language])
             end
 
             male    = map(x -> convert(Int64,x=="Male"), filtered)
@@ -597,68 +608,47 @@ function statistical_tests()
             neutral = map(x -> convert(Int64,x=="Neutral"), filtered)
 
             p_MF = pvalue(OneSampleTTest(male,female,0),tail=:right)
+            p_FM = pvalue(OneSampleTTest(male,female,0),tail=:left)
             p_MN = pvalue(OneSampleTTest(male,neutral,0),tail=:right)
+            p_NM = pvalue(OneSampleTTest(male,neutral,0),tail=:left)
             p_NF = pvalue(OneSampleTTest(neutral,female,0),tail=:right)
+            p_FN = pvalue(OneSampleTTest(neutral,female,0),tail=:left)
 
[…]

### 2018-04-23 stats.jl
Extreme bookkeeping

@@ -220,9 +220,8 @@ function histograms_compare()
         color=:Data,
         Geom.bar(position=:dodge),
         Guide.xticks(ticks=[1:12;]),
-        
-        Theme(key_max_columns=1,key_label_font_size=6pt))
-    draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 7.50inch, 5.625inch),p)
+        Theme(key_max_columns=1,key_label_font_size=6pt, key_position = :top, key_title_font_size=12pt, key_label_font_size=8pt))
+    draw(PDF("Paper/pictures/histogram-compare-gt-real.pdf", 5.625inch, 5.625inch),p)
 end
 
 function rename_category(label)
@@ -371,10 +370,20 @@ function barplots_category()
         x="Category", y="Count", color="Gender",
         Coord.Cartesian(ymin=0,ymax=100),
         Geom.bar(position=:stack),
-        Theme(background_color="white",grid_color="gray",bar_highlight=colorant"black",bar_spacing=1mm),
+        
+        Theme(
+            background_color="white",
+            grid_color="gray",
+            bar_highlight=colorant"black",
+            bar_spacing=1mm,
+            key_position = :right,
+            key_title_font_size=12pt,
+            key_label_font_size=8pt,
+            plot_padding=[0pt,5mm,5mm,0pt]),
+        
         Guide.ylabel("\%"),
         Guide.xticks(orientation=:vertical))
-    draw(PDF("Paper/pictures/barplot-gender-by-category.pdf", 7.50inch, 5.625inch),p)
+    draw(PDF("Paper/pictures/barplot-gender-by-category.pdf", 5.625inch, 5.625inch),p)
 end
 
 function barplots_language()
@@ -427,10 +436,18 @@ function barplots_language()
         x="Language", y="Count", color="Gender",
         Coord.Cartesian(ymin=0,ymax=100),
         Geom.bar(position=:stack),
-        Theme(background_color="white",bar_highlight=colorant"black",bar_spacing=1mm),
+        Theme(
+            background_color="white",
+            grid_color="gray",
+            bar_highlight=colorant"black",
+            bar_spacing=1mm,
+            key_position = :right,
+            key_title_font_size=12pt,
+            key_label_font_size=8pt,
+            plot_padding=[0pt,5mm,5mm,0pt]),
         Guide.ylabel("\%"),
[…]
